### 2021-09-10 HW0/hw0.tex
init hw0

\documentclass{article}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath, amsthm, verbatim, enumerate, xcolor, mathpartir}

\input{macros}

\title{IIT CS595: Topics \& Applications in Programming Languages\\
  {\large Homework 0: Rule induction, expressions, type safety}}
\author{Your name here}

\begin{document}
\maketitle

\section{Logistics}

\paragraph{Writing up Solutions.}
If you're looking at this file, you're probably (at least thinking about)
writing up your solutions in LaTeX. Nice! If you need help, see the writeup
PDF for a link to resources and other options.

\paragraph{Collaboration.}
Please see the collaboration policy on the course website.

\paragraph{Submission.}
Submit your answers as a single .pdf or .doc file (see above) on Blackboard
under
the correct assignment.

\section{Evaluation}
\begin{task}
  Evaluate the following expression using the dynamics semantics from class,
  showing each step:

  (Put your answers in the ... and copy and paste the line to add more lines
  as necessary).

  \[
  \begin{array}{l l}
    &
    \kwlet{x}{\kwn{2}}{x+(\kwlet{x}{\kwn{1}}{\kwlet{y}{x+\kwn{1}}{x+y}})}\\
    \step & ...\\
    \step & ...\\
  \end{array}
  \]
\end{task}

\section{Red-Black Trees}
In data structures, a red-black tree is a binary tree whose nodes are colored
either red or black.
%
Red-black trees must obey the following rules:
\begin{itemize}
\item Leaf nodes must be black.
\item No red node has red children.
\item Any path from the root to a leaf contains the same number of black nodes.
\end{itemize}

We can express red-black trees using the following small language
(note that this syntax doesn't enforce the rules above; we'll do that later).

\[
\begin{array}{r l l l}
  Trees & \tree & \bnfdef & \leaf \bnfalt \bnode{\tree}{\tree} \bnfalt
  \rnode{\tree}{\tree}\\
  Colors & \colr & \bnfdef & \bc \bnfalt \rc\\
\end{array}
\]
where~$\leaf$ is a leaf and~$\bnode{\tree_1}{\tree_2}$
and~$\rnode{\tree_1}{\tree_2}$ represent black and red nodes, respectively,
with children~$\tree_1$ and~$\tree_2$.

We will define a judgment~$\tree\istree{\colr}{n}$, meaning that~$\tree$ is
a valid red-black tree whose root node has the color~$\colr$ (which is
either~$\bc$ or~$\rc$) and all of whose paths from root to leaf have
exactly~$n$ black nodes.

{
  \centering
  \def \MathparLineskip {\lineskip=0.43cm}
  \begin{mathpar}

    \Rule{RBT-1}
         {\strut}
         {\leaf \istree{\bc}{1}}
    \and
    \Rule{RBT-2}
         {\tree_1 \istree{\colr_1}{n}\\
           \tree_2 \istree{\colr_2}{n}}
         {\bnode{\tree_1}{\tree_2} \istree{\bc}{n+1}}
    \and
    \Rule{RBT-3}
         {\tree_1 \istree{\bc}{n}\\
           \tree_2 \istree{\bc}{n}}
         {\rnode{\tree_1}{\tree_2} \istree{\rc}{n}}
  \end{mathpar}
}

Rule~(RBT-1) says that leaves are valid trees as long as they are
black: a leaf by definition has~1 black node on any path from root to leaf.
%
Rule~(RBT-2) allows a black node with two children as long as both
children are valid red-black trees with the same number of black nodes on any
path (regardless of what color their roots are); the result is a tree with a
black root and~$n+1$ black nodes on any path.
%
Rule~(RBT-3) enforces that red nodes do not have red children by
requiring in the premises that both children have black roots.

Define~$\nodes{\tree}$ to be the number of nodes, including leaves, in a
tree:

\[
\begin{array}{l l l}
  \nodes{\leaf} & = & 1\\
  \nodes{\bnode{\tree_1}{\tree_2}} & = & 1 + \nodes{\tree_1} + \nodes{\tree_2}\\
  \nodes{\rnode{\tree_1}{\tree_2}} & = & 1 + \nodes{\tree_1} + \nodes{\tree_2}
\end{array}
\]

\begin{task}
  Prove by rule induction that for any tree~$\tree$,
  color~$\colr \in \{\rc, \bc\}$ and~$n \geq 1$,
  if~$\tree \istree{\colr}{n}$, then~$\nodes{\tree} \geq 2^{n-1}$.

  {\em You must} use rule induction, not any other proof technique.

  \textbf{Note:} This, together with one or two other properties of
  red-black trees, proves that a valid red-black tree is approximately
  balanced.
\end{task}

\textbf{Answer:}

\section{Booleans}

In this task, you'll extend the {\Elang} language from class with Booleans:

\[
\begin{array}{r l l l}
  \mathit{Types} & \tau & \bnfdef &
  \kwint \bnfalt \kwstring \bnfalt \kwbool\\
  \mathit{Expressions} & e & \bnfdef &
  x \bnfalt
  \kwn{n} \bnfalt
  \kws{s} \bnfalt
  \kwtrue \bnfalt
  \kwfalse \bnfalt
  e = e \bnfalt
  e + e \bnfalt
  e \cat e \bnfalt
  \kwlen{e} \bnfalt
  \kwlet{x}{e}{e} \bnfalt
  \kwif{e}{e}{e}
\end{array}
\]

The expressions~$\kwtrue$ and~$\kwfalse$ are values and have their usual
meanings.
%
The expression~$\kwif{e_1}{e_2}{e_3}$ should evaluate~$e_1$. If it evaluates
to~$\kwtrue$, it should continue evaluating~$e_2$, otherwise it should
continue evaluating~$e_3$.
%
The other branch {\em should not} be evaluated (this isn't possible to express
in {\Elang}, but consider the code $\kwif{x = \kwn{0}}{\kwn{0}}{\kwn{42} / x}$: we certainly
don't want to evaluate the else branch when the condition is true!).
%
We've also added an integer equality test~$e_1 = e_2$ to have an interesting
way of producing Booleans (this is an operation on integers only, not on
Booleans or strings).
%
Recall the dynamic semantics from class, now extended with the rules
for~$e_1 = e_2$.

\textbf{Note:} There's a lot here, but don't panic:
the only new rules here are (S-11) through (S-14). The rest are just there
as a reminder. Your job is only to add Booleans to the language.

{
  \centering
  \def \MathparLineskip {\lineskip=0.43cm}
  \begin{mathpar}
    \Rule{V-1}
         {\strut} % This leaves the proper spacing above the line of an axiom
         {\kwn{n}\val}
    \and
    \Rule{V-2}
         {\strut}
         {\kws{s}\val}
    \and
    \Rule{S-1}
         {\strut}
         {\kwn{n_1} + \kwn{n_2} \step \kwn{n_1 + n_2}}
    \and
    \Rule{S-2}
         {\strut}
         {\kws{s_1} \cat \kws{s_2} \step \kws{s_1s_2}}
    \and
    \Rule{S-3}
         {\strut}
         {\kwlen{\kws{s}} \step \kwn{\kwlen{s}}}
    \and
    \Rule{S-4}
         {e_1 \step e_1'}
         {e_1 + e_2 \step e_1' + e_2}
    \and
    \Rule{S-5}
         {e_2 \step e_2'}
         {\kwn{n_1} + e_2 \step \kwn{n_1} + e_2'}
    \and
    \Rule{S-6}
         {e_1 \step e_1'}
         {e_1 \cat e_2 \step e_1' \cat e_2}
    \and
    \Rule{S-7}
         {e_2 \step e_2'}
         {\kws{s_1} \cat e_2 \step \kws{s_1} \cat e_2'} 
    \and
    \Rule{S-8}
         {e \step e'}
         {\kwlen{e} \step \kwlen{e'}}
    \and
    \Rule{S-9}
         {e_1 \step e_1'}
         {\kwlet{x}{e_1}{e_2} \step \kwlet{x}{e_1'}{e_2}}
    \and
    \Rule{S-10}
         {v \val}
         {\kwlet{x}{v}{e_2} \step \sub{v}{x}{e_2}}
    \and
    \Rule{S-11}
         {e_1 \step e_1'}
         {e_1 = e_2 \step e_1' = e_2}
    \and
    \Rule{S-12}
         {e_2 \step e_2'}
         {\kwn{n_1} = e_2 \step \kwn{n_1} = e_2'}
    \and
    \Rule{S-13}
         {\strut}
         {\kwn{n} = \kwn{n} \step \kwtrue}
    \and
    \Rule{S-14}
         {n_1 \neq n_2}
         {\kwn{n_1} = \kwn{n_2} \step \kwfalse}
  \end{mathpar}
}

\begin{task}\label{task:dyn}
  Write the new inference rules for the dynamic semantics of Booleans and the
  if-else construct.
  You should have 2 new rules for the judgment~$e\val$ and 3 new rules
  for the judgment~$e \step e$.\\
  (\textbf{Hint:} only one of these will be a search rule.)
\end{task}

\textbf{Answer:}

We also extend the typing rules with the new rule for equality testing and
Booleans.

{
  \centering
  \def \MathparLineskip {\lineskip=0.43cm}
  \begin{mathpar}
    \Rule{T-8}
         {\typed{\ctx}{e_1}{\kwint}\\
           % Use \\ to separate multiple premises
           \typed{\ctx}{e_2}{\kwint}}
         {\typed{\ctx}{e_1 = e_2}{\kwbool}}
    \and
    \Rule{T-9}
         {\strut}
         {\typed{\ctx}{\kwtrue}{\kwbool}}
    \and
    \Rule{T-10}
         {\strut}
         {\typed{\ctx}{\kwtrue}{\kwbool}}
    \and
    \Rule{T-11}
         {\typed{\ctx}{e_1}{\kwbool}\\
           \typed{\ctx}{e_2}{\tau}\\
           \typed{\ctx}{e_3}{\tau}}
         {\typed{\ctx}{\kwif{e_1}{e_2}{e_3}}{\tau}}
  \end{mathpar}
}

Note that rule (T-11) doesn't require that the two branches of the conditional
have a particular type (e.g.~$\kwint$).
%
The use of the same metavariable~$\tau$ {\em does} however, mean that the
two branches~$e_2$ and~$e_3$ must have the {\em same} type, which is then the
type of the whole expression
(this makes sense: how could we possibly give a type to the expression
$\kwif{n = \kwn{0}}{\kwn{42}}{\kws{\text{Oops}}}$?).

\begin{task}
  Prove the cases of the Preservation theorem for the new rules you
  added in Task~\ref{task:dyn}.
\end{task}

\textbf{Answer:}

\begin{task}\label{task-cf}
  State (you do not have to prove it) the new case of the Canonical Forms
  lemma for Booleans.
\end{task}

\textbf{Answer:}

\begin{task}
  Prove the cases of the Progress theorem for the new rules
  (T-9) through (T-11). You may (and should) use the new case of Canonical
  Forms from Task~\ref{task-cf}.
\end{task}

\textbf{Answer:}

\end{document}

### 2021-09-14 HW0/hw0.tex
complete task 2

@@
   {\large Homework 0: Rule induction, expressions, type safety}}
-\author{Your name here}
+\author{Dustin Van Tate Testa}
 
@@
     \kwlet{x}{\kwn{2}}{x+(\kwlet{x}{\kwn{1}}{\kwlet{y}{x+\kwn{1}}{x+y}})}\\
-    \step & ...\\
-    \step & ...\\
+    \step & \kwn{2}+(\kwlet{x}{\kwn{1}}{\kwlet{y}{x+\kwn{1}}{x+y}})\\
+    \step & \kwn{2}+(\kwlet{y}{\kwn{1}+\kwn{1}}{\kwn{1}+y}\\
+    \step & \kwn{2}+(\kwlet{y}{\kwn{2}}{\kwn{1}+y}\\
+    \step & \kwn{2}+(\kwn{1}+\kwn{2})\\
+    \step & \kwn{2}+\kwn{3}\\
+    \step & \kwn{5}\\
   \end{array}
@@
 \textbf{Answer:}
+
+- Notice use of shorthand $\kwlen{\tree}$ to indicate the value of $n$ (black height) for tree \tree \\
+
+Base Case: 
+\begin{itemize}
+    \item all cases when n = 1:
+        \begin{itemize}
+            \item {\tree = \leaf}: Rule~(RBT-1) confirms that when T is a single black node, n = 1.\\
+            The proposition holds for this case as $\nodes{\leaf} = 1 \geq 2^{n-1} = 2^{1-1} = 1$.
+
+            \item {\tree = \rnode{\leaf}{\leaf}}: rule~(RBT-3) confirms that when $\tree$ is a red node,
+            n is the same as it's children, which when they're both leaves makes n = 1.\\
+            The proposition holds for this case as follows:\\
+            LHS: $\nodes{\rnode{\leaf}{\leaf}} = 1 + \nodes{\leaf} + \nodes{\leaf} = 1 + 1 + 1 = 3$\\
+            RHS: $2^{n-1} = 2 ^ {1 - 1} = 2^0 = 1$\\
+            giving $3 \geq 1$ 
+        \end{itemize}
+\end{itemize}
+
+Inductive case assuming $\nodes{\tree} \geq 2^{n-1}$:
+\begin{itemize}
+    \item Case when tree's top element is a black node
+        \[
+        \begin{array}{l l l l}
+            \nodes{\bnode{\tree_1}{\tree_2}} & \geq & 2^{\kwlen{\tree}-1} & \text{the proposition for this case}\\
+            1 + \nodes{\tree_1} + \nodes{\tree_2} & \geq & 1 + 2 ^ {\kwlen{\tree_1} - 1} + 2 ^ {\kwlen{\tree_2} - 1} & \text{apply rule (RBT-2)}\\
+            1 & \geq & 1  & \text{apply inductive hypothesis}\\
+        \end{array}
+        \]
+
+    \item Case when tree's top element is a red node
+        \[
+        \begin{array}{l l l l}
+            \nodes{\rnode{\tree_1}{\tree_2}} & \geq & 2^{\kwlen{\tree}-1} & \text{the proposition for this case}\\
+            1 + \nodes{\tree_1} + \nodes{\tree_2} & \geq & 2 ^ {\kwlen{\tree_1} - 1} + 2 ^ {\kwlen{\tree_2} - 1} & \text{apply rule (RBT-3)}\\
+            1 & \geq & 0  & \text{apply inductive hypothesis}\\
+        \end{array}
+        \]
+\end{itemize}
 

### 2021-09-17 HW0/hw0.tex
fix prof typo, task 3
HW0 done

@@
 \begin{itemize}
-    \item Case when tree's top element is a black node
+    \item Case when $\tree = \bnode{...}{...}$ -- tree's top element is a black node
         \[
@@
 
-    \item Case when tree's top element is a red node
+    \item Case when $\tree = \rnode{...}{...}$ -- tree's top element is a red node
         \[
@@
 
+{
+    \centering
+    \def \MathparLineskip{\lineskip=0.43cm}
+    \begin{mathpar}
+        \Rule{V-3}
+            {\strut}
+            {\kwfalse \val}
+        \and
+        \Rule{V-4}
+            {\strut}
+            {\kwtrue \val}
+        \and
+        \Rule{S-15}
+            {e_1 \step e_1'}
+            {\kwif{e_0}{e_1}{e_2} \step \kwif{e_0'}{e_1}{e_2}}
+        \and
+        \Rule{S-16}
+            {\strut}
+            {\kwif{\kwtrue}{e_1}{e_2} \step e_1}
+        \and
+        \Rule{S-17}
+            {\strut}
+            {\kwif{\kwfalse}{e_1}{e_2} \step e_2}
+    \end{mathpar}
+}
+
 We also extend the typing rules with the new rule for equality testing and
@@
          {\strut}
-         {\typed{\ctx}{\kwtrue}{\kwbool}}
+         {\typed{\ctx}{\kwfalse}{\kwbool}}
     \and
@@
 
+Proposition: if $\typed{\ctx}{e}{\tau}$ and $e \step e'$ then $\typed{\ctx}{e'}{\tau}$
+\begin{itemize}
+    \item S-16: When $e = \kwif{\kwtrue}{e_1}{e_2}$\\ \step $e' = e_1$\\
+        by inversion on rule (T-11) $\typed{\ctx}{\kwif{e_0}{e_1}{e_2}}{\tau}$ 
+            and  $\typed{\ctx}{e_1}{\tau}$\\
+        with induction on $e_1 \step e_1'$ this confirms that the type of the stepped value is preserved.
+
+    \item S-17: When $e = \kwif{\kwfalse}{e_1}{e_2}$\\ \step $e' = e_2$\\
+        by inversion on rule (T-11) $\typed{\ctx}{\kwif{e_0}{e_1}{e_2}}{\tau}$ 
+            and  $\typed{\ctx}{e_2}{\tau}$\\
+        with induction on $e_2 \step e_2'$ this confirms that the type of the stepped value is preserved.
+
+    \item S-15: When $e = \kwif{e_0}{e_1}{e_2}$\\ \step $e' = \kwif{e_0'}{e_1}{e_2}$\\
+        by inversion on rule (T-11), $\typed{\ctx}{e_0}{\kwbool}$, $\typed{\ctx}{e_1}{\tau}$, 
+            and $\typed{\ctx}{e_2}{\tau}$\\
+        by induction we can assume $e_0 \step e_0'$ with $\typed{\ctx}{e_0'}{\kwbool}$\\
+        thus we can apply (T-11) to assert that the type is the same.
+\end{itemize}
+
 \begin{task}\label{task-cf}
@@
 \textbf{Answer:}
+if $e \val$ and $\typed{}{e}{\kwbool}$ then $e$ is either $\kwtrue$ or $\kwfalse$.
 
@@
 
+Proposition: if $\typed{}{e}{\tau}$ then $e \val$ or $e \step e'$
+\begin{itemize}
+    \item T-9: $e = \kwtrue$, $\tau = \kwbool$\\
+        by V-4 $e \val$
+    \item T-10: $e = \kwfalse$, $\tau = \kwbool$\\
+        by V-3 $e \val$
+    \item T-11: $e = \kwif{e_1}{e_2}{e_3}$\\
+        by inversion on rule T-11 $\typed{}{e_1}{\kwbool}$\\
+        by induction, either $e_1 \val$ or $e_1 \step e_1'$\\
+        by the Canonical Forms lemma for from Task~\ref{task-cf} $e_1$ can hold either \kwtrue or \kwfalse
+        \begin{itemize}
+            \item \kwtrue : S-16 confirms that $\kwif{\kwtrue}{e_2}{e_3} \step e_2$ and thus $e' = e_2$
+            \item \kwfalse : S-17 confirms that $\kwif{\kwfalse}{e_2}{e_3} \step e_3$ and thus $e' = e_3$
+        \end{itemize}
+
+\end{itemize}
+
 \end{document}

### 2021-10-04 HW1/hw1.tex
i should start on this

\documentclass{article}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath, amsthm, verbatim, enumerate, xcolor, mathpartir}

\input{macros}

\title{IIT CS595: Topics \& Applications in Programming Languages\\
  {\large Homework 1: Curry-Howard, Expressions continued}}
\author{Your name here}

\begin{document}
\maketitle

\section*{Logistics}
The submission and collaboration instructions (as well as the preference
for submissions typeset in LaTeX) are the same as for HW0.

\textbf{In particular:}
Submit your answers as a single .pdf or .doc file  on Blackboard
under
the correct assignment.

\section{The Curry-Howard Correspondence}

We discussed the Curry-Howard Correspondence in class as a correspondence
between STLC and Intuitionistic Propositional Logic (IPL).
%
We can also add universal quantification to both.
%
On the logic side, this gives us intuitionistic second-order logic,
where you can express
things like $\forall A. \forall B. A \land B \limp B \land A$
(``for all propositions $A$ and $B$, $A$ and $B$ implies $B$ and $A$'').
%
On the PL side, it gives us STLC with ``forall'' types (or, equivalently,
System F with notations for sums and products defined).
%
The type corresponding to the above logic statement would be
$\kwforall{A}{\kwforall{B}}{\kwarr{(\kwprod{A}{B})}{(\kwprod{B}{A})}}$---
the logical~$\forall$ and its type equivalent even conveniently use the
same symbol.

In Classical Logic, we have the bi-implication
$\forall A. \forall B. (A \limp B) \lbimp (\lnot A \lor B)$.
%
Only one direction of this implication is true constructively (i.e., in
intuitionistic second-order logic).

\begin{task}\label{task:ch-type}
  Write the STLC (with~$\forall$) type corresponding to
  the logical implication $\forall A. \forall B. (\lnot A \lor B) \limp (A \limp B)$.
\end{task}

\textbf{Answer:}

\begin{task}
  Prove the implication above by giving an STLC term of the type you wrote in
  Task~\ref{task:ch-type}.
\end{task}

\textbf{Answer:}

It turns out that, for any classically true proposition~$P$, $\lnot \lnot P$
is true constructively (even if~$P$ is not)\footnote{If you're wondering
  whether this fact itself is meaningful in programming languages through
  the Curry-Howard correspondence, Google ``continuation passing transformation''.}.

\begin{task}
  We've discussed that the Law of the Excluded Middle (now restated with
  $\forall$),
  $\forall A. A \lor \lnot A$, is
  not true constructively. Show that
  $\lnot \lnot (\forall A. A \lor \lnot A)$ \textit{is} true constructively by
  giving an STLC term of type

  \[\kwforall{\alpha}{\kwarr{(\kwarr{(\kwsum{\alpha}{(\kwarr{\alpha}{\kwvoid})})}{\kwvoid})}{\kwvoid}}\]

  \textbf{Hint: }
    The term will be of the form
    \[\kwtfun{\alpha}{\kwfun{f}{(\kwarr{(\kwsum{\alpha}{(\kwarr{\alpha}{\kwvoid})})}{\kwvoid})}{~e}}\]
    for some $e$ such that~$\ftyped{\alpha}{\ctx}{e}{\kwvoid}$,
    where~$\ctx = f:(\kwarr{(\kwsum{\alpha}{(\kwarr{\alpha}{\kwvoid})})}{\kwvoid})$.
    Can you construct terms~$e_1$ and $e_2$ such that
    $\ftyped{\alpha}{\ctx}{e_1}{\kwarr{(\kwarr{\alpha}{\kwvoid})}{\kwvoid}}$
    and~$\ftyped{\alpha}{\ctx}{e_2}{\kwarr{\alpha}{\kwvoid}}$?
    How can you then use~$e_1$ and~$e_2$ to construct the function body,~$e$?
\end{task}

\textbf{Answer:}

\section{Division by zero}

In the past, we've used type systems and the ideas of type safety (progress
and preservation) to prevent run-time type errors.
%
Here, we'll use a type system to prevent another run-time error: division by
zero.
%
Consider a language we'll call~\ezlang{}, whose syntax and dynamics are below.

{
  \centering
  \[
  \begin{array}{l l l}
    \tau & \bnfdef & \kwnz \bnfalt \kwmz\\
    e & \bnfdef &
    \kwn{n} \bnfalt
    e + e \bnfalt
    e - e \bnfalt
    e * e \bnfalt
    e / e
  \end{array}
  \]
}

{
  \centering
  \def \MathparLineskip {\lineskip=0.43cm}
  \begin{mathpar}

    \Rule{V-1}
         {\strut}
         {\kwn{n} \val}
    \and
    \Rule{S-1}
         {e_1 \step e_1'\\
         \text{\textit{op}} \in \{+, -, *, /\}}
         {e_1 \mathbin{\text{\textit{op}}} e_2 \step
           e_1' \mathbin{\text{\textit{op}}} e_2}
    \and
    \Rule{S-2}
         {e_2 \step e_2'\\
         \text{\textit{op}} \in \{+, -, *, /\}}
         {\kwn{n_1} \mathbin{\text{\textit{op}}} e_2 \step
           \kwn{n_1} \mathbin{\text{\textit{op}}} e_2'}
   \and
   \Rule{S-3}
        {\strut}
        {\kwn{n_1} + \kwn{n_2} \step \kwn{n_1 + n_2}}
   \and
   \Rule{S-4}
        {\strut}
        {\kwn{n_1} - \kwn{n_2} \step \kwn{n_1 - n_2}}
   \and
   \Rule{S-5}
        {\strut}
        {\kwn{n_1} * \kwn{n_2} \step \kwn{n_1 n_2}}
   \and     
   \Rule{S-6}
        {n_2 \neq 0}
        {\kwn{n_1} / \kwn{n_2} \step \kwn{n_1 / n_2}}
  \end{mathpar}
}

This is basically our expression language {\elang} from class
(without~$\kw{let}$), but with only integers and with subtraction,
multiplication and (integer) division added.
%
To save space and time, we have only two search rules: rules (S-1) and
(S-2) apply for all~4 arithmetic operations in place of \textit{op}.
%
Note that rule (S-6), as we'd hope, allows division only if~$\kwn{n_2}$ is
nonzero.
%
That means that a division by zero, e.g.,~$\kwn{1} / \kwn{0}$, is a ``stuck''
expression: it's not a value and it can't step.
%
So in order to make Progress and Preservation hold, we need a type system
that rules out division by zero.
%
The {\ezlang} language has only integers, so we don't need types to
distinguish integers from strings, but we still have two
types:~$\kwnz$ (``not zero'') and~$\kwmz$ (``maybe zero''), which we'll use to
check whether an expression {\em might} evaluate to zero: the idea will be
that if~$\etyped{e}{\kwnz}$, then~$e$ definitely does not evaluate to zero
(that is, if~$e \step^* \kwn{n}$, then~$n \neq 0$).
%
We'll only allow division by expressions that definitely won't evaluate to
zero.

\begin{task}
  Fill in the blanks in the typing rules for {\ezlang}.
  %
  The first two are done for you: any integer can have type~$\kwmz$, but
  it can only have~$\kwnz$ if it's not 0.
  %
  There are two rules for multiplication, depending on whether or not both
  operands are nonzero (think about why).

  \textbf{Hint:} Think about, for each operation (addition, subtraction,
  multiplication, division), whether performing the
  operation on two integers
  can result in zero, depending on whether the two operands can be zero.
  {\em Integers can be negative.}

\end{task}

%% Lucky you, you're using the template: you can just replace each
%% \blank below with the answer.
{
  \centering
  \def \MathparLineskip {\lineskip=0.43cm}
  \begin{mathpar}

    \Rule{T-1}
         {\strut}
         {\etyped{\kwn{n}}{\kwmz}}
    \and
    \Rule{T-2}
         {n \neq 0}
         {\etyped{\kwn{n}}{\kwnz}}
    \and
    \Rule{T-3}
         {\etyped{e_1}{\kwmz}\\
           \etyped{e_2}{\kwmz}}
         {\etyped{e_1 + e_2}{\blank}}
    \and
    \Rule{T-4}
         {\etyped{e_1}{\kwmz}\\
           \etyped{e_2}{\kwmz}}
         {\etyped{e_1 - e_2}{\blank}}
    \and
    \Rule{T-5}
         {\etyped{e_1}{\kwnz}\\
           \etyped{e_2}{\kwnz}}
         {\etyped{e_1 * e_2}{\blank}}
    \and
    \Rule{T-6}
         {\etyped{e_1}{\kwmz}\\
           \etyped{e_2}{\kwmz}}
         {\etyped{e_1 * e_2}{\blank}}
    \and
    \Rule{T-7}
         {\etyped{e_1}{\kwmz}\\
           \etyped{e_2}{\blank}}
         {\etyped{e_1 / e_2}{\blank}}
  \end{mathpar}
}

You may find it concerning that, for example, there's no rule for
$e_1 + e_2$ when~$e_1$ has type~$\kwnz$ and~$e_2$ has type~$\kwmz$.
%
This actually isn't a problem, because\footnote{This is, in fact, a form
  of {\em subtyping}, which you may be familiar with from languages like Java,
  where whenever something has one type, it also has a more general type.
  We'll study this in more detail later in the course.}:

\begin{thm}\label{thm:subtyping}
  For any expression~$e$, if~$\etyped{e}{\kwnz}$ then~$\etyped{e}{\kwmz}$.
\end{thm}

\begin{task}
  Prove Theorem~\ref{thm:subtyping} by induction on the derivation of
  $\etyped{e}{\kwnz}$.

  \textbf{Hint:} If you've set up the typing rules correctly, there should
  only be two cases. If a typing rule has a conclusion of the form
  $\etyped{e}{\kwmz}$, we don't need to consider it because it can't be used
  to derive our assumption~$\etyped{e}{\kwnz}$.
\end{task}

\textbf{Answer:}

The Canonical Forms lemma is similar to before, but now includes the fact that
values of type~$\kwnz$ are not zero.

\begin{lemma}[Canonical Forms]
  If~$\etyped{e}{\tau}$ and~$e \val$, then:
  \begin{enumerate}
  \item If~$\tau = \kwnz$, then~$e = \kwn{n}$ for some~$n \neq 0$.
  \item If~$\tau = \kwmz$, then~$e = \kwn{n}$ for some~$n$.
  \end{enumerate}
\end{lemma}

Here are the statements of Progress and Preservation for {\ezlang}.

\begin{thm}[Progress]
  If~$\etyped{e}{\tau}$ then~$e \val$ or there exists~$e'$ such that
  $e \step e'$.
\end{thm}

\begin{thm}[Preservation]
  If~$\etyped{e}{\tau}$ and~$e \step e'$ then~$\etyped{e'}{\tau}$.
\end{thm}

\begin{task}
  Prove the Progress and Preservation theorems for {\ezlang}.

  \textbf{Hints/Comments -- you'll want to read these.}
  \begin{itemize}
  \item For Progress, you only need to prove the cases for (T-1), (T-2) and
    (T-7)\footnote{This is {\em not} for the same reason that you only need
    to prove two cases for Theorem~\ref{thm:subtyping}; the other cases
    for Progress need to be proven too, this is just me being nice and not
    making you do them :)}.
  \item For Preservation, you only need to prove the cases for (S-3),
    (S-4), (S-5) and (S-6)\footnote{Same.}.
  \item In the (S-5) case for Preservation, if you do inversion on the typing
    rules, there will actually be {\em two} rules you need to consider in your
    inversion and thus, two cases (think about why).
  \item You can use without proof any (true!) facts of basic arithmetic.
  \end{itemize}
\end{task}

\textbf{Answer:}

\begin{task}
  Give an expression in {\ezlang} that {\em does not} perform a division by
  zero, but is still {\em not well-typed} by the above typing rules.
  Suggest a change, addition, refinement, etc., to the type system that could
  make your expression well-typed (but would still rule out expressions that
  divide by zero).
  You don't need to actually implement your suggestion, or even argue that
  it's possible or practical, and it's fine if your solution still rules out
  some ``OK'' programs (programs that don't divide by zero).
  This is just to get you to start thinking about
  how to get what you want with type systems.
\end{task}

\textbf{Answer:}

\end{document}

### 2021-10-08 HW1/hw1.tex
hw1

@@
 \usepackage{amsmath, amsthm, verbatim, enumerate, xcolor, mathpartir}
+\usepackage{amsfonts} % for "\mathbb" macro
 
@@
 \textbf{Answer:}
+\kwforall {\alpha} {\kwforall {\beta} {\kwarr{(\kwsum{(\kwarr{\alpha}{\kwvoid})}{\beta})}({\kwarr{\alpha}{\beta}})}}
 
@@
 \textbf{Answer:}
+% this is probably not right...
+\kwtfun{\alpha}{\kwtfun{\beta}{
+    \\\hspace*{10mm} 
+    \kwfun{b}{\beta}{
+        \\\hspace*{15mm}
+        \kwfun{l}{\kwsum{(\kwarr{\alpha}{\kwvoid})}{\beta}}{
+            \\\hspace*{20mm} 
+            \kwfun{a}{\alpha}{
+                \\\hspace*{25mm} 
+                \kwcase{l}
+                    {x} {b}
+                    {y} {y}}}}}}
+
+When invoked with a value of type $\beta$, the above term will have the type specified in task 1.\\
 
@@
 \textbf{Answer:}
+    \kwtfun{\alpha}{
+        \kwfun{f}{\kwarr {(\kwsum{\alpha}{(\kwarr{\alpha}{\kwvoid})})}{\kwvoid}}{
+        \kwapp{f}{(\kwfun{a}{\kwsum{\alpha}{(\kwarr{\alpha}{\kwvoid})}}{
+            \kwcase{a}{x}{\kwapp{f}{x}}{y}{\kwapp{f}{y}}})}}}
 
@@
            \etyped{e_2}{\kwmz}}
-         {\etyped{e_1 + e_2}{\blank}}
+         {\etyped{e_1 + e_2}{\kwmz}}
     \and
@@
            \etyped{e_2}{\kwmz}}
-         {\etyped{e_1 - e_2}{\blank}}
+         {\etyped{e_1 - e_2}{\kwmz}}
     \and
@@
            \etyped{e_2}{\kwnz}}
-         {\etyped{e_1 * e_2}{\blank}}
+         {\etyped{e_1 * e_2}{\kwnz}}
     \and
@@
            \etyped{e_2}{\kwmz}}
-         {\etyped{e_1 * e_2}{\blank}}
+         {\etyped{e_1 * e_2}{\kwmz}}
     \and
@@
          {\etyped{e_1}{\kwmz}\\
-           \etyped{e_2}{\blank}}
-         {\etyped{e_1 / e_2}{\blank}}
+           \etyped{e_2}{\kwnz}}
+         {\etyped{e_1 / e_2}{\kwmz}}
   \end{mathpar}
@@
 \textbf{Answer:}
+\begin{itemize}
+    \item Rule (T-1): already of type \kwmz 
+    \item Rule (T-2): $e = \kwn{n}$ thus apply (T-1)
+    \item Rule (T-3): already of type \kwmz
+    \item Rule (T-4): already of type \kwmz
+    \item Rule (T-5): $e = e_1 * e_2$ \\
+        by inversion $\etyped{e_1}{\kwnz}$ and $\etyped{e_2}{\kwnz}$\\
+        by induction $\etyped{e_1}{\kwmz}$ and $\etyped{e_2}{\kwmz}$\\
+        apply rule (T-6) giving $\etyped{e}{\kwmz}$
+    \item Rule (T-6): already of type \kwmz
+    \item Rule (T-7): already of type \kwmz
+\end{itemize}
 
@@
 
+\textbf{Progress:}
+\begin{itemize}
+    \item (T-1): $e = \kwn{n}$ and $\tau = \kwmz$ \\
+        according to (V-1) $e \val$.
+    \item (T-2): $e = \kwn{n}$ and $\tau = \kwnz$ \\
+        according to (V-1) $e \val$.
+    \item (T-7): $e = e_1 / e_2$ and $\tau = \kwmz$ \\
+        by inversion $\etyped{e_2}{\kwnz}$\\
+        by inversion on (T-2) $e_2 \neq 0$\\
+        by induction $e_1$ and $e_2$ either step or are values\\ 
+        if they are values we can use the canonical forms lemma to show that e steps via (S-6)\\
+        if they step then the expression steps, proving the proposition.
+\end{itemize}
+
+\textbf{Preservation:}
+\begin{itemize}
+    \item (S-3): $e = e_1 + e_2$ and $\tau = \kwmz$\\
+        by inversion on (T-3) $\etyped{e_1}{\kwmz}$ and $\etyped{e_2}{\kwmz}$
+        (note that theorem 1 says this includes \kwnz)\\
+        thus by induction on rule (T-3) $\etyped{e'}{\kwmz}$
+    \item (S-4): $e = e_1 - e_2$ and $\tau = \kwmz$\\
+        by inversion on (T-3) $\etyped{e_1}{\kwmz}$ and $\etyped{e_2}{\kwmz}$
+        (note that theorem 1 says this includes \kwnz)\\
+        thus by induction on rule (T-3) $\etyped{e'}{\kwmz}$
+    \item (S-5): $e = e_1 * e_2$
+        there are two cases to consider
+        \begin{enumerate}
+            \item when $\etyped{e_1}{\kwnz}$ and $\etyped{e_2}{\kwnz}$, $\tau = \kwnz$\\
+                by inversion on (T-5) $\etyped{e_1}{\kwnz}$ and $\etyped{e_2}{\kwnz}$\\
+                thus by induction $\etyped{e}{\kwnz}$
+            \item when $\etyped{e_1}{\kwmz}$ or $\etyped{e_2}{\kwmz}$, $\tau = \kwmz$\\
+                by inversion on (T-6) $\etyped{e_1}{\kwmz}$ and $\etyped{e_2}{\kwmz}$ 
+                (note that theorem 1 says this includes \kwnz)\\
+                thus by induction $\etyped{e'}{\kwmz}$
+        \end{enumerate}
+    \item (S-6): $e = e_1 / e_2$ and $\tau = \kwmz$\\
+        by inversion on (T-7) $\etyped{e_1}{\kwmz}$ and $\etyped{e_2}{\kwnz}$\\
+        thus by induction $\etyped{e'}{\kwmz}$
+\end{itemize}
+
 \begin{task}
@@
 
+\textbf{Example Expression:} $555 / (1 + 4)$ \\
+We could extend the type system, allowing in addition to tracking zero-ness we also track the signedness of numbers.
+So we could have some new addition rules like the following shorthand for the type rules where u* means unsigned ($\ge 0$) and s* means signed ($\mathbb{Z}$)
+
+\begin{enumerate}
+    \item umz + umz : umz
+    \item umz + unz : unz
+    \item unz + umz : unz
+    \item unz + unz : unz\\
+
+    \item smz + smz : smz
+    \item smz + snz : smz
+    \item snz + smz : smz
+    \item snz + snz : smz\\
+
+    \item depending on implementation, it might also make sense to add negative signed types
+\end{enumerate}
+
+So in the example we could have used rule 4 to verify that the denominator would not be zero.
+This would still not work with all valid expressions but at least this could potentially be done at compile time.
+
+Also partial evaluation could potentially help.
+
 \end{document}

### 2021-10-27 HW2/hw2.tex
hw2 assigned

\documentclass{article}
\usepackage[left=1in,right=1in,top=1in,bottom=1in]{geometry}
\usepackage{amsmath, amsthm, verbatim, enumerate, xcolor, mathpartir}

\input{macros}

\title{IIT CS595: Topics \& Applications in Programming Languages\\
  {\large Homework 2: Memory and Cost Semantics}}
\author{Your Name Here}
\date{Out: Tuesday, Oct. 19\\
  Due: Wednesday, Nov. 3 11:59pm CDT\\
  %\vspace{0.5cm}
  %\updated{Sep. 15}
}

\begin{document}
\maketitle

%\textbf{This assignment contains~{\numwtasks} written tasks and~{\numptasks}
%  programming tasks, for a total of~{\totpoints} points.}

\section*{Logistics}
The submission and collaboration instructions (as well as the preference
for submissions typeset in LaTeX) are the same as for HW0 and HW1.

\textbf{In particular:}
Submit your answers as a single .pdf or .doc file  on Blackboard
under
the correct assignment.

\section{Allocation and Memory Use}

As we discussed in class, cost semantics can be useful for more than just
counting the number of steps a program will take to execute.
%
In this assignment, we'll use them to track memory usage.
%
Consider this new variant of STLC:

\[
\begin{array}{r l l}
  \tau & \bnfdef & \kwunitt \bnfalt
  \kwarr{\tau}{\tau} \bnfalt
  \kwprod{\tau}{\tau}\\
  e & \bnfdef & x \bnfalt
  \kwunit \bnfalt
  \kwfun{x}{\tau}{e} \bnfalt
  \kwapp{e}{e} \bnfalt
  \kwpair{e}{e} \bnfalt
  \kwvpair{e}{e} \bnfalt
  \kwfst{e} \bnfalt
  \kwsnd{e}\\
\end{array}
\]

We now have two forms for pairs: $\kwpair{e_1}{e_2}$ and $\kwvpair{e_1}{e_2}$.
The latter form, $\kwvpair{e_1}{e_2}$, is used for a pair that has fully
evaluated and been allocated in memory (this models how structures like pairs
are executed in real languages; when they are constructed, memory for the
structure is allocated, e.g., using \texttt{malloc}).
%
This is made clearer in the dynamic semantics.
%
Like when we covered imperative languages, the state has a store~$\store$.
%
These stores are actually just lists of all of the pairs that have been
produced/allocated.
%
We'll use~$\stjoin{\store_1}{\store_2}$ to combine stores
and~$|\store|$ to refer to the size of a store (i.e., the length of the list).

The key different rules are (V-3) and (S-6).
%
Note from the (V-\#) rules that~$\kwpair{e_1}{e_2}$ is {\em not} a value,
even if~$e_1$ and~$e_2$ are values.
%
Instead, (S-6) says that when~$v_1$ and~$v_2$ are values,~$\kwpair{v_1}{v_2}$
takes an additional step to allocate the memory for the pair by adding it
to the store and steps to~$\kwvpair{v_1}{v_2}$, which is now a value%
\footnote{If we really wanted to faithfully model how this works in real
  languages, we would return a representation of the location in memory where
  the pair is stored (basically a pointer), and look it up in memory if we
  need to get the components of the pair later. The way we're doing it here
  is a little contrived but avoids a bunch of cumbersome notation I didn't
  want to burden you with.}.
%
For the same reason,~$\kwfst{e}$ and~$\kwsnd{e}$ evaluate~$e$ until
it looks like~$\kwvpair{v_1}{v_2}$: the pair must be allocated before we can
take its components and if we have a pair~$\kwvpair{v_1}{v_2}$, we know it's
been allocated.

Make sure you understand the language and the distinctions made above
before continuing.
%
If you have questions, feel free to ask for clarifications on Discord or
at office hours.

{
  \centering
  \def \MathparLineskip {\lineskip=0.43cm}
  \begin{mathpar}
    \Rule{V-1}
         {\strut}
         {\kwunit\val}
    \and
    \Rule{V-2}
         {\strut}
         {\kwfun{x}{\tau}{e}\val}
    \and
    \Rule{V-3}
         {v_1 \val\\
           v_2 \val}
         {\kwvpair{v_1}{v_2} \val}
    \and
    \Rule{S-1}
         {\cfg{\store}{e_1} \step \cfg{\store'}{e_1'}}
         {\cfg{\store}{\kwapp{e_1}{e_2}}
           \step
           \cfg{\store'}{\kwapp{e_1'}{e_2}}}
    \and
    \Rule{S-2}
         {\cfg{\store}{e_2} \step \cfg{\store'}{e_2'}}
         {\cfg{\store}{\kwapp{(\kwfun{x}{\tau}{e})}{e_2}}
           \step
           \cfg{\store'}{\kwapp{(\kwfun{x}{\tau}{e})}{e_2'}}}
    \and
    \Rule{S-3}
         {v\val}
         {\cfg{\store}{\kwapp{(\kwfun{x}{\tau}{e})}{v}}
           \step
           \cfg{\store}{\sub{v}{x}{e}}}
    \and
    \Rule{S-4}
         {\cfg{\store}{e_1} \step \cfg{\store'}{e_1'}}
         {\cfg{\store}{\kwpair{e_1}{e_2}}
           \step
           \cfg{\store'}{\kwpair{e_1'}{e_2}}}
    \and
    \Rule{S-5}
         {v_1\val\\
           \cfg{\store}{e_2} \step \cfg{\store'}{e_2'}}
         {\cfg{\store}{\kwpair{v_1}{e_2}}
           \step
           \cfg{\store'}{\kwpair{v_1}{e_2'}}}
    \and
    \Rule{S-6}
         {v_1\val\\
           v_2\val}
         {\cfg{\store}{\kwpair{v_1}{v_2}}
           \step
           \cfg{\stjoin{\store}{\stsing{\kwvpair{v_1}{v_2}}}}
               {\kwvpair{v_1}{v_2}}
         }
    \and
    \Rule{S-7}
         {\cfg{\store}{e} \step \cfg{\store'}{e'}}
         {\cfg{\store}{\kwfst{e}} \step \cfg{\store'}{\kwfst{e'}}}
    \and
    \Rule{S-8}
         {\strut}
         {\cfg{\store}{\kwfst{\kwvpair{v_1}{v_2}}}
           \step \cfg{\store}{v_1}}
    \and
    \Rule{S-9}
         {\cfg{\store}{e} \step \cfg{\store'}{e'}}
         {\cfg{\store}{\kwsnd{e}} \step \cfg{\store'}{\kwsnd{e'}}}
    \and
    \Rule{S-10}
         {\strut}
         {\cfg{\store}{\kwsnd{\kwvpair{v_1}{v_2}}}
         \step \cfg{\store}{v_2}}
\end{mathpar}
}

The cost semantics to track memory usage then should correspond not to the
number of steps a program will take to execute, but the number of pairs it
allocates, that is, the length of~$\store$ when it is done.
%
I've filed out the big-step semantics for the language for you.
%
Note that the big-step semantics {\em don't} carry around a $\store$; that's
too low-level a detail for the abstract big-step semantics.

{
  \centering
  \def \MathparLineskip {\lineskip=0.43cm}
  \begin{mathpar}
    \Rule{E-1}
         {v\val}
         {\eval[0]{v}{v}}
    \and
    \Rule{E-2}
         {\eval[\blank]{e_1}{\kwfun{x}{\tau}{e}}\\
           \eval[\blank]{e_2}{v}\\
           \eval[\blank]{\sub{v}{x}{e}}{v'}}
         {\eval[\blank]{\kwapp{e_1}{e_2}}{v'}}
    \and
    \Rule{E-3}
         {\eval[\blank]{e_1}{v_1}\\
           \eval[\blank]{e_2}{v_2}
         }
         {\eval[\blank]{\kwpair{e_1}{e_2}}{\kwvpair{v_1}{v_2}}}
    \and
    \Rule{E-4}
         {\eval[\blank]{e}{\kwvpair{v_1}{v_2}}}
         {\eval[\blank]{\kwfst{e}}{v_1}}
    \and
    \Rule{E-5}
         {\eval[\blank]{e}{\kwvpair{v_1}{v_2}}}
         {\eval[\blank]{\kwsnd{e}}{v_2}}
  \end{mathpar}
}

\begin{task}
  Fill in the blanks in the big-step semantics to turn it into a cost semantics
  for tracking memory usage.
  %
  If you're having trouble deciding exactly what operations to ``charge'' for,
  look ahead at the theorems we want to prove about the cost semantics.
  %
  (Conversely, if you're having trouble proving the results later, you
  might want to come back and re-evaluate whether you need to change the
  cost semantics.)
\end{task}

We'll now prove that the cost semantics corresponds to the memory used by
the small-step semantics.
%
That is:
\[
\eval[n]{e}{v} \Leftrightarrow (\cfg{\estore}{e} \step^* \cfg{\store}{v}
\text{~and~} n = |\store|)
\]
where~$|\store|$ is the size of the store and~$\estore$ is the empty store.
%
As you'd expect, $|\estore| = 0$.

We prove this in several pieces.
%
Recall that the forward direction involves proving several annoying lemmas.
%
Here's the first annoying lemma.
%
It says that if~$e \step^* e'$ and $e$ is actually a part of a bigger
expression, that bigger expression also steps appropriately.
%
It has 6 parts for each kind of ``bigger expression.''

\begin{lemma}\label{lem:compose-exps}
  \begin{enumerate}
  \item
    If~$\cfg{\store}{e_1} \step^* \cfg{\store'}{e_1'}$
    then~$\cfg{\store}{\kwapp{e_1}{e_2}} \step^*
    \cfg{\store'}{\kwapp{e_1'}{e_2}}$.
  \item
    If~$\cfg{\store}{e_2} \step^* \cfg{\store'}{e_2'}$
    then~$\cfg{\store}{\kwapp{e_1}{e_2}} \step^*
    \cfg{\store'}{\kwapp{e_1}{e_2'}}$.
  \item
    If~$\cfg{\store}{e_1} \step^* \cfg{\store'}{e_1'}$
    then~$\cfg{\store}{\kwpair{e_1}{e_2}} \step^*
    \cfg{\store'}{\kwpair{e_1'}{e_2}}$.
  \item
    If~$\cfg{\store}{e_2} \step^* \cfg{\store'}{e_2'}$
    then~$\cfg{\store}{\kwpair{e_1}{e_2}} \step^*
    \cfg{\store'}{\kwpair{e_1}{e_2'}}$.
  \item
    If~$\cfg{\store}{e} \step^* \cfg{\store'}{e'}$
    then~$\cfg{\store}{\kwfst{e}} \step^*
    \cfg{\store'}{\kwfst{e}}$.
  \item
    If~$\cfg{\store}{e} \step^* \cfg{\store'}{e'}$
    then~$\cfg{\store}{\kwsnd{e}} \step^*
    \cfg{\store'}{\kwsnd{e}}$.
  \end{enumerate}
\end{lemma}

To do the proof, recall that the~$\step^*$ judgment is also defined
inductively:

{
  \centering
  \begin{mathpar}
    \Rule{M-1}
         {\strut}
         {\cfg{\store}{e} \step^* \cfg{\store}{e}}
    \and
    \Rule{M-2}
         {\cfg{\store}{e} \step \cfg{\store''}{e''}\\
           \cfg{\store''}{e''} \step^* \cfg{\store'}{e'}}
         {\cfg{\store}{e} \step^* \cfg{\store'}{e'}}
  \end{mathpar}
}

The parts are all reasonably similar.
I'll prove one, you'll prove another.

\begin{proof}
  \begin{enumerate}
  \item
    By induction on the derivation of
    $\cfg{\store}{e_1} \step^* \cfg{\store'}{e_1'}$.
    \begin{itemize}
    \item (M-1). Then~$e_1 = e_1'$ and~$\store = \store'$.
      By (M-1),
      $\cfg{\store}{\kwapp{e_1}{e_2}} \step^*
      \cfg{\store}{\kwapp{e_1}{e_2}}$.
    \item (M-2)
      Then~$\cfg{\store}{e_1} \step \cfg{\store''}{e_1''}$
      and $\cfg{\store''}{e_1''} \step^* \cfg{\store'}{e_1'}$.
      By (S-1),
      $\cfg{\store}{\kwapp{e_1}{e_2}} \step \cfg{\store''}{\kwapp{e_1''}{e_2}}$.
      By induction,
      $\cfg{\store''}{\kwapp{e_1''}{e_2}} \step^* \cfg{\store'}{\kwapp{e_1'}{e_2}}$
      so by (M-2),
      $\cfg{\store}{\kwapp{e_1}{e_2}} \step^* \cfg{\store'}{\kwapp{e_1'}{e_2}}$.
    \end{itemize}
  \end{enumerate}
\end{proof}

\begin{task}
  Prove part 5 of Lemma~\ref{lem:compose-exps}.
\end{task}

\begin{proof}
  Your proof here.
\end{proof}

We also need to know that if we have two sequences of steps, we can ``glue''
them together.

\begin{lemma}\label{lem:compose-steps}
  If~$\cfg{\store}{e} \step^* \cfg{\store'}{e'}$
  and~$\cfg{\store'}{e'} \step^* \cfg{\store''}{e''}$
  then~$\cfg{\store}{e} \step^* \cfg{\store''}{e''}$.
\end{lemma}

\begin{task}
  Prove Lemma~\ref{lem:compose-steps}.
  You have two choices of what to induct on. Take an educated guess based on
  the rules you'll have to apply to conclude that
  $\cfg{\store}{e} \step^* \cfg{\store''}{e''}$.
  If the proof isn't working out, go back and try inducting on the other
  assumption.
\end{task}

\begin{proof}
  Your proof here.
\end{proof}

Now we can prove the forward direction of the theorem.
%
Note that we don't actually prove
that~$\cfg{\estore}{e} \step^* \cfg{\store}{v}$.
%
We actually need to be able to start with a non-empty store in order to
have a strong enough induction hypothesis.
%
Think about why.
%
We also prove that when~$e$ steps to~$v$ starting with store~$\store$,
the resulting store is strictly an extension of~$\store$: that is,
it's~$\stjoin{\store}{\store'}$, where~$\store'$ contains what we've added
to the store while stepping.
%
This means you don't have to worry about tracking things getting de-allocated
or changed (we don't have garbage collection.)

\begin{thm}\label{thm:forward}
  If $\eval[n]{e}{v}$ then for all~$\store$, it is the case that
  $\cfg{\store}{e} \step^* \cfg{\stjoin{\store}{\store'}}{v}$
  and $|\store'| = n$.
\end{thm}

I'll prove one case as an example.

\begin{proof}
  By induction on the derivation of $\eval[n]{e}{v}$.
  \begin{itemize}

  \item (E-2).
    Then~$e = \kwapp{e_1}{e_2}$
    and~$\eval[n_1]{e_1}{\kwfun{x}{\tau}{e}}$
    and~$\eval[n_2]{e_2}{v}$
    and~$\eval[n_3]{\sub{v}{x}{e}}{v'}$ and~$n = n_1 + n_2 + n_3$.
    By induction,
    \[
    \cfg{\store}{e_1} \step^*
    \cfg{\stjoin{\store}{\store_1}}{\kwfun{x}{\tau}{e}}
    \text{~and~}\cfg{\stjoin{\store}{\store_1}}{e_2} \step^*
    \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v}
    \text{~and~}\cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}
    {\sub{v}{x}{e}} \step^*
    \cfg{\stjoin{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\store_3}}
        {v'}
    \]
    and
    \[|\store_1| = n_1\text{~and~}|\store_2| = n_2\text{~and~}|\store_3| = n_3\]
    By Lemma~\ref{lem:compose-exps},
    $\cfg{\store}{e} \step^*
    \cfg{\stjoin{\store}{\store_1}}{\kwapp{(\kwfun{x}{\tau}{e})}{e_2}}$
    and~$\cfg{\stjoin{\store}{\store_1}}{\kwapp{(\kwfun{x}{\tau}{e})}{e_2}}
    \step^*
    \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}
          {\kwapp{(\kwfun{x}{\tau}{e})}{v}}$.\\
          By Lemma~\ref{lem:compose-steps},
      $\cfg{\store}{e} \step^*
        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}
            {(\kwapp{\kwfun{x}{\tau}{e})}{v}}$.\\
      By (S-3) and Lemma~\ref{lem:compose-steps},
      $\cfg{\store}{e} \step^*
      \cfg{\stjoin{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\store_3}}
          {v'}$.
          We have
      $|\stjoin{\store_1}{\stjoin{\store_2}{\stsing{\kwvpair{v_1}{v_2}}}}| =
      |\store_1| + |\store_2| + |\store_3| = n$.

\item (E-1).
\item (E-3).
\item (E-4).
\item (E-5).
  \end{itemize}
\end{proof}

\begin{task}
  Prove the remaining cases of Theorem~\ref{thm:forward}.
  There are many similarities between the cases, but some small, important
  differences. Copying and pasting between cases is allowed and encouraged,
  but make sure you know what to change!

  \textbf{Note:} I actually proved the {\em second} case; you still have
  to go back and prove (E-1) as well as (E-3) through (E-5).
\end{task}

The reverse direction is pretty straightforward given a key lemma introduced
in class.
%
The lemma says that if~$\eval[n]{e'}{v}$ and~$e$ steps to~$e'$,
then~$\eval[n+k]{e}{v}$, where~$k$ counts any pairs allocated by that step.

\begin{lemma}\label{lem:closed}
  If~$\eval[n]{e'}{v}$
  and~$\cfg{\store}{e} \step \cfg{\stjoin{\store}{\store'}}{e'}$
  then~$\eval[n+k]{e}{v}$
  and~$|\store'| = k$.
\end{lemma}
  \begin{proof}
    By induction on the derivation of 
    $\cfg{\store}{e} \step \cfg{\stjoin{\store}{\store'}}{e'}$.
    \begin{itemize}
    \item (S-1)
      Then~$e = \kwapp{e_1}{e_2}$ and~$e' = \kwapp{e_1'}{e_2}$.
      By inversion on (E-2),
      $\eval[n_1]{e_1'}{\kwfun{x}{\tau}{e}}$
      and~$\eval[n_2]{e_2}{v}$
      and~$\eval[n_3]{\sub{v}{x}{e}}{v'}$
      and~$n = n_1 + n_2 + n_3$.
      By induction,
      $\eval[n_1 + k]{e_1}{\kwfun{x}{\tau}{e}}$
      and~$|\store'| = k$.
      By (E-2),
      $\eval[n+k]{e}{v}$.

    \item Your other cases here...
    \end{itemize}
  \end{proof}

\begin{task}
  Prove the remaining cases of Lemma~\ref{lem:closed}.
  There are many similarities between the cases, but some small, important
  differences. Copying and pasting between cases is allowed and encouraged,
  but make sure you know what to change!
\end{task}

Given that lemma, the theorem is pretty straightforward to prove.
%
I'll just do it so you can be done with the homework.

\begin{thm}
  If $\cfg{\store}{e} \step^* \cfg{\stjoin{\store}{\store'}}{v}$ and~$v \val$,
  then~$\eval[n]{e}{v}$ and~$n = |\store'|$.
\end{thm}
\begin{proof}
  By induction on the derivation of $\cfg{\estore}{e} \step^* \cfg{\store}{v}$.
  If~$e = v,$ then $\store = \estore$ and by (E-1), $\eval[0]{v}{v}$. 
  Otherwise,
  $\cfg{\estore}{e} \step \cfg{\stjoin{\estore}{\store_1}}{e'}
  \step^* \cfg{\stjoin{\estore}{\stjoin{\store_1}{\store_2}}}{v}$.
  By induction,~$\eval[n]{e'}{v}$
  and~$n = |\store_2|$.
  By Lemma~\ref{lem:closed}, $\eval[n + k]{e}{v}$
  where~$k = |\store_1|$.
  We have~$n + k = |\stjoin{\store_1}{\store_2}|$.
\end{proof}

\end{document}

### 2021-11-06 HW2/hw2.tex
finish hw2

@@
   {\large Homework 2: Memory and Cost Semantics}}
-\author{Your Name Here}
+\author{Dustin Van Tate Testa}
 \date{Out: Tuesday, Oct. 19\\
@@
     \Rule{E-2}
-         {\eval[\blank]{e_1}{\kwfun{x}{\tau}{e}}\\
-           \eval[\blank]{e_2}{v}\\
-           \eval[\blank]{\sub{v}{x}{e}}{v'}}
-         {\eval[\blank]{\kwapp{e_1}{e_2}}{v'}}
+         {\eval[n_1]{e_1}{\kwfun{x}{\tau}{e}}\\
+           \eval[n_2]{e_2}{v}\\
+           \eval[n_3]{\sub{v}{x}{e}}{v'}}
+         {\eval[n_1 + n_2 + n_3]{\kwapp{e_1}{e_2}}{v'}}
     \and
     \Rule{E-3}
-         {\eval[\blank]{e_1}{v_1}\\
-           \eval[\blank]{e_2}{v_2}
+         {\eval[n_1]{e_1}{v_1}\\
+           \eval[n_2]{e_2}{v_2}
          }
-         {\eval[\blank]{\kwpair{e_1}{e_2}}{\kwvpair{v_1}{v_2}}}
+         {\eval[n_1 + n_2 + 1]{\kwpair{e_1}{e_2}}{\kwvpair{v_1}{v_2}}}
     \and
     \Rule{E-4}
-         {\eval[\blank]{e}{\kwvpair{v_1}{v_2}}}
-         {\eval[\blank]{\kwfst{e}}{v_1}}
+         {\eval[n]{e}{\kwvpair{v_1}{v_2}}}
+         {\eval[n]{\kwfst{e}}{v_1}}
     \and
     \Rule{E-5}
-         {\eval[\blank]{e}{\kwvpair{v_1}{v_2}}}
-         {\eval[\blank]{\kwsnd{e}}{v_2}}
+         {\eval[n]{e}{\kwvpair{v_1}{v_2}}}
+         {\eval[n]{\kwsnd{e}}{v_2}}
   \end{mathpar}
@@
 \begin{proof}
-  Your proof here.
+  By induction on the derivation of
+    $\cfg{\store}{e} \step^* \cfg{\store'}{e'}$.
+    \begin{itemize}
+    \item (M-1). Then~$e = e'$ and~$\store = \store'$.
+      By (M-1),
+      $\cfg{\store}{\kwfst{e}} \step^*
+      \cfg{\store}{\kwfst{e}}$.
+    \item (M-2)
+      Then~$\cfg{\store}{e} \step \cfg{\store''}{e''}$ 
+      and $\cfg{\store''}{e''} \step^* \cfg{\store}{e'}$.\\
+      By (S-7), 
+      $\cfg{\store}{\kwfst{e}} \step \cfg{\store''}{\kwfst{e''}}$.
+      By induction,
+      $\cfg{\store''}{\kwfst{e''}} \step^* \cfg{\store'}{\kwfst{e'}}$
+      so by (M-2),
+      $\cfg{\store}{\kwfst{e}} \step^* \cfg{\store'}{\kwfst{e'}}$
+    \end{itemize}
 \end{proof}
@@
   Your proof here.
+  \begin{itemize}
+    \item by induction on derivation of the step judgement, either
+        \begin{itemize}
+            \item $e = e' = e''$ and $\store = \store' = \store''$:\\
+                in which case for both step judgements (M-1) would be applied
+
+            \item $e' = e''$ and $\store' = \store''$:\\
+                in which case (M-2) is applied to the first step and (M-1) applied after
+                % \begin{mathpar}
+                %     \inferrule*
+                %     {\inferrule* {$ $}{\cfg{\store}{e} \step \cfg{\store'''}{e'''}}\\
+                %       \inferrule* {$ $}{\cfg{\store'''}{e'''} \step^* \cfg{\store'}{e'}}}
+                %     {\cfg{\store}{e} \step^* \cfg{\store'}{e'}}
+                % \end{mathpar}
+            \item $e = e'$ and $\store = \store'$:\\
+                in which case (M-1) is applied to make the first step and then (M-2) after
+
+            \item all $e$ and all $\store$ values are unique:\\
+                in which case (M-2) is applied twice
+        \end{itemize}
+  \end{itemize}
 \end{proof}
@@
           We have
-      $|\stjoin{\store_1}{\stjoin{\store_2}{\stsing{\kwvpair{v_1}{v_2}}}}| =
+      $|\stjoin{\store_1}{\stjoin{\store_2}{\store_3}}| =
       |\store_1| + |\store_2| + |\store_3| = n$.
@@
 \item (E-1).
+Then $e \val$ and $\eval[0]{e}{v}$ and $e = v$ and $n = 0$ which by lemma 1 translates to $\cfg{\estore}{e} \step^* \cfg{\store}{e}$.
+Because values cannot step according to the defined small-step semantics rules (exhaustion of rules (V- 1..3)), the $\step^*$ judgement must follow rule (M-1), thus $\estore = \store$ and $|\store| = n = 0$.
+
 \item (E-3).
+    Then~$e = \kwpair{e_1}{e_2}$
+    and~$\eval[n_1]{e_1}{v_1}$
+    and~$\eval[n_2]{e_2}{v_2}$
+    and~$\eval[1]{\kwpair{v_1}{v_2}}{\kwvpair{v_1}{v_2}}$
+    and~$n = n_1 + n_2 + 1$.
+    By induction,
+    \[
+    \cfg{\store}{e_1} 
+    \step^* \cfg{\stjoin{\store}{\store_1}}{v_1}
+        \text{~and~}
+    \cfg{\store}{e_2}
+    \step^* \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_2}
+        \text{~and~}
+    \cfg{\store}{e_2} 
+    \step^* \cfg{\stjoin{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\store_3}}{v_2}
+    \]
+    and
+    \[|\store_1| = n_1\text{~and~}|\store_2| = n_2\text{~and~}|\store_3| = 1\]
+    By Lemma~\ref{lem:compose-exps},
+    $
+        \cfg{\store}{e} 
+        \step^* \cfg{\stjoin{\store}{\store_1}}{\kwpair{v_1}{e_1}}
+    $and~$
+        \cfg{\stjoin{\store}{\store_1}}{\kwpair{v_1}{e_1}}
+        \step^* \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\kwpair{v_1}{v_2}}
+    $.\\
+    By Lemma~\ref{lem:compose-steps},
+      $\cfg{\store}{e} 
+        \step^*
+         \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\kwpair{v_1}{v_2}}$.\\
+    By (S-6) and Lemma~\ref{lem:compose-steps},
+      $\cfg{\store}{e} \step^*
+      \cfg{\stjoin{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{\stsing{\kwvpair{v_1}{v_2}}}}
+          {\kwvpair{v_1}{v_2}}$.
+          We have
+      $|\stjoin{\store_1}{\stjoin{\store_2}{\stsing{\kwvpair{v_1}{v_2}}}}| =
+      |\store_1| + |\store_2| + |\stsing{\kwvpair{v_1}{v_2}}| = n$.
+
 \item (E-4).
+    Then~$e = \kwfst{e_1}$
+    and~$\eval[n_1]{e_1}{\kwvpair{v_1}{v_2}}$
+    and~$\eval[0]{e}{v_1}$
+    and~$n = n_1 + 0$.
+    By induction,
+    \[
+    \cfg{\store}{e_1} \step^* \cfg{\stjoin{\store}{\store_1}}{\kwvpair{v_1}{v_2}}
+    \text{~and~}
+    \cfg{\stjoin{\store}{\store_1}}{\kwfst{\kwvpair{v_1}{v_2}}} 
+        \step^* 
+        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_1}
+    \]
+    and
+    \[|\store_1| = |\store_2| = n_1 \text{~and~} n = n_1 \]
+    By Lemma~\ref{lem:compose-exps},
+    $
+        \cfg{\store}{e} \step^* \cfg{\stjoin{\store}{\store_1}}{\kwfst{\kwvpair{v_1}{v_2}}}
+    $~and~$
+        \cfg{\stjoin{\store}{\store_1}}{\kwfst{\kwvpair{v_1}{v_2}}} 
+        \step^* 
+        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_1}
+    $.\\
+    By Lemma~\ref{lem:compose-steps},
+      $\cfg{\store}{e} \step^*
+        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_1}$.\\
+    By (S-8) and Lemma~\ref{lem:compose-steps},
+      $\cfg{\store}{e} \step^*
+      \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_1}$.
+          We have
+      $|\stjoin{\store_1}{\store_2}| =
+      |\store_1| = n$.
+
 \item (E-5).
+    Then~$e = \kwsnd{e_1}$
+    and~$\eval[n_1]{e_1}{\kwvpair{v_1}{v_2}}$
+    and~$\eval[0]{e}{v_1}$
+    and~$n = n_1 + 0$.
+    By induction,
+    \[
+    \cfg{\store}{e_1} \step^* \cfg{\stjoin{\store}{\store_1}}{\kwvpair{v_1}{v_2}}
+    \text{~and~}
+    \cfg{\stjoin{\store}{\store_1}}{\kwsnd{\kwvpair{v_1}{v_2}}} 
+        \step^* 
+        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_2}
+    \]
+    and
+    \[|\store_1| = |\store_2| = n_1 \text{~and~} n = n_1 \]
+    By Lemma~\ref{lem:compose-exps},
+    $
+        \cfg{\store}{e} \step^* \cfg{\stjoin{\store}{\store_1}}{\kwsnd{\kwvpair{v_1}{v_2}}}
+    $~and~$
+        \cfg{\stjoin{\store}{\store_1}}{\kwsnd{\kwvpair{v_1}{v_2}}} 
+        \step^* 
+        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_2}
+    $.\\
+    By Lemma~\ref{lem:compose-steps},
+      $\cfg{\store}{e} \step^*
+        \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_2}$.\\
+    By (S-9) and Lemma~\ref{lem:compose-steps},
+      $\cfg{\store}{e} \step^*
+      \cfg{\stjoin{\stjoin{\store}{\store_1}}{\store_2}}{v_2}$.
+          We have
+      $|\stjoin{\store_1}{\store_2}| =
+      |\store_1| = n$.
+
   \end{itemize}
@@
       By (E-2),
+      $\eval[n+k]{e}{v'}$.
+
+    \item (S-2) Then~$e = \kwapp{\kwfun{x}{\tau}{e_1}}{e_2}$
+        and~$e' = \kwapp{\kwfun{x}{\tau}{e_1}}{e_2'}$.
+        By inversion on (E-2),
+        $\eval[0]{\kwfun{x}{\tau}{e_1}}{\kwfun{x}{\tau}{e_1}}$
+        and~$\eval[n_2]{e_2}{v'}$
+        and~$\eval[n_3]{\sub{v'}{x}{e_1}}{v}$
+        and~$n = 0 + n_2 + n_3$.
+        By induction,
+        $\eval[n' + k]{e_2}{v'}$
+        and~$|\store'| = k$.
+        By (E-2),
         $\eval[n+k]{e}{v}$.
 
-    \item Your other cases here...
+    \item (S-3) Then~$e = \kwapp{\kwfun{x}{\tau}{e_1}}{v'}$
+        and~$e' = \kwapp{\kwfun{x}{\tau}{e_1}}{e_2'}$
+        and~$v' \val$.
+        By inversion on (E-2),
+        $\eval[0]{\kwfun{x}{\tau}{e_1}}{\kwfun{x}{\tau}{e_1}}$
+        and~$\eval[0]{v'}{v'}$
+        and~$\eval[n_1]{\sub{v'}{x}{e_1}}{v}$
+        and~$n = 0 + 0 + n_1$.
+        By induction,
+        $\eval[n' + k]{\sub{v'}{x}{e_1}}{v}$
+        $|\store'| = k$.
+        By (E-2),
+        $\eval[n+k]{e}{v}$.
+
+    \item (S-4) Then~$e = \kwpair{e_1}{e_2}$
+        and~$e' = \kwpair{e_1'}{e_2}$.
+        By inversion on (E-3),
+        $\eval[n_1]{e_1}{v_1}$
+        and~$\eval[n_2]{e_2}{v_2}$
+        and~$\eval[1]{\kwpair{v_1}{v_2}}{\kwvpair{v_1}{v_2}}$
+        and~$n = n_1 + n_2 + 1$.
+        By induction,
+        $\eval[n_1 + k]{e_1}{v_1}$
+        $|\store'| = k$.
+        By (E-3),
+        $\eval[n+k]{e}{v}$
+    \item (S-5) Then~$e = \kwpair{v_1}{e_2}$
+        and~$e' = \kwpair{v_1}{e_2'}$.
+        By inversion on (E-3),
+        $\eval[0]{v_1}{v_1}$
+        and~$\eval[n_2]{e_2}{v_2}$
+        and~$\eval[1]{\kwpair{v_1}{v_2}}{\kwvpair{v_1}{v_2}}$
+        and~$n = 0 + n_2 + 1$.
+        By induction,
+        $\eval[n_2 + k]{e_2}{v_2}$
+        $|\store'| = k$.
+        By (E-3),
+        $\eval[n+k]{e}{v}$
+    \item (S-6) Then~$e = \kwpair{v_1}{v_2}$
+        and~$e' = \kwpair{v_1}{v_2}$.
+        By inversion on (E-3),
+        $\eval[0]{v_1}{v_1}$
+        and~$\eval[0]{v_2}{v_2}$
+        and~$\eval[1]{\kwpair{v_1}{v_2}}{\kwvpair{v_1}{v_2}}$
+        and~$n = 0 + 0 + 1$.
+        By induction,
+        $\eval[n_2 + k]{e}{\kwvpair{v_1}{v_2}}$
+        $|\store'| = k = 1$.
+        By (E-3),
+        $\eval[n+k]{e}{v}$
+
+    \item (S-7) Then~$e = \kwfst{e_1}$ and~$e' = \kwfst{e_1'}$.
+        By inversion on (E-4),
+        $\eval[n_1]{e_1}{\kwvpair{v_1}{v_2}}$
+        and~$\eval[n_1]{\kwfst{e_1}}{v_1}$
+        and~$n = n_1$.
+        By induction,
+        $\eval[n_1 + k]{e}{v_1}$
+        $|\store'| = k$.
+        By (E-4),
+        $\eval[n+k]{e}{v}$
+
+    \item (S-8) Then~$e = \kwfst{\kwvpair{v_1}{v_2}}$ and~$e' = v_1$.
+        By inversion on (E-4),
+        $\eval[0]{\kwvpair{v_1}{v_2}}{\kwvpair{v_1}{v_2}}$
+        $\eval[0]{e'}{v_1}$, so $n = 0$.
+        By (S-8), $\store' = \estore$, so $k = 0$.
+        By (E-4),
+        $\eval[n+k]{e}{v}$
+
+    \item (S-9) Then~$e = \kwsnd{e_1}$ and~$e' = \kwsnd{e_1'}$.
+        By inversion on (E-4),
+        $\eval[n_1]{e_1}{\kwvpair{v_1}{v_2}}$
+        and~$\eval[n_1]{\kwfst{e_1}}{v_2}$
+        and~$n = n_1$.
+        By induction,
+        $\eval[n_1 + k]{e}{v_1}$
+        $|\store'| = k$.
+        By (E-5),
+        $\eval[n+k]{e}{v}$
+
+    \item (S-10) Then~$e = \kwfst{\kwvpair{v_1}{v_2}}$ and~$e' = v_1$.
+        By inversion on (E-5),
+        $\eval[0]{\kwvpair{v_1}{v_2}}{\kwvpair{v_1}{v_2}}$
+        $\eval[0]{e'}{v_2}$, so $n = 0$.
+        By (S-10), $\store' = \estore$, so $k = 0$.
+        By (E-5),
+        $\eval[n+k]{e}{v}$
+
     \end{itemize}

### 2021-11-12 project/proposal.tex
Draft

\documentclass{article}

% Language setting
% Replace `english' with e.g. `spanish' to change the document language
\usepackage[english]{babel}

% Set page size and margins
% Replace `letterpaper' with`a4paper' for UK/EU standard size
\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}

% Useful packages
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage[colorlinks=true, allcolors=blue]{hyperref}

\title{Verifying a Postfix Language \\
    \large CS595 Final Project Proposal}
\author{Dustin Van Tate Testa and Andrew Neth}

\begin{document}
\maketitle
\begin{abstract}
We intend to formalize and give some proofs of soundness to a subset of \href{https://github.com/dvtate/postfix-haskell}{a functional, postfix language}.
\end{abstract}

\section{The Language}
This section describes some of the key features of the original language. It is included only to show that the subset is a good representation of the original language and thus can be skipped.

\subsection{Syntax}
\begin{itemize}
    \item \textbf{Primitive Literals:} The language supports numeric and string literals although in the paper we will likely only cover numbers.

    \item \textbf{Lambdas:} Lambdas aka macros are enclosed within \texttt{(: ... )} and can have type annotations, for example \texttt{((I32 I32) (I32) rec: ... )}.

    \item \textbf{Tuples:} Tuples are enclosed within parenthesis. There is no reason to include separators like commas.

    \item \textbf{Identifiers:} there are two types of identifiers, escaped identifiers which start with \texttt{\$} are generally used for assignment and unescaped identifiers which are used for to invoke the value to which the identifier refers. To use an identifier \texttt{x} within a namespace \texttt{n} one can use a dot, ie - \texttt{\$n.x}

    \item \textbf{Functions:} function overloading is the only supported form of branching in the language. The syntax for the \texttt{fun} operator is as follows.

    \texttt{(($..._1$)($..._2$ I32): Condition ) (: Action ) \$Identifier fun}.

    In the final paper this will likely get replaced with a ternary.

    \item \textbf{Other Operators:} Although the majority of primitive operators are defined in the standard library, the compiler actually has a few builtins as well. Athough these are very important to actually using the language they aren't worthy of listing out in their entirety here. Some notable examples include: \texttt{require}, \texttt{use}, \texttt{import}, \texttt{export}, \texttt{asm}, \texttt{namespace}, \texttt{type}, \texttt{unpack}, \texttt{class}, \texttt{=}, \texttt{|}, \texttt{@}, \texttt{\~}, \texttt{==}, \texttt{make}

\end{itemize}

\subsection{Type System}
The language currently features an incomplete type system (with some compile-time-only values not fitting into the type system). However for the paper we will simplify it to exclude these values (notably namespaces, string literals, etc.) as they don't do anything interesting.

\begin{itemize}
    \item \textbf{Primitives:} Types supported by WebAssembly: \texttt{I32 I64 F32 F64}
    \item \textbf{Tuples:} Types which represent the concept of adjacent values on the stack. And thus actually represent multiple values when compiled.
    \begin{itemize}
        \item \textbf{Relevant operators:} \texttt{pack unpack}.
        \item \textbf{Example:} \texttt{(F32 F32 F32) \$FloatVec3 =}.
    \end{itemize}

    \item \textbf{Unions:} Created using the \texttt{|} operator, either of it's two operands can satisfy it. At present, all union types must be resolved at compile time, but this should be resolved once the garbage collector is finished.

    \item \textbf{Unit \& Void}: \texttt{Unit} is the type of the empty tuple and \texttt{Void} does not typecheck.

    \item \textbf{\texttt{Any}}: matches with any given type/value. Useful for pattern matching.

    \item \textbf{\texttt{Arrow} Types}: the \texttt{Arrow} operator is used for macro types. 
    Example: \texttt{(F32 F32) (F32) Arrow \$BinaryOperator =}

    \item \textbf{Classes:} A similar concept in other languages would likely be 'traits', adding a class to a type gives it functionality designated by that class. Thus we can define \texttt{I32 class \$Color =} and then use bitwise operators in order to make the \texttt{Color} class represent a 32bit packed RGBA value. In addition to there are a number of more advanced ways to use the class

    \begin{itemize}
        \item \textbf{Instantiating:} the \texttt{make} operator is used to add a class to a value. So for the RGBA class we can do \texttt{0xff00ff80 Color make \$half\_magenta =}

        \item \textbf{Subclasses:} We can make a class of our color class to add new functionality, for example \texttt{Color class \$RenderEngineColor =}

        \item \textbf{Parametric Classes:} In addition to taking types as arguments, the class operator can take a macro, this allows one to describe more complex types. One such usage is to define type operators.

        \texttt{(: \$T = (T T T) ) class \$Vec3 =}\\
        \texttt{(1 2 3) I32 Vec3 make \$pos =}

        Note that here \texttt{class} is technically optional and the following would be the same but less strict.

        \texttt{(: \$T = (T T T) ) \$Vec3 =}\\
        \texttt{(1 2 3) \$pos =}\\
        so that \texttt{pos type I32 Vec3 == :data} gives 1

        In the paper we will avoid ambiguity likely take the approach of adding 'type macros' with a brackets syntax instead.

        \item \textbf{Recursive Classes:} By marking the macro in the class as recursive it's only accessed as a reference to an object stored on the heap. This allows the programmer to define recursive types. Not currently supported.

        \texttt{(rec: \$T = (T List T) Unit | ) class \$List =}\\
        \texttt{( 1 ( 2 () I32 List make ) I32 List make ) I32 List make}
    \end{itemize}
\end{itemize}

\section{The Subset}
Here we describe at high level the subset of the language to formalize.

\section{The Formalization}
Here we describe the some ideas for rules, context, etc.

\end{document}

### 2021-11-16 project/proposal.tex
submitting this

@@
 \documentclass{article}
-
-% Language setting
-% Replace `english' with e.g. `spanish' to change the document language
 \usepackage[english]{babel}
@@
 % Set page size and margins
-% Replace `letterpaper' with`a4paper' for UK/EU standard size
 \usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
@@
 
-\section{The Language}
+\section{Motivation}
+WebAssembly is a typed, stack-machine based assembly language which is wildely supported across all major browsers.
+A number of new languages have come about with the goal of providing better support for the platform.
+Additionally it has proven the utility of typed stack-machines as a reasonably safe way to write programs.
+
+There appears to be little research into stack-based, functional languages (likely due to these languages not being common), thus more exploration on the topic could likely produce some value.
+
+The language this paper would be focused on is currently being used to power \href{https://ridder.xyz/game}{some insignificant demos}, but maybe in some distant future it will end up as the backend for a visual programming tool or some other application where guarantees of safety which could be produced here would be very important.
+
+\section{Preliminary Work - The Language}
 This section describes some of the key features of the original language. It is included only to show that the subset is a good representation of the original language and thus can be skipped.
@@
     \item \textbf{Identifiers:} there are two types of identifiers, escaped identifiers which start with \texttt{\$} are generally used for assignment and unescaped identifiers which are used for to invoke the value to which the identifier refers.
-To use an identifier \texttt{x} within a namespace \texttt{n} one can use a dot, ie - \texttt{\$n.x}
+
+    In the language, identifiers can reference both typed and untyped values, but likely will be limited to typed values in the paper.
 
@@
 Athough these are very important to actually using the language they aren't worthy of listing out in their entirety here.
-Some notable examples include: \texttt{require}, \texttt{use}, \texttt{import}, \texttt{export}, \texttt{asm}, \texttt{namespace}, \texttt{type}, \texttt{unpack}, \texttt{class}, \texttt{=}, \texttt{|}, \texttt{@}, \texttt{\~}, \texttt{==}, \texttt{make}
-
+Some examples include: \texttt{require}, \texttt{use}, \texttt{import}, \texttt{export}, \texttt{asm}, \texttt{namespace}, \texttt{type}, \texttt{unpack}, \texttt{class}, \texttt{=}, \texttt{|}, \texttt{@}, \texttt{\~}, \texttt{==}, \texttt{make}
 \end{itemize}
 
-\subsection{Type System}
+    Combining these concepts we can make the following program.
+
+    \begin{verbatim}
+    # Import basic math and logic
+    "stdlib/prelude.phs" require use
+
+    # Define a recursive factorial lambda
+    ((I32)(I32) rec:
+        $n =
+        (: 1 ) (: 1 ) $branch fun
+        (: n 1 > ) (: n 1 - fac n * ) $branch fun
+        branch
+    ) $fac =
+
+    # Export factorial to the host environment
+    (I32) (: fac ) "factorial" export
+    \end{verbatim}
+
+\subsection{Preliminary Work - Type System}
 The language currently features an incomplete type system (with some compile-time-only values not fitting into the type system).
@@
     \item \textbf{Unions:} Created using the \texttt{|} operator, either of it's two operands can satisfy it.
-At present, all union types must be resolved at compile time, but this should be resolved once the garbage collector is finished.
+At present, all union types must be resolved at compile time (ie - unions can exist as types but not values), but this should be resolved once the garbage collector is finished.
 
@@
 This allows the programmer to define recursive types.
-Not currently supported.
+Will be added once GC is stable.
 
@@
 
-\section{The Subset}
-Here we describe at high level the subset of the language to formalize.
+\section{Prelimnary work - The Subset}
+Here we describe at a high level the subset of the language to formalize.
 
-\section{The Formalization}
-Here we describe the some ideas for rules, context, etc.
+\subsection{Branching}
+As mentioned before a ternary operator would be a reasonable subset of branching via function overloading. The tern operator is defined as an operator which
+
+\begin{verbatim}
+# Tern takes an I32 and two values of the same type
+# and gives one of those two values
+$tern ~ :type # :type - (I32 $A $A) ($A) Arrow
+
+# As it is a subset we can define it in terms of functions
+((I32 Any Any): type swap type == ) (: ( $c $a $b ) = b ) $tern fun
+((I32 Any Any): type swap type == && ) (: ( $c $a $b ) = a ) $tern fun
+\end{verbatim}
+
+\subsection{Lambdas}
+Via the same syntax already defined
+
+\subsection{Recursion}
+Need to think about how best to do this in a formalization but there was a lot of time that had to go into making type inference work for recursive functions (implementation + 3 pages of notes).
+
+\subsection{Tuples}
+N-ary tuples represent one of the more problematic parts of the language, notably \texttt{unpack} being the only operator with a type signature dependant on the type of it's input, additionally I have no clue how to define an arrow type with a specified but non-constant number of return values.
+In order to avoid expanding the scope of this assignment we may replace the \texttt{unpack} operator with a set of two operators with similar functionality.
+
+\subsection{Unions}
+Unions should be an easy addition.
+
+\begin{verbatim}
+    (()(I32 F32 |): 12 ) $value =
+    value type I32 == 1 0 tern $value_is_int =
+\end{verbatim}
+
+\subsection{Identifiers}
+Although the system for identifiers is a bit different from most languages we've covered, it's not too strange.
+
+\subsection{Types}
+Although types, which are treated as untyped stack values, aren't pretty we could take the approach of having a second stack (for operations on types) and second store (for type variables), instead of the implementation's approach.
+Or we can have a type judgement in addition to a value judgement.
+
+\subsection{Operators, Keywords and Constants}
+So in summary our builtin operators would be: 
+\begin{itemize}
+    \item \texttt{+} - numeric add
+    \item \texttt{*} - numeric mul
+    \item \texttt{if-then-else} - ternary operator
+    \item \texttt{unpack} - destruct tuple (an attempt will be made)
+    \item \texttt{\~} - Extract value from an escaped identifier without invoking it (less important for safety but useful for defining additional operators using these as base)
+    \item \texttt{=} - the expression \texttt{e \$x = ...} is equivalent to \texttt{let x = e in ...}. But note that there are no 'bound variables' in this language.
+    \item \texttt{type} - get type of argument (equivalent to \texttt{type()} in python
+    \item \texttt{==} - current plan is to have an ambiguous == operator which accepts either 2 types or 2 equivalently typed values like in the language
+    \item \texttt{|} - used to form type unions
+    \item \texttt{Arrow} - lambda types
+    \item \texttt{Unit} - keyword to distinguish from the empty tuple ()
+    \item \texttt{Void} - used due to lack of syntax for empty union
+\end{itemize}
+
+\subsection{Notably Not Included}
+
+\subsubsection{Classes}
+It would be nice to include classes in our language (notably recursive classes). However this would involve subtyping and some other concepts so might not have time to add them to the model
+
+As much as we would like to include classes (notably recursive classes), it would likely add excess complexity, requiring subtyping (on N dimensions); untyped lambdas which can operate on types; and likely some other painpoints.
+And it's not clear how they could make the system unsafe as apart from being used to define recursive types are simply specialized versions of preexisting types.
+
+\subsubsection{Functions}
+Really the only reason I opted for the function overloading system instead of normal branching in the original language was in order to make it easier to extend the language.
+Function overloading seems like it would be needlessly painful to formalize and doesn't contribute to our proof of soundness any more than branching via a ternary.
+
+\section{Proposed Work}
+\subsection{The Formalization}
+Using big step semantics most of these aren't too scary. Just need to sit down and combine ideas.
+
+\subsection{Proof of Soundness}
+Proving theorems of Progress and Preservation should sufficiently prove a baseline of safety.
+
+\subsection{Findings, Resolution?, Conclusions}
+It's entirely possible that we could encounter problems with the language design, although pretty unlikely since we're even requiring type signatures on lambdas.
+
+\section{Plan Distribution of Work}
+"Mostly" is used to indicate any >50\% but likely not independent
+\subsection{Timeline}
+\begin{itemize}
+    \item Week of Nov 15: Further research and discussion over formalization, rules written down, mostly Tate
+    \item Week of Nov 22: Tate and Andrew will each independently work on the soundness proofs (Tate Progress, Andrew preservation, although subject to change)
+    \item Week of Nov 29: 
+        \begin{itemize}
+            \item Peer-review proofs - equal work
+            \item Combine proofs into paper - mostly Andrew
+            \item Start on presentation - mostly Andrew
+        \end{itemize}
+    \item Week of Dec 6: 
+        \begin{itemize} 
+            \item Attempt to resolve any findings - mostly Tate
+            \item finalize paper presentation - equal work
+        \end{itemize}
+\end{itemize}
+
+\subsection{Things to Research}
+\begin{itemize}
+    \item Formalization for Forth or PostScript, stack-based languages which don't follow the variable binding
+\end{itemize}
 

### 2021-11-17 project/proposal.tex
Update proposal.tex

@@
 \documentclass{article}
+
+% Language setting
+% Replace `english' with e.g. `spanish' to change the document language
 \usepackage[english]{babel}
@@
 % Set page size and margins
+% Replace `letterpaper' with`a4paper' for UK/EU standard size
 \usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
@@
 \usepackage[colorlinks=true, allcolors=blue]{hyperref}
+
+\input{macros}
 
@@
 
-    \item \textbf{Lambdas:} Lambdas aka macros are enclosed within \texttt{(: ... )} and can have type annotations, for example \texttt{((I32 I32) (I32) rec: ... )}.
+    \item \textbf{closures:} closures aka macros are enclosed within \texttt{(: ... )} and can have type annotations, for example \texttt{((I32 I32) (I32) rec: ... )}.
 
@@
 
-    # Define a recursive factorial lambda
+    # Define a recursive factorial closure
     ((I32)(I32) rec:
@@
 
-\subsection{Lambdas}
+\subsection{closures}
 Via the same syntax already defined
@@
     \item \texttt{|} - used to form type unions
-    \item \texttt{Arrow} - lambda types
+    \item \texttt{Arrow} - closure types
     \item \texttt{Unit} - keyword to distinguish from the empty tuple ()
@@
 \subsubsection{Classes}
-It would be nice to include classes in our language (notably recursive classes). However this would involve subtyping and some other concepts so might not have time to add them to the model
-
-As much as we would like to include classes (notably recursive classes), it would likely add excess complexity, requiring subtyping (on N dimensions); untyped lambdas which can operate on types; and likely some other painpoints.
+As much as we would like to include classes (notably recursive classes), it would likely add excess complexity, requiring subtyping (on N dimensions); untyped closures which can operate on types; and likely some other painpoints.
 And it's not clear how they could make the system unsafe as apart from being used to define recursive types are simply specialized versions of preexisting types.
@@
 Function overloading seems like it would be needlessly painful to formalize and doesn't contribute to our proof of soundness any more than branching via a ternary.
+As the case for functions having potentially no possible branch results in a type error in the complete language and this is simplified by having an else clause.
+
+\section{Syntax}
+Taking the ideas from the above rant into account we arrive at something along these lines
+\[
+\begin{array}{r l}
+    \tau \bnfdef & \kwInt \bnfalt \kwLong \bnfalt \kwFloat \bnfalt \kwDouble \bnfalt 
+        (\tau^*) \bnfalt \kwUnit \bnfalt \kwVoid \bnfalt \tau_1~ \tau_2~ + 
+        \bnfalt ~ \kwArrow{\tau_1} {\tau_2} \bnfalt \kwType{e}\\
+
+    \tau^* \bnfdef & \varepsilon \bnfalt \tau^* ~ \tau \bnfalt \kwunpack{\tau}\\
+
+    e \bnfdef & e~e~+ \bnfalt ~e~e~\times \bnfalt ~ \kwtern{e}{e}{e} 
+        \bnfalt e~e~== \bnfalt ~ \tau~ \tau~ ==
+        \bnfalt ~ ( e^* )
+        \bnfalt \kwmacro{\tau^*}{S} \bnfalt \kwfix{e}{\tau}
+        \bnfalt \kwid{x} \bnfalt \kwunesc{x}\\
+
+    e^* \bnfdef & \varepsilon \bnfalt e^* ~ e \bnfalt \kwunpack{e}\\
+
+    S \bnfdef & S~ S \bnfalt e~ \kweid{x} = \bnfalt ~e
+
+\end{array}
+\]
+
+Notes
+\begin{itemize}
+    \item A syntactic fixed point combinator was chosen instead of $\kwrec{\tau^*}{\tau^*}{e^*}$
+    \item notice that unpack can have any non-negative number of output values depending on the given input
+    \item $\kwunpack{e}$ and $\kwunpack{\tau}$ can give 0 or more outputs depending on the input tuple
+\end{itemize}
 
@@
 \subsection{Findings, Resolution?, Conclusions}
-It's entirely possible that we could encounter problems with the language design, although pretty unlikely since we're even requiring type signatures on lambdas.
+It's entirely possible that we could encounter problems with the language design, although pretty unlikely since we're even requiring type signatures on closures.
 

### 2021-11-25 project/paper.tex
start on final paper

--- 2021-11-17 project/proposal.tex
@@
 % Useful packages
-\usepackage{amsmath}
+\usepackage{amsmath, amsthm, verbatim, enumerate, xcolor, mathpartir}
 \usepackage{graphicx}
@@
 \begin{abstract}
-We intend to formalize and give some proofs of soundness to a subset of \href{https://github.com/dvtate/postfix-haskell}{a functional, postfix language}.
+A formalization and some proofs of soundness is given to a subset of \href{https://github.com/dvtate/postfix-haskell}{a functional, postfix language}.
 \end{abstract}
@@
     \tau \bnfdef & \kwInt \bnfalt \kwLong \bnfalt \kwFloat \bnfalt \kwDouble \bnfalt 
-        (\tau^*) \bnfalt \kwUnit \bnfalt \kwVoid \bnfalt \tau_1~ \tau_2~ + 
+        (\tau^*) \bnfalt \kwUnit \bnfalt \kwVoid \bnfalt \tau_1~ \tau_2~ \texttt{|} 
         \bnfalt ~ \kwArrow{\tau_1} {\tau_2} \bnfalt \kwType{e}\\
@@
 
-    S \bnfdef & S~ S \bnfalt e~ \kweid{x} = \bnfalt ~e
+    S \bnfdef & \varepsilon \bnfalt S~ S \bnfalt e~ \kweid{x} = \bnfalt ~e
 
@@
 
-\section{Proposed Work}
-\subsection{The Formalization}
-Using big step semantics most of these aren't too scary. Just need to sit down and combine ideas.
+\section{Semantics}
+We've expanded the macro value syntax to also capture lexically scoped variables. These are represented as with a brackets syntax as shown: $(( \tau^* )[ x \step (), y \step 123 ]: ... )$.
 
-\subsection{Proof of Soundness}
-Proving theorems of Progress and Preservation should sufficiently prove a baseline of safety.
+{
+    \centering
+    \def \MathparLineskip {\lineskip=0.43cm}
+    \begin{mathpar}
+        \Rule{Num Lits}
+             {\forall ~ T \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
+             {\kwnumlit{n}{T} \val}
+        \and
 
-\subsection{Findings, Resolution?, Conclusions}
-It's entirely possible that we could encounter problems with the language design, although pretty unlikely since we're even requiring type signatures on closures.
+        % Derivation of e* -> e*'
+        \Rule{$e^* \step~ e^{*\prime}$ - Part 1}
+            {e_1 \step e_1'}
+            {e_1 ~ ... ~ e_n \step e_1' ~ ... ~ e_n}
+        \and
+        \Rule{$e^* \step~ e^{*\prime}$ - Part 2}
+            {e_1 \step e_1'
+                \\v_i \val ~ \forall v_i \in v*}
+            {v* ~ e_1 ~ ... ~ e_n \step v* ~ e_1' ~ ... ~ e_n}
+        \and
 
-\section{Plan Distribution of Work}
-"Mostly" is used to indicate any >50\% but likely not independent
-\subsection{Timeline}
-\begin{itemize}
-    \item Week of Nov 15: Further research and discussion over formalization, rules written down, mostly Tate
-    \item Week of Nov 22: Tate and Andrew will each independently work on the soundness proofs (Tate Progress, Andrew preservation, although subject to change)
-    \item Week of Nov 29: 
-        \begin{itemize}
-            \item Peer-review proofs - equal work
-            \item Combine proofs into paper - mostly Andrew
-            \item Start on presentation - mostly Andrew
-        \end{itemize}
-    \item Week of Dec 6: 
-        \begin{itemize} 
-            \item Attempt to resolve any findings - mostly Tate
-            \item finalize paper presentation - equal work
-        \end{itemize}
-\end{itemize}
+        % Tuple literals evaluation
+        \Rule{Tuple Lits - 1}
+            {e^* \step e^{*\prime}}
+            {( e^* ) \step ( e^{*\prime} )}
+        \Rule{Tuple Lits - 2}
+            {v_i \val ~ \forall v_i \in v*}
+            {(v*) \val}
+        \and
 
-\subsection{Things to Research}
-\begin{itemize}
-    \item Formalization for Forth or PostScript, stack-based languages which don't follow the variable binding
-\end{itemize}
+        % unpack operator
+        \Rule{Tuple Value Unpack - 1}
+            {v_i \val ~ \forall i \in [1, n]}
+            {\kwunpack{(v_1 ... v_n)} \step v_1 ... v_n}
+        \Rule{Tuple Value Unpack - 2}
+            {\strut}
+            {\kwunpack{()} \step \varepsilon}
+        \and
+        \Rule{Tuple Type Unpack - 1}
+            {\strut}
+            {\kwunpack{(\tau_1 ... \tau_n)} \step \tau_1 ... \tau_n}
+        \Rule{Tuple Type Unpack - 2}
+            {\strut}
+            {\kwunpack{\kwUnit} \step \varepsilon}
+
+        % math
+        \Rule{Numeric Sum}
+            {v_1 \val \\ v_2 \val \\ v_3 = v_1 + v_2}
+            {v_1~ v_2~ + \step v_3}
+        \Rule{Numeric Product}
+            {v_1 \val \\ v_2 \val \\ v_3 = v_1 * v_2}
+            {v_1~ v_2~ \times \step v_3}
+        \and
+
+        % ternary
+        \Rule{Ternary 1}
+            {v_1 \val \\ v_2 \val}
+            {\kwtern{\kwnumlit{0}{\kwInt}}{v_1}{v_2} \step v_2}
+        \Rule{Ternary 2}
+            {v_1 \val \\ v_2 \val \\ n \neq 0}
+            {\kwtern{\kwnumlit{n}{\kwInt}}{v_1}{v_2} \step v_1}
+
+    \end{mathpar}
+}
 

### 2021-11-25 project/proposal.tex
move some work to the paper

@@
 
-\section{Syntax}
-Taking the ideas from the above rant into account we arrive at something along these lines
-\[
-\begin{array}{r l}
-    \tau \bnfdef & \kwInt \bnfalt \kwLong \bnfalt \kwFloat \bnfalt \kwDouble \bnfalt 
-        (\tau^*) \bnfalt \kwUnit \bnfalt \kwVoid \bnfalt \tau_1~ \tau_2~ + 
-        \bnfalt ~ \kwArrow{\tau_1} {\tau_2} \bnfalt \kwType{e}\\
-
-    \tau^* \bnfdef & \varepsilon \bnfalt \tau^* ~ \tau \bnfalt \kwunpack{\tau}\\
-
-    e \bnfdef & e~e~+ \bnfalt ~e~e~\times \bnfalt ~ \kwtern{e}{e}{e} 
-        \bnfalt e~e~== \bnfalt ~ \tau~ \tau~ ==
-        \bnfalt ~ ( e^* )
-        \bnfalt \kwmacro{\tau^*}{S} \bnfalt \kwfix{e}{\tau}
-        \bnfalt \kwid{x} \bnfalt \kwunesc{x}\\
-
-    e^* \bnfdef & \varepsilon \bnfalt e^* ~ e \bnfalt \kwunpack{e}\\
-
-    S \bnfdef & S~ S \bnfalt e~ \kweid{x} = \bnfalt ~e
-
-\end{array}
-\]
-
-Notes
-\begin{itemize}
-    \item A syntactic fixed point combinator was chosen instead of $\kwrec{\tau^*}{\tau^*}{e^*}$
-    \item notice that unpack can have any non-negative number of output values depending on the given input
-    \item $\kwunpack{e}$ and $\kwunpack{\tau}$ can give 0 or more outputs depending on the input tuple
-\end{itemize}
-
 \section{Proposed Work}

### 2021-11-26 project/paper.tex
organize, start on typing section

@@
 
-\section{Preliminary Work - The Language}
+\section{The Language}
 This section describes some of the key features of the original language. It is included only to show that the subset is a good representation of the original language and thus can be skipped.
@@
 
-\section{Prelimnary work - The Subset}
+\section{The Subset}
 Here we describe at a high level the subset of the language to formalize.
@@
         \bnfalt \kwmacro{\tau^*}{S} \bnfalt \kwfix{e}{\tau}
-        \bnfalt \kwid{x} \bnfalt \kwunesc{x}\\
+        \bnfalt \kwid{x} \bnfalt e ~ \textbf{@} 
+        \bnfalt S~ S \bnfalt e~ \kweid{x} = \\
 
     e^* \bnfdef & \varepsilon \bnfalt e^* ~ e \bnfalt \kwunpack{e}\\
-
-    S \bnfdef & \varepsilon \bnfalt S~ S \bnfalt e~ \kweid{x} = \bnfalt ~e
 
@@
 \begin{itemize}
-    \item A syntactic fixed point combinator was chosen instead of $\kwrec{\tau^*}{\tau^*}{e^*}$
+    \item A syntactic fixed point combinator was chosen instead of $\kwrec{\tau^*}{\tau^*}{e^*}$. It's possible that one could be implemented in the language itself.
     \item notice that unpack can have any non-negative number of output values depending on the given input
@@
 \section{Semantics}
-We've expanded the macro value syntax to also capture lexically scoped variables. These are represented as with a brackets syntax as shown: $(( \tau^* )[ x \step (), y \step 123 ]: ... )$.
 
+% TODO we should split this into subsections
+\subsection{Value Judgements}
+Types come in the form of values and expressions just like other language features.
 {
@@
     \begin{mathpar}
-        \Rule{Num Lits}
-             {\forall ~ T \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
-             {\kwnumlit{n}{T} \val}
+        % number literals
+        \Rule{V-1}
+             {\forall ~ \tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
+             {\kwnumlit{n}{\tau} \val}
         \and
 
+        % type values
+        \Rule{V-2}{\strut}{\kwInt \val} \and
+        \Rule{V-3}{\strut}{\kwLong \val} \and
+        \Rule{V-4}{\strut}{\kwFloat \val} \and
+        \Rule{V-5}{\strut}{\kwDouble \val} \and
+        \Rule{V-6}{\tau_1 \val \\ \tau_2 \val}{\tau_1 \tau_2 \texttt{|} \val} \and
+        \Rule{V-7}{\tau_1 \val \\ \tau_2 \val}{\kwArrow{\tau_1}{\tau_2} \val} \and
+        % fixed point combinator
+        \Rule{Fixed Point}{e \val \\ \tau \val}{\kwfix{e}{\tau} \val}
+    \end{mathpar}  
+}
+
+\subsection{Step Judgement}
+Expressions are evaluated from left to right until everything is a value
+{
+    \centering
+    \def \MathparLineskip {\lineskip=0.43cm}
+    \begin{mathpar}
         % Derivation of e* -> e*'
-        \Rule{$e^* \step~ e^{*\prime}$ - Part 1}
+        \Rule{STEP-1}
             {e_1 \step e_1'}
@@
         \and
-        \Rule{$e^* \step~ e^{*\prime}$ - Part 2}
+        \Rule{STEP-2}
             {e_1 \step e_1'
@@
         \and
+    \end{mathpar}  
+}
+
+\subsection{Macros, Identifiers and Scoping}
+The store, $\sigma$, consists of a list of lists of mappings from identifiers to values.
+Notice the addition of an auxiliary $\kw{end\_scope}$ operator which removes the current scope.
+A macro literal steps to a macro value by first stepping it's input types until they're all values before substituting all of the values in the store within it's body (forming a closure).
+Calling a macro value pushes a new empty mappings list onto the store, places the macro's expressions and a $\kw{end\_scope}$ operator onto the stack, enabling local identifiers.
+{
+    \centering
+    \def \MathparLineskip {\lineskip=0.43cm}
+    \begin{mathpar}    
+        % Closures capture lexically scoped variables via substitution 
+        \Rule{LET-1}
+            {v \val
+                \\ \sigma = [ [...], ... [ x_1 \step v_1 ... x_n \step v_n] ]
+                \\ \sigma' = [ [...], ... [ x_1 \step v_1 ... x_n \step v_n, x_{n+1} \step v_{n+1}] ]}
+            {\sigma; v ~ \kweid{x} = \step \sigma'; \varepsilon}
+
+        \Rule{MACRO-1}
+            {\tau^* \step \tau^{\prime*}}
+            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^{\prime*}}{e^*}}
+
+        \Rule{MACRO-2}
+            {\uplus^{|\sigma|}_{i = 1} \sigma(i) = [x_1 \step v_1 ~ ... ~ x_n \step v_n]
+                \\ \tau_i \val~ \forall~ \tau_i \in \tau*}
+            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^*}{[v_1/x_1]...[v_n/x_n] e^*} \val}
+
+        \Rule{CALL-1}
+            {\sigma' = \sigma \cup [~] }
+            {\sigma; \kwmacro{\tau^*}{e^*} \textbf{@} \step \sigma'; e^* \kw{end\_scope}}
+
+        \Rule{CALL-2}
+            {\sigma' = \text{$\sigma$ but with the last list of mappings removed}}
+            {\sigma; \kw{end\_scope} \step \sigma'; \varepsilon} 
+    \end{mathpar}  
+}
+
+\subsection{Others}
+{
+    \centering
+    \def \MathparLineskip {\lineskip=0.43cm}
+    \begin{mathpar}
 
         % Tuple literals evaluation
-        \Rule{Tuple Lits - 1}
+        \Rule{TUPLES-1}
             {e^* \step e^{*\prime}}
             {( e^* ) \step ( e^{*\prime} )}
-        \Rule{Tuple Lits - 2}
+        \Rule{TUPLES-2}
             {v_i \val ~ \forall v_i \in v*}
@@
         % unpack operator
-        \Rule{Tuple Value Unpack - 1}
-            {v_i \val ~ \forall i \in [1, n]}
+        \Rule{TUPLES-3}
+            {v_i \val ~ \forall~ i \in [1, n]}
             {\kwunpack{(v_1 ... v_n)} \step v_1 ... v_n}
-        \Rule{Tuple Value Unpack - 2}
+        \Rule{TUPLES-4}
             {\strut}
@@
         \and
-        \Rule{Tuple Type Unpack - 1}
-            {\strut}
+        \Rule{TUPLES-5}
+            {\tau_i \val ~ \forall~ i \in [1, n]}
             {\kwunpack{(\tau_1 ... \tau_n)} \step \tau_1 ... \tau_n}
-        \Rule{Tuple Type Unpack - 2}
+        \Rule{TUPLES-6}
             {\strut}
             {\kwunpack{\kwUnit} \step \varepsilon}
+        \and
 
         % math
-        \Rule{Numeric Sum}
+        \Rule{ADD}
             {v_1 \val \\ v_2 \val \\ v_3 = v_1 + v_2}
             {v_1~ v_2~ + \step v_3}
-        \Rule{Numeric Product}
+        \Rule{MUL}
             {v_1 \val \\ v_2 \val \\ v_3 = v_1 * v_2}
@@
         % ternary
-        \Rule{Ternary 1}
+        \Rule{TERN-1}
             {v_1 \val \\ v_2 \val}
             {\kwtern{\kwnumlit{0}{\kwInt}}{v_1}{v_2} \step v_2}
-        \Rule{Ternary 2}
+        \Rule{TERN-2}
             {v_1 \val \\ v_2 \val \\ n \neq 0}
             {\kwtern{\kwnumlit{n}{\kwInt}}{v_1}{v_2} \step v_1}
+        \and
 
+        % typeof operator
+        \Rule{TYPE-1}{v \val \\ \etyped{v}{\tau}}{\kwType{v} \step \tau} \and
+        % Move this to type rules
+        % \Rule{Type Operator}
+        %     {\tau^* = \tau_1 ~...~ \tau_n \\
+        %      v^* = \etyped{v_1}{\tau_1} ~...~ \etyped{v_1}{\tau_1} \\
+        %      m = \kwmacro{\tau^*}{e^*} \\
+        %      m \val }
+        %     {\kwType{m} \step \kwArrow{(\tau^*)}{\kwType{(v^*~ m~ \textbf{@})}}}
     \end{mathpar}
@@
 
+\section{Typing}
+
+\subsection{Math and Logic}
+{
+    \centering
+    \def \MathparLineskip {\lineskip=0.43cm}
+    \begin{mathpar}
+        \Rule{T-1}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwInt}}{\kwInt}}
+        \and \Rule{T-2}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwLong}}{\kwLong}}
+        \and \Rule{T-3}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwFloat}}{\kwFloat}}
+        \and \Rule{T-4}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwDouble}}{\kwDouble}}
+        \and \Rule{T-5}
+            {\typed{\ctx}{e_1}{\tau} 
+                \\ \typed{\ctx}{e_2}{\tau} 
+                \\ \tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}}
+            {\typed{\ctx}{e_1~ e_2~ +}{\tau}}
+        \and \Rule{T-6}
+            {\typed{\ctx}{e_1}{\tau} 
+                \\ \typed{\ctx}{e_2}{\tau} 
+                \\ \tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}}
+            {\typed{\ctx}{e_1~ e_2~ \times}{\tau}}
+        \and \Rule{T-7}
+            {\typed{\ctx}{e_1}{\kwInt}
+                \\ \typed{\ctx}{e_2}{\tau}
+                \\ \typed{\ctx}{e_3}{\tau}}
+            {\typed{\ctx}{\kwtern{e_1}{e_2}{e_3}}{\tau}}
+    \end{mathpar}
+}
 \end{document}

### 2021-11-27 project/paper.tex
Formatting, start type rules

@@
 
+% ???
+\usepackage{syntax}
+
 \input{macros}
@@
 \title{Verifying a Postfix Language \\
-    \large CS595 Final Project Proposal}
+    \large CS595 Final Project}
 \author{Dustin Van Tate Testa and Andrew Neth}
@@
 \section{Motivation}
-WebAssembly is a typed, stack-machine based assembly language which is wildely supported across all major browsers.
+WebAssembly is a typed, stack-machine based assembly language which is widely supported across all major browsers.
 A number of new languages have come about with the goal of providing better support for the platform.
@@
 
-    \item \textbf{closures:} closures aka macros are enclosed within \texttt{(: ... )} and can have type annotations, for example \texttt{((I32 I32) (I32) rec: ... )}.
+    \item \textbf{Closures:} closures aka macros are enclosed within \texttt{(: ... )} and can have type annotations, for example \texttt{((I32 I32) (I32) rec: ... )}.
 
@@
     \item \textbf{Other Operators:} Although the majority of primitive operators are defined in the standard library, the compiler actually has a few builtins as well.
-Athough these are very important to actually using the language they aren't worthy of listing out in their entirety here.
+Although these are very important to actually using the language they aren't worthy of listing out in their entirety here.
 Some examples include: \texttt{require}, \texttt{use}, \texttt{import}, \texttt{export}, \texttt{asm}, \texttt{namespace}, \texttt{type}, \texttt{unpack}, \texttt{class}, \texttt{=}, \texttt{|}, \texttt{@}, \texttt{\~}, \texttt{==}, \texttt{make}
@@
 \section{Syntax}
-Taking the ideas from the above rant into account we arrive at something along these lines
-\[
-\begin{array}{r l}
-    \tau \bnfdef & \kwInt \bnfalt \kwLong \bnfalt \kwFloat \bnfalt \kwDouble \bnfalt 
-        (\tau^*) \bnfalt \kwUnit \bnfalt \kwVoid \bnfalt \tau_1~ \tau_2~ \texttt{|} 
-        \bnfalt ~ \kwArrow{\tau_1} {\tau_2} \bnfalt \kwType{e}\\
+Taking the ideas from the above rant into account we arrive at something along these lines:
 
-    \tau^* \bnfdef & \varepsilon \bnfalt \tau^* ~ \tau \bnfalt \kwunpack{\tau}\\
+\renewcommand{\grammarlabel}[2]{#1 \hfill#2}
+\begin{grammar}
+    <$\tau$> ::= $\kwInt$
+    |    $\kwLong$
+    |    $\kwFloat$
+    |    $\kwDouble$
+    \alt $(\tau^*)$
+    |    $\kwUnit$
+    |    $\kwVoid$
+    |    $\tau_1~ \tau_2~ \texttt{|}$
+    |    $\kwArrow{\tau_1} {\tau_2}$
+    |    $\kwType{e}$
 
-    e \bnfdef & e~e~+ \bnfalt ~e~e~\times \bnfalt ~ \kwtern{e}{e}{e} 
-        \bnfalt e~e~== \bnfalt ~ \tau~ \tau~ ==
-        \bnfalt ~ ( e^* )
-        \bnfalt \kwmacro{\tau^*}{S} \bnfalt \kwfix{e}{\tau}
-        \bnfalt \kwid{x} \bnfalt e ~ \textbf{@} 
-        \bnfalt S~ S \bnfalt e~ \kweid{x} = \\
+    <$\tau^*$> ::= $\varepsilon$ | $\tau^* ~ \tau$ | $\kwunpack{\tau}$
 
-    e^* \bnfdef & \varepsilon \bnfalt e^* ~ e \bnfalt \kwunpack{e}\\
+    <$e$> ::= $e~ e~ +$
+    |    $e~ e~ \times$
+    |    $\kwtern{e}{e}{e} $
+    |    $e~ e~ ==$
+    |    $\tau~ \tau~ ==$
+    \alt $(e^*)$
+    |    $\kwmacro{\tau^*}{S}$
+    |    $\kwfix{e}{\tau}$
+    \alt $\kwid{x}$
+    |    $e~ \textbf{@} $
+    |    $S~ S$
+    |    $e~ \kweid{x} =$
 
-\end{array}
-\]
+    <$e^*$> ::= $\varepsilon$ | $e^* ~ e$ | $\kwunpack{e}$
+\end{grammar}
 
@@
 \subsection{Macros, Identifiers and Scoping}
-The store, $\sigma$, consists of a list of lists of mappings from identifiers to values.
+The store, $\sigma$, consists of a list of \textit{scopes}, where each scope is a mapping from identifiers to values.
 Notice the addition of an auxiliary $\kw{end\_scope}$ operator which removes the current scope.
-A macro literal steps to a macro value by first stepping it's input types until they're all values before substituting all of the values in the store within it's body (forming a closure).
-Calling a macro value pushes a new empty mappings list onto the store, places the macro's expressions and a $\kw{end\_scope}$ operator onto the stack, enabling local identifiers.
+A macro literal steps to a macro value by first stepping its input types until they're all values before substituting all of the values in the store within its body (forming a closure).
+Calling a macro value pushes a new empty mappings list onto the store, then places the macro's expressions and an $\kw{end\_scope}$ operator onto the stack, enabling local identifiers.
 {
@@
                 \\ \sigma = [ [...], ... [ x_1 \step v_1 ... x_n \step v_n] ]
-                \\ \sigma' = [ [...], ... [ x_1 \step v_1 ... x_n \step v_n, x_{n+1} \step v_{n+1}] ]}
+                \\ \sigma' = [ [...], ... [ x_1 \step v_1 ... x_n \step v_n, x_{n+1} \step v] ]}
             {\sigma; v ~ \kweid{x} = \step \sigma'; \varepsilon}
 
-        \Rule{MACRO-1}
+        \and \Rule{MACRO-1}
             {\tau^* \step \tau^{\prime*}}
@@
 
-        \Rule{MACRO-2}
+        \and \Rule{MACRO-2}
             {\uplus^{|\sigma|}_{i = 1} \sigma(i) = [x_1 \step v_1 ~ ... ~ x_n \step v_n]
@@
 
-        \Rule{CALL-1}
+        \and \Rule{CALL-1}
             {\sigma' = \sigma \cup [~] }
@@
 
-        \Rule{CALL-2}
-            {\sigma' = \text{$\sigma$ but with the last list of mappings removed}}
+        \and \Rule{CALL-2}
+            {\sigma = [s_1,...s_n,s_{n+1}] \\ \sigma' = [s_1,...s_n]}
             {\sigma; \kw{end\_scope} \step \sigma'; \varepsilon}
+
+        \and \Rule{Identifier}
+            {[ ... ~ \kwid{x} \step v ~ ... ] \in \sigma}
+            {\sigma; \kwid{x} \step \sigma; v}
     \end{mathpar}  
@@
 
+\subsection{Step Judgement}
+{
+    \centering
+    \begin{mathpar}
+    \Rule{T-0}
+        {\tau \step \tau' \\ \typed{\ctx}{e}{\tau}}
+        {\typed{\ctx}{e}{\tau'}}
+    \end{mathpar}
+}
+
 \subsection{Math and Logic}
@@
 }
+
+\subsection{Tuples}
+{
+    \centering
+    \def \MathparLineskip {\lineskip=0.43cm}
+    \begin{mathpar}
+        \Rule{Empty Tuple}{\strut}{\typed{\ctx}{()}{\kwUnit}}
+        \and \Rule{Tuple Type}
+            {\tau^* = \{ \tau_i ~\text{such that}~ \typed{\ctx}{e_i}{\tau_i}~ \forall ~ e_i \in e^* \} }
+            {\typed{\ctx}{(e^*)}{(\tau^*)}}
+        \and \Rule{Unpack}
+            {\typed{\ctx}{(e^*)}{(\tau^*)}}
+            {\typed{\ctx}{\kwunpack{(e^*)}}{\tau^*}}
+        % equivalent types?
+        % \and \Rule{Unpack types}
+        %     {\typed{\ctx}{e*}{\kwunpack{(\tau^*)}}}
+        %     {\typed{\ctx}{e*}{\tau*}}
+    \end{mathpar}
+}
+
+\subsection{Macros}
+{
+    \centering
+    \def \MathparLineskip {\lineskip=0.43cm}
+    \begin{mathpar}
+
+    \end{mathpar}
+}
 \end{document}

### 2021-11-29 project/paper.tex
Refactor rules w/ type semantics

@@
     |    $\kwUnit$
-    |    $\kwVoid$
     |    $\tau_1~ \tau_2~ \texttt{|}$
@@
     \alt $(e^*)$
-    |    $\kwmacro{\tau^*}{S}$
+    |    $\kwmacro{\tau^*}{e^*}$
     |    $\kwfix{e}{\tau}$
     \alt $\kwid{x}$
+    |    $\kwid{T}$
     |    $e~ \textbf{@} $
-    |    $S~ S$
     |    $e~ \kweid{x} =$
@@
 \section{Semantics}
+\subsection{Notation}
+To annotate stack values we used the notation of two values being written next to each other, as they would be written in the language itself.
+The values at the bottom of the stack (furthest to the left) are often irrelevant, and thus can be omitted with this notation.
+The superscript * notation (ie - $e^*$) is used to annotate 0 or more of the preceding.
+This is explicitly defined in the syntax but implicitly expanded in the semantics.
 
-% TODO we should split this into subsections
 \subsection{Value Judgements}
-Types come in the form of values and expressions just like other language features.
+{
+    \centering
+    \def \MathparLineskip {\lineskip=0.43cm}
+    \begin{mathpar}
+
+        % fixed point combinator
+        \Rule{Fixed Point}{e \val \\ \tau \prim}{\kwfix{e}{\tau} \val}
+    \end{mathpar}
+}
+
+\subsection{Value and Step Judgement}
+Expressions are evaluated from left to right until everything is a value
 {
@@
 
-        % type values
-        \Rule{V-2}{\strut}{\kwInt \val} \and
-        \Rule{V-3}{\strut}{\kwLong \val} \and
-        \Rule{V-4}{\strut}{\kwFloat \val} \and
-        \Rule{V-5}{\strut}{\kwDouble \val} \and
-        \Rule{V-6}{\tau_1 \val \\ \tau_2 \val}{\tau_1 \tau_2 \texttt{|} \val} \and
-        \Rule{V-7}{\tau_1 \val \\ \tau_2 \val}{\kwArrow{\tau_1}{\tau_2} \val} \and
-        % fixed point combinator
-        \Rule{Fixed Point}{e \val \\ \tau \val}{\kwfix{e}{\tau} \val}
-    \end{mathpar}  
-}
+        % Tuple literals evaluation
+        \Rule{TUPLES-1}
+            {e^* \step e^{*\prime}}
+            {( e^* ) \step ( e^{*\prime} )}
+        \Rule{TUPLES-2}
+            {v_i \val ~ \forall v_i \in v^*}
+            {(v^*) \val}
+        \and
 
-\subsection{Step Judgement}
-Expressions are evaluated from left to right until everything is a value
-{
-    \centering
-    \def \MathparLineskip {\lineskip=0.43cm}
-    \begin{mathpar}
         % Derivation of e* -> e*'
@@
             {e_1 \step e_1'
-                \\v_i \val ~ \forall v_i \in v*}
-            {v* ~ e_1 ~ ... ~ e_n \step v* ~ e_1' ~ ... ~ e_n}
+                \\v_i \val ~ \forall~ v_i \in v^*}
+            {v^* ~ e_1 ~ ... ~ e_n \step v^* ~ e_1' ~ ... ~ e_n}
         \and
@@
         % Closures capture lexically scoped variables via substitution 
-        \Rule{LET-1}
+        \Rule{LET}
             {v \val
@@
 
-        \and \Rule{MACRO-1}
-            {\tau^* \step \tau^{\prime*}}
-            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^{\prime*}}{e^*}}
+        \and \Rule{Identifier}
+            {[ ... ~ \kwid{x} \step v ~ ... ] \in \sigma}
+            {\sigma; \kwid{x} \step \sigma; v}
 
@@
             {\uplus^{|\sigma|}_{i = 1} \sigma(i) = [x_1 \step v_1 ~ ... ~ x_n \step v_n]
-                \\ \tau_i \val~ \forall~ \tau_i \in \tau*}
+                \\ \tau_i \val~ \forall~ \tau_i \in \tau^*}
             {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^*}{[v_1/x_1]...[v_n/x_n] e^*} \val}
@@
 
-        \and \Rule{Identifier}
-            {[ ... ~ \kwid{x} \step v ~ ... ] \in \sigma}
-            {\sigma; \kwid{x} \step \sigma; v}
     \end{mathpar}  
@@
 
-        % Tuple literals evaluation
-        \Rule{TUPLES-1}
-            {e^* \step e^{*\prime}}
-            {( e^* ) \step ( e^{*\prime} )}
-        \Rule{TUPLES-2}
-            {v_i \val ~ \forall v_i \in v*}
-            {(v*) \val}
-        \and
-
         % unpack operator
@@
             {\kwunpack{()} \step \varepsilon}
-        \and
-        \Rule{TUPLES-5}
-            {\tau_i \val ~ \forall~ i \in [1, n]}
-            {\kwunpack{(\tau_1 ... \tau_n)} \step \tau_1 ... \tau_n}
-        \Rule{TUPLES-6}
-            {\strut}
-            {\kwunpack{\kwUnit} \step \varepsilon}
         \and
@@
         \and
+    \end{mathpar}
+}
 
-        % typeof operator
-        \Rule{TYPE-1}{v \val \\ \etyped{v}{\tau}}{\kwType{v} \step \tau} \and
-        % Move this to type rules
-        % \Rule{Type Operator}
-        %     {\tau^* = \tau_1 ~...~ \tau_n \\
-        %      v^* = \etyped{v_1}{\tau_1} ~...~ \etyped{v_1}{\tau_1} \\
-        %      m = \kwmacro{\tau^*}{e^*} \\
-        %      m \val }
-        %     {\kwType{m} \step \kwArrow{(\tau^*)}{\kwType{(v^*~ m~ \textbf{@})}}}
+\section{Semantics for Types}
+Because there are operations involving types, we have to have a equivalents to $\val$ and $\step$ judgements for types.
+These are provided by $\prim$ and $\tstep$ respectively with $\prim$ loosely referring to "primitive" types.
+
+\subsection{Primitives}
+{
+    \centering
+    \def \MathparLineskip {\lineskip=0.43cm}
+    \begin{mathpar}
+        \Rule{P-I32}{\strut}{\kwInt \prim} \and
+        \Rule{P-I64}{\strut}{\kwLong \prim} \and
+        \Rule{P-F32}{\strut}{\kwFloat \prim} \and
+        \Rule{P-F64}{\strut}{\kwDouble \prim} \and
+        \Rule{P-sum}{\tau_1 \prim \\ \tau_2 \prim}{\tau_1 \tau_2 \texttt{|} \prim} \and
+        \Rule{P-Arrow}{\tau_1 \prim \\ \tau_2 \prim}{\kwArrow{\tau_1}{\tau_2} \prim} \and
+    \end{mathpar}  
+}
+
+\subsection{Step Judgement and Type operator}
+Expressions are evaluated from left to right until everything is a value
+{
+    \centering
+    \def \MathparLineskip {\lineskip=0.43cm}
+    \begin{mathpar}
+        % Derivation of e* -> e*'
+        \Rule{STEP-1}
+            {\tau_1 \tstep \tau_1'}
+            {\tau_1 ~ ... ~ \tau_n \tstep \tau_1' ~ ... ~ \tau_n}
+        \and
+        \Rule{STEP-2}
+            {\tau_1 \tstep \tau_1'
+                \\v_i \prim ~ \forall v_i \in v^*}
+            {\tau^* ~ \tau_1 ~ ... ~ \tau_n \tstep v^* ~ \tau_1' ~ ... ~ \tau_n}
+        \and
+
+        \Rule{TYPE-1}{v \val \\ \etyped{v}{\tau}}{\kwType{v} \tstep \tau} 
+    \end{mathpar}  
+}
+
+\subsection{Type Identifiers}
+Type identifiers use the same store as used for values, however their identifiers are guaranteed to not conflict as type identifiers start with a capitol letter vs lower-case for value identifiers.
+Because we use the same store the step semantics for closures and scopes continue to work.
+{
+    \centering
+    \def \MathparLineskip{\lineskip=0.43cm}
+    \begin{mathpar}
+        \Rule{Type-Let}
+            {\tau \prim
+                \\ \sigma = [ [...], ... [ T_1 \tstep \tau_1 ... T_n \tstep \tau_n] ]
+                \\ \sigma' = [ [...], ... [ T_1 \tstep \tau_1 ... T_n \tstep \tau_n, T_{n+1} \tstep \tau] ]}
+            {\sigma; \tau ~ \kweid{T} = \tstep \sigma'; \varepsilon}
+
+        \and \Rule{Type-Id}
+            {[ ... ~ \kwid{T} \tstep \tau ~ ... ] \in \sigma}
+            {\sigma; \kwid{T} \tstep \sigma; \tau}
+
+        \and \Rule{MACRO-1}
+            {\tau^* \tstep \tau^{*\prime}}
+            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^{*\prime}}{e^*}}
+    \end{mathpar}
+}
+
+\subsection{Others}
+Notably the step
+{
+    \centering
+    \def \MathparLineskip{\lineskip=0.43cm}
+    \begin{mathpar}
+        % Tuple type literals evaluation
+        \Rule{TTUPLES-1}
+            {\tau^* \tstep \tau^{*\prime}}
+            {( \tau^* ) \tstep ( \tau^{*\prime} )}
+        \Rule{TTUPLES-2}
+            {\tau_i \prim ~ \forall \tau_i \in \tau^*}
+            {(\tau^*) \prim}
+        \and
+
+        % unpack operator
+        \Rule{TTUPLES-3}
+            {\tau_i \prim ~ \forall~ i \in [1, n]}
+            {\kwunpack{(\tau_1 ... \tau_n)} \tstep \tau_1 ... \tau_n}
+        \Rule{TTUPLES-4}
+            {\strut}
+            {\kwunpack{\kwUnit} \tstep \varepsilon}
+        \and
+
+        % fixed point combinator
+        \Rule{Fixed Point}{e \val \\ \tau \prim}{\kwfix{e}{\tau} \val}
     \end{mathpar}
@@
     \Rule{T-0}
-        {\tau \step \tau' \\ \typed{\ctx}{e}{\tau}}
+        {\typed{\ctx}{e}{\tau} \\ \tau \tstep \tau'}
         {\typed{\ctx}{e}{\tau'}}
@@
             {\typed{\ctx}{\kwunpack{(e^*)}}{\tau^*}}
-        % equivalent types?
-        % \and \Rule{Unpack types}
-        %     {\typed{\ctx}{e*}{\kwunpack{(\tau^*)}}}
-        %     {\typed{\ctx}{e*}{\tau*}}
     \end{mathpar}
@@
 \subsection{Macros}
+This is why we needed the step judgement :sweatsmile:
 {
@@
     \begin{mathpar}
+        % Move this to type rules   
+        \Rule{Macro}
+            {\tau^* = \tau_1 ~...~ \tau_n \\
+             v^* = \etyped{v_1}{\tau_1} ~...~ \etyped{v_n}{\tau_n} \\
+             m = \kwmacro{\tau^*}{e^*} \\
+             m \val }
+            {\typed{\ctx}{m}{\kwArrow{(\tau^*)}{\kwType{(v^*~ m~ \textbf{@})}}}}
+
+        \Rule{Call}
+            {\typed{\ctx}{e_1}{\kwArrow{\tau^*}{\tau^{*\prime}}}
+                \\ \typed{\ctx}{e^*}{\tau^*}}
+            {\typed{\ctx}{e^*~ e_1~ \textbf{@}}{\tau^{*\prime}}}
+
+        \Rule{Identifier}
+            {\ctx(\kwid{x}) = \tau}
+            {\typed{\ctx}{\kwid{x}}{\tau}}
 
@@
 }
+
+\section{Proofs}
+\subsection{Theorem - Progress}
+If $\etyped{e^*}{\tau^*}$ then either $e^*\step e^{*\prime}$ or $e_i \val~ \forall~ e_i \in e^*$.
+\begin{proof}
+    By induction on the derivation of $\etyped{e^*}{\tau^*}$.
+    \begin{itemize}
+        \item (T-\{1,2,3,4\})
+            Then $e^* = e = \kwnumlit{n}{\tau}$, where $\tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble \}$.
+            By (V-1), $\kwnumlit{n}{\tau} \val$, so $e^* \val$.
+
+        \item (T-5)
+            Then $e^* = e = e_1~ e_2~ +$ and $\kwtyped{e_1}{\tau}$ and $\kwtyped{e_2}{\tau}$ for some $\tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble \}$.
+
+            By induction, $e_1 \step e_1'$ or $e_1 val$.
+            If $e_1 \step e_1'$, then by \Rulename{STEP-1}, $e \step e_1'~ e_2~ +$.
+
+            If $e_1 \val$, then by induction, $e_2 \step e_2'$ or $e_2 val$.
+            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP-2}, $e \step e_1~ e_2'~ +$.
+
+            If $e_1 \val$ and $e_2 \val$, then $e_1 = \kwnumlit{n_1}{\tau}, e_2 = \kwnumlit{n_2}{\tau}$, and by \Rulename{ADD}, $e+ \step \kwnumlit{n_1 + n_2}{\tau}$.
+    \end{itemize}
+\end{proof}
+
+\subsection{Theorem - Preservation}
+If $\typed{\ctx}{e^*}{\tau^*}$ and $e^* \step e^{*\prime}$ then $\typed{\ctx}{e^{*\prime}}{\tau^{*\prime}}$ where $\tau^{*\prime} \tstep^* \tau^*$
+\begin{proof}
+
+\end{proof}
+
 \end{document}

### 2021-11-30 project/paper.tex
fix some rules, remove sum types

@@
 \usepackage[colorlinks=true, allcolors=blue]{hyperref}
-
-% ???
 \usepackage{syntax}
@@
     |    $\kwUnit$
-    |    $\tau_1~ \tau_2~ \texttt{|}$
+    % |    $\tau_1~ \tau_2~ \texttt{|}$
     |    $\kwArrow{\tau_1} {\tau_2}$
@@
 \end{grammar}
-
 Notes
@@
     \item A syntactic fixed point combinator was chosen instead of $\kwrec{\tau^*}{\tau^*}{e^*}$. It's possible that one could be implemented in the language itself.
-    \item notice that unpack can have any non-negative number of output values depending on the given input
+    \item Tuples and Tuple types are enclosed in parens
     \item $\kwunpack{e}$ and $\kwunpack{\tau}$ can give 0 or more outputs depending on the input tuple
@@
 
-\subsection{Value Judgements}
+\subsection{Value and Step Judgement}
+The step judgement states that expressions are evaluated from left to right until everything is a value. Values are defined further as we go through.
 {
@@
     \begin{mathpar}
+        % Derivation of e* -> e*'
+        \Rule{STEP}
+            {e_1 \step e_1'
+                \\v_i \val ~ \forall~ v_i \in v^*}
+            {v^* ~ e_1 ~ ... ~ e_n \step v^* ~ e_1' ~ ... ~ e_n}
+        \and
 
-        % fixed point combinator
-        \Rule{Fixed Point}{e \val \\ \tau \prim}{\kwfix{e}{\tau} \val}
-    \end{mathpar}
-}
-
-\subsection{Value and Step Judgement}
-Expressions are evaluated from left to right until everything is a value
-{
-    \centering
-    \def \MathparLineskip {\lineskip=0.43cm}
-    \begin{mathpar}
         % number literals
@@
             {(v^*) \val}
-        \and
-
-        % Derivation of e* -> e*'
-        \Rule{STEP-1}
-            {e_1 \step e_1'}
-            {e_1 ~ ... ~ e_n \step e_1' ~ ... ~ e_n}
-        \and
-        \Rule{STEP-2}
-            {e_1 \step e_1'
-                \\v_i \val ~ \forall~ v_i \in v^*}
-            {v^* ~ e_1 ~ ... ~ e_n \step v^* ~ e_1' ~ ... ~ e_n}
         \and
@@
 \subsection{Others}
+Notice the use of the \href{https://en.wikipedia.org/wiki/Kronecker_delta}{Kronecker Delta} to define the equals operator.
 {
@@
             {v_1~ v_2~ + \step v_3}
+        \and
         \Rule{MUL}
@@
 
-        % ternary
+        % Comparisons
+        \Rule{EQUIV}
+            {v_1 \val \\ v_2 \val}
+            {v_1~ v_2~ ==~ \step \kwnumlit{(\delta_{v_1 v_2})}{\kwInt}}
+
+        % Ternary
         \Rule{TERN-1}
@@
 \section{Semantics for Types}
-Because there are operations involving types, we have to have a equivalents to $\val$ and $\step$ judgements for types.
-These are provided by $\prim$ and $\tstep$ respectively with $\prim$ loosely referring to "primitive" types.
+Because there are operations involving types, we have to have a equivalents to $\val$ and $\step$ judgements for types. These are provided by $\prim$ and $\tstep$ respectively.
 
@@
         \Rule{P-F64}{\strut}{\kwDouble \prim} \and
-        \Rule{P-sum}{\tau_1 \prim \\ \tau_2 \prim}{\tau_1 \tau_2 \texttt{|} \prim} \and
+        %\Rule{P-sum}{\tau_1 \prim \\ \tau_2 \prim}{\tau_1 \tau_2 \texttt{|} \prim} \and
         \Rule{P-Arrow}{\tau_1 \prim \\ \tau_2 \prim}{\kwArrow{\tau_1}{\tau_2} \prim} \and
@@
 
-\subsection{Step Judgement and Type operator}
-Expressions are evaluated from left to right until everything is a value
+\subsection{Step Judgement}
+Type expressions are evaluated left to right until everything is $\prim$.
 {
@@
     \begin{mathpar}
-        % Derivation of e* -> e*'
-        \Rule{STEP-1}
-            {\tau_1 \tstep \tau_1'}
-            {\tau_1 ~ ... ~ \tau_n \tstep \tau_1' ~ ... ~ \tau_n}
+        % Derivation of T* -> T*'
+        \Rule{TSTEP}
+            {\tau_1 \tstep \tau_1'
+                \\\tau_i \prim ~ \forall \tau_i \in \tau^*}
+            {\tau^* ~ \tau_1 ~ ... ~ \tau_n \tstep \tau^* ~ \tau_1' ~ ... ~ \tau_n}
         \and
-        \Rule{STEP-2}
-            {\tau_1 \tstep \tau_1'
-                \\v_i \prim ~ \forall v_i \in v^*}
-            {\tau^* ~ \tau_1 ~ ... ~ \tau_n \tstep v^* ~ \tau_1' ~ ... ~ \tau_n}
-        \and
-
-        \Rule{TYPE-1}{v \val \\ \etyped{v}{\tau}}{\kwType{v} \tstep \tau} 
     \end{mathpar}  
@@
             {\sigma; \kwid{T} \tstep \sigma; \tau}
-
-        \and \Rule{MACRO-1}
-            {\tau^* \tstep \tau^{*\prime}}
-            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^{*\prime}}{e^*}}
     \end{mathpar}
@@
 
-\subsection{Others}
-Notably the step
+\subsection{Tuples}
 {
@@
         % Tuple type literals evaluation
-        \Rule{TTUPLES-1}
+        \Rule{Tuple Types - 1}
             {\tau^* \tstep \tau^{*\prime}}
             {( \tau^* ) \tstep ( \tau^{*\prime} )}
-        \Rule{TTUPLES-2}
+        \Rule{Tuple Types - 2}
             {\tau_i \prim ~ \forall \tau_i \in \tau^*}
@@
         % unpack operator
-        \Rule{TTUPLES-3}
+        \Rule{Tuple Types - 3}
             {\tau_i \prim ~ \forall~ i \in [1, n]}
             {\kwunpack{(\tau_1 ... \tau_n)} \tstep \tau_1 ... \tau_n}
-        \Rule{TTUPLES-4}
+        \Rule{Tuple Types - 4}
             {\strut}
@@
         \and
+    \end{mathpar}
+}
+
+\subsection{Others}
+{
+    \centering
+    \def \MathparLineskip{\lineskip=0.43cm}
+    \begin{mathpar}
+        \Rule{Type Equiv}
+            {\tau_1 \prim \\ \tau_2 \prim}
+            {\tau_1~ \tau_2~ ==~ \tstep~ \kwnumlit{(\delta_{\tau_1 \tau_2})}{\kwInt}}
 
         % fixed point combinator
-        \Rule{Fixed Point}{e \val \\ \tau \prim}{\kwfix{e}{\tau} \val}
+        \and \Rule{Fixed Point}{e \val \\ \tau \prim}{\kwfix{e}{\tau} \val}
+
+        % input types for macros can step
+        \and \Rule{Macro Type Step}
+            {\tau^* \tstep \tau^{*\prime}}
+            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^{*\prime}}{e^*}}
+
+        % typeof operator
+        \Rule{TYPE-1}{v \val \\ \etyped{v}{\tau}}{\kwType{v} \tstep \tau} 
     \end{mathpar}
@@
 \section{Typing}
-
+Note that because our notation is defined as ignoring irrelevant values there could be  
 \subsection{Step Judgement}
@@
             {\typed{\ctx}{\kwtern{e_1}{e_2}{e_3}}{\tau}}
+        \and \Rule{T-8}
+            {\typed{\ctx}{e_1}{\tau}
+                \\ \typed{\ctx}{e_2}{\tau}
+                \\ \tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble \}}
+            {\typed{\ctx}{e_1~ e_2~ ==~}{\kwInt}}
+        \and \Rule{T-9}
+            {\tau_1 \prim
+                \\ \tau_2 \prim}
+            {\typed{\ctx}{\tau_1~ \tau_2~ ==~}{\kwInt}}
     \end{mathpar}
@@
 
-        \Rule{Call}
+        \and \Rule{Call}
             {\typed{\ctx}{e_1}{\kwArrow{\tau^*}{\tau^{*\prime}}}
@@
 
-        \Rule{Identifier}
+        \and \Rule{Identifier}
             {\ctx(\kwid{x}) = \tau}
             {\typed{\ctx}{\kwid{x}}{\tau}}
+        \and \Rule{Let}
+            {\typed{\ctx}{e}{\tau}
+                \\ \ctx(\kwid{x}) = e}
+            {\typed{\ctx}{e \kwid{x}}{\varepsilon}}
 
+        \and \Rule{Fix}
+            {\tau_1 \prim
+                \\ \typed{\ctx}{e}{\tau_2}
+                \\ \tau_2 \prim
+                \\ \tau_2 = \kwArrow{\tau^*_3 ~ \kwArrow{\tau^*_3}{\tau^*_4}}{\tau^*_4}}
+            {\typed{\ctx}{\kwfix{e}{\tau_1}}{\kwArrow{\tau^*_3}{\tau^*_4}}}
     \end{mathpar}
@@
         \item (T-\{1,2,3,4\})
-            Then $e^* = e = \kwnumlit{n}{\tau}$, where $\tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble \}$.
+            Then $e^* = \kwnumlit{n}{\tau}$, where $\tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}$.
             By (V-1), $\kwnumlit{n}{\tau} \val$, so $e^* \val$.
@@
         \item (T-5)
-            Then $e^* = e = e_1~ e_2~ +$ and $\kwtyped{e_1}{\tau}$ and $\kwtyped{e_2}{\tau}$ for some $\tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble \}$.
+            Then $e^* = e_1~ e_2~ +$ and $\etyped{e_1}{\tau}$ and $\etyped{e_2}{\tau}$ for some $\tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}$.
 
-            By induction, $e_1 \step e_1'$ or $e_1 val$.
-            If $e_1 \step e_1'$, then by \Rulename{STEP-1}, $e \step e_1'~ e_2~ +$.
+            By induction, $e_1 \step e_1'$ or $e_1 \val$.
+            If $e_1 \step e_1'$, then by \Rulename{STEP-1}, $e^* \step e_1'~ e_2~ +$.
 
-            If $e_1 \val$, then by induction, $e_2 \step e_2'$ or $e_2 val$.
-            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP-2}, $e \step e_1~ e_2'~ +$.
+            If $e_1 \val$, then by induction, $e_2 \step e_2'$ or $e_2 \val$.
+            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP-2}, $e^* \step e_1~ e_2'~ +$.
 
-            If $e_1 \val$ and $e_2 \val$, then $e_1 = \kwnumlit{n_1}{\tau}, e_2 = \kwnumlit{n_2}{\tau}$, and by \Rulename{ADD}, $e+ \step \kwnumlit{n_1 + n_2}{\tau}$.
+            If $e_1 \val$ and $e_2 \val$, then by CF, $e_1 = \kwnumlit{n_1}{\tau}, e_2 = \kwnumlit{n_2}{\tau}$, and by \Rulename{ADD}, $e^* \step \kwnumlit{n_1 + n_2}{\tau}$.
+
+        \item (T-6)
+            $[\times / +][\Rulename{MUL} / \Rulename{ADD}]$ (T-5)
+
+        \item (T-7)
+            Then $e^* = \kwtern{e_1}{e_2}{e_3}$ and $\etyped{e_1}{\kwInt}$. %and $\etyped{e_2}{\tau}$ and $\etyped{e_3}{\tau}$ for some $\tau$.
+
+            By induction, $e_1 \step e_1'$ or $e_1 \val$.
+            If $e_1 \step e_1'$, then by \Rulename{STEP-1}, $e^* \step \kwtern{e_1'}{e_2}{e_3}$.
+
+            If $e_1 \val$, then by induction, $e_2 \step e_2'$ or $e_2 \val$.
+            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP-2}, $e^* \step \kwtern{e_1}{e_2'}{e_3}$.
+
+            If $e_1 \val$ and $e_2 \val$, then by induction, $e_3 \step e_3'$ or $e_3 \val$.
+            If $e_1 \val$ and $e_2 \val$ and $e_3 \step e_3'$, then by \Rulename{STEP-2}, $e \step \kwtern{e_1}{e_2}{e_3'}$.
+
+            If $e_1 \val$ and $e_2 \val$ and $e_3 \val$, then by CF, $e_1 = \kwnumlit{n}{\kwInt}$.
+
+            If $n = 0$, then by \Rulename{TERN-1}, $e^* \step e_3$.
+            Otherwise, $n \neq 0$, so by \Rulename{TERN-2}, $e^* \step e_2$.
+
+        \item \Rulename{Empty Tuple}
+            Then $e^* = ()$. By \Rulename{TUPLES-2}, $() \val$.
+
+        \item \Rulename{Tuple Type}
+            Then $e^* = (e^*_t) = (e_1~ e_2~ ...~ e_n)$.
+
+            By induction, $\forall e_i \in e^*_t$, $e_i \val$ or $e_i \step e_i'$.
+
+            % Then either all elements are values or there exists some *first* steppable element
+            % if there's a steppable element, then e^*_t steps using STEP-{1,2}, so e^* steps using TUPLES-1
+            % if there are no steppable elements, then all elements are values, so e^* \val by TUPLES-2
     \end{itemize}

### 2021-12-01 project/paper.tex
day before presentation

@@
 \section{Syntax}
-Taking the ideas from the above rant into account we arrive at something along these lines:
+Here some operations also include their stack arguments, this is intended only for clarity as to what their function is as their arguments are come from the stack when it's their turn to be evaluated, not syntactically.
 
@@
     |    $\kwDouble$
-    \alt $(\tau^*)$
+    |    $(\tau^*)$
     |    $\kwUnit$
@@
     |    $\kwArrow{\tau_1} {\tau_2}$
-    |    $\kwType{e}$
-
-    <$\tau^*$> ::= $\varepsilon$ | $\tau^* ~ \tau$ | $\kwunpack{\tau}$
+    \alt $\kwType{e}$
+    |    $e~ \kweid{T} =$
+    |    $\kwid{T}$
+    |    $\kwunpack{\tau}$
 
@@
     |    $\kwfix{e}{\tau}$
-    \alt $\kwid{x}$
-    |    $\kwid{T}$
+    |    $\kwnumlit{n}{\kwInt}$
+    |    $\kwnumlit{n}{\kwLong}$
+    |    $\kwnumlit{n}{\kwFloat}$
+    |    $\kwnumlit{n}{\kwDouble}$
+    \alt $e~ \kweid{x} =$
+    |    $\kwid{x}$
     |    $e~ \textbf{@} $
-    |    $e~ \kweid{x} =$
+    |    $\kwunpack{e}$
 
-    <$e^*$> ::= $\varepsilon$ | $e^* ~ e$ | $\kwunpack{e}$
+    <$prog$> ::= $e^*$ | $\tau^*$ | $prog~ prog$ | $\varepsilon$
 \end{grammar}
@@
 \begin{itemize}
+    \item A program consists of a list of $op^*$ expressions
+    \item $\varepsilon$ is used to denote an empty program/expression
     \item A syntactic fixed point combinator was chosen instead of $\kwrec{\tau^*}{\tau^*}{e^*}$. It's possible that one could be implemented in the language itself.
@@
 
-\section{Semantics}
+\section{Small Step Semantics For Values}
 \subsection{Notation}
@@
             {v \val
-                \\ \sigma = [ [...], ... [ x_1 \step v_1 ... x_n \step v_n] ]
-                \\ \sigma' = [ [...], ... [ x_1 \step v_1 ... x_n \step v_n, x_{n+1} \step v] ]}
+                \\ \sigma = [ [...], ... [ x_1 \step v_1 ~...~ x_n \step v_n] ]
+                \\ \sigma' = [ [...], ... [ x_1 \step v_1 ~...~ x_n \step v_n, x_{n+1} \step v] ]}
             {\sigma; v ~ \kweid{x} = \step \sigma'; \varepsilon}
@@
         \and \Rule{Identifier}
-            {[ ... ~ \kwid{x} \step v ~ ... ] \in \sigma}
+            { \kwid{x} \step v \in \sigma(-1)}
             {\sigma; \kwid{x} \step \sigma; v}
 
-        \and \Rule{MACRO-2}
-            {\uplus^{|\sigma|}_{i = 1} \sigma(i) = [x_1 \step v_1 ~ ... ~ x_n \step v_n]
-                \\ \tau_i \val~ \forall~ \tau_i \in \tau^*}
-            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^*}{[v_1/x_1]...[v_n/x_n] e^*} \val}
-
         \and \Rule{CALL-1}
-            {\sigma' = \sigma \cup [~] }
-            {\sigma; \kwmacro{\tau^*}{e^*} \textbf{@} \step \sigma'; e^* \kw{end\_scope}}
+            {\sigma' = \sigma \cup [~]
+                \\ v = \kwmacro{\tau^*}{e^*}
+                \\ v \val}
+            {\sigma; v~ \textbf{@} \step \sigma'; e^* \kw{end\_scope}}
 
@@
     \begin{mathpar}
-
         % unpack operator
         \Rule{TUPLES-3}
-            {v_i \val ~ \forall~ i \in [1, n]}
-            {\kwunpack{(v_1 ... v_n)} \step v_1 ... v_n}
-        \Rule{TUPLES-4}
-            {\strut}
-            {\kwunpack{()} \step \varepsilon}
-        \and
+            {v_i \val ~\forall~ v_i \in v^*}
+            {\kwunpack{(v^*)} \step v^*}
 
@@
 
-\section{Semantics for Types}
+\section{Small Step Semantics for Types}
 Because there are operations involving types, we have to have a equivalents to $\val$ and $\step$ judgements for types. These are provided by $\prim$ and $\tstep$ respectively.
@@
 \subsection{Step Judgement}
-Type expressions are evaluated left to right until everything is $\prim$.
+Type expressions are evaluated left to right until everything is $\prim$. Types and expressions have equal precedence and both can be present on the stack.
 {
@@
         \Rule{TSTEP}
-            {\tau_1 \tstep \tau_1'
-                \\\tau_i \prim ~ \forall \tau_i \in \tau^*}
-            {\tau^* ~ \tau_1 ~ ... ~ \tau_n \tstep \tau^* ~ \tau_1' ~ ... ~ \tau_n}
-        \and
+            {\tau \tstep \tau'
+                \\\tau_i \prim ~ \forall~ \tau_i \in \tau_1^*}
+            {\tau_1^* ~ \tau ~ \tau_2^* \tstep \tau_1^* ~ \tau' ~ \tau_2^*}
+
+        \and \Rule{TSTEP - Mixed}
+            {\tau^* \tstep \tau^{*\prime}
+                \\ v_i \val ~ \forall v_i \in v^*}
+            {v^*~ \tau^*~ e^*~ \tstep v^*~ \tau^{*\prime}~ e^*}
+        \and \Rule{STEP - Mixed}
+            {e^* \step e^{*\prime}
+                \\ \tau_i \prim ~ \forall~ \tau_i \in \tau^*}
+            {\tau^*~ e^* ~ \tau_2^*  \step \tau^* ~ e^{*\prime} ~ \tau_1}
     \end{mathpar}
@@
 
+        \and \Rule{MACRO-2}
+            {\sigma = [ ..., [x_1 \step v_1 ~ ... ~ x_n \step v_n, 
+                ~ T_1 \tstep \tau_1 ~ ... ~ T_n \tstep \tau_n ]]
+                \\ \tau_i \val~ \forall~ \tau_i \in \tau^*}
+            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^*}{[v_1/x_1]...[v_n/x_n][\tau_1/T_1]...[\tau_n/T_n] e^*} \val}
+
         \and \Rule{Type-Id}
-            {[ ... ~ \kwid{T} \tstep \tau ~ ... ] \in \sigma}
+            { \kwid{T} \step \tau \in \sigma(-1)}
             {\sigma; \kwid{T} \tstep \sigma; \tau}
@@
         \and
-
         % unpack operator
         \Rule{Tuple Types - 3}
-            {\tau_i \prim ~ \forall~ i \in [1, n]}
-            {\kwunpack{(\tau_1 ... \tau_n)} \tstep \tau_1 ... \tau_n}
-        \Rule{Tuple Types - 4}
-            {\strut}
-            {\kwunpack{\kwUnit} \tstep \varepsilon}
-        \and
+            {\tau_i \prim ~ \forall~ \tau_i \in \tau^*}
+            {\kwunpack{(\tau^*)} \tstep \tau_1 ... \tau_n}
     \end{mathpar}
@@
 \section{Typing}
-Note that because our notation is defined as ignoring irrelevant values there could be  
+Note that we're using a somewhat weird notation for a stack machine (normally there's arrows representing stack transformation). This aligns with how we defined our syntax.
+
 \subsection{Step Judgement}
@@
 \subsection{Macros}
-This is why we needed the step judgement :sweatsmile:
+This is why we needed the step judgement
 {
@@
              m \val }
-            {\typed{\ctx}{m}{\kwArrow{(\tau^*)}{\kwType{(v^*~ m~ \textbf{@})}}}}
+            {\typed{\ctx}{\kwmacro{\tau^*}{e^*}}{\kwArrow{(\tau^*)}{\kwType{(v^*~ m~ \textbf{@})}}}}
 
@@
             {\typed{\ctx}{e}{\tau}
-                \\ \ctx(\kwid{x}) = e}
-            {\typed{\ctx}{e \kwid{x}}{\varepsilon}}
+                \\ \ctx(\kwid{x}) = \tau}
+            {\typed{\ctx}{e~ \kwid{x} =}{\varepsilon}}
 
@@
             {\tau_1 \prim
-                \\ \typed{\ctx}{e}{\tau_2}
-                \\ \tau_2 \prim
-                \\ \tau_2 = \kwArrow{\tau^*_3 ~ \kwArrow{\tau^*_3}{\tau^*_4}}{\tau^*_4}}
-            {\typed{\ctx}{\kwfix{e}{\tau_1}}{\kwArrow{\tau^*_3}{\tau^*_4}}}
+                \\ \tau_1 = \kwArrow{(\tau^*_3)}{(\tau^*_4)}
+                \\ \typed{\ctx}{e}{\kwArrow{(\tau^*_3 ~ \tau_1)}{(\tau^*_4)}}}
+            {\typed{\ctx}{\kwfix{e}{\tau_1}}{\tau_1}}
     \end{mathpar}
@@
             By induction, $e_1 \step e_1'$ or $e_1 \val$.
-            If $e_1 \step e_1'$, then by \Rulename{STEP-1}, $e^* \step e_1'~ e_2~ +$.
+            If $e_1 \step e_1'$, then by \Rulename{STEP}, $e^* \step e_1'~ e_2~ +$.
 
             If $e_1 \val$, then by induction, $e_2 \step e_2'$ or $e_2 \val$.
-            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP-2}, $e^* \step e_1~ e_2'~ +$.
+            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP}, $e^* \step e_1~ e_2'~ +$.
 
@@
             By induction, $e_1 \step e_1'$ or $e_1 \val$.
-            If $e_1 \step e_1'$, then by \Rulename{STEP-1}, $e^* \step \kwtern{e_1'}{e_2}{e_3}$.
+            If $e_1 \step e_1'$, then by \Rulename{STEP}, $e^* \step \kwtern{e_1'}{e_2}{e_3}$.
 
             If $e_1 \val$, then by induction, $e_2 \step e_2'$ or $e_2 \val$.
-            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP-2}, $e^* \step \kwtern{e_1}{e_2'}{e_3}$.
+            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP}, $e^* \step \kwtern{e_1}{e_2'}{e_3}$.
 
             If $e_1 \val$ and $e_2 \val$, then by induction, $e_3 \step e_3'$ or $e_3 \val$.
-            If $e_1 \val$ and $e_2 \val$ and $e_3 \step e_3'$, then by \Rulename{STEP-2}, $e \step \kwtern{e_1}{e_2}{e_3'}$.
+            If $e_1 \val$ and $e_2 \val$ and $e_3 \step e_3'$, then by \Rulename{STEP}, $e \step \kwtern{e_1}{e_2}{e_3'}$.
 
@@
 
-            % Then either all elements are values or there exists some *first* steppable element
-            % if there's a steppable element, then e^*_t steps using STEP-{1,2}, so e^* steps using TUPLES-1
-            % if there are no steppable elements, then all elements are values, so e^* \val by TUPLES-2
+            If all elements are values, then $(e^*_t) \val$ by \Rulename{TUPLES-2}.
+
+            Otherwise, one or more elements can step, including a leftmost steppable element $e_j \step e_j'$.
+            Then, by \Rulename{STEP}, $e^*_t \step [e_j' / e_j] e^*_t$, so by \Rulename{Tuples-1}, $e^* \step ([e_j' / e_j] e^*_t)$.
+
+        \item \Rulename{Unpack}
+            Then $e^* = \kwunpack{(e^*_t)}$.
+            By induction, either $(e^*_t) \step (e^{*'}_t)$ or $(e^*_t) \val$.
+
+            If $(e^{*'}_t) \step (e^{*'}_t)$, then by \Rulename{STEP}, $\kwunpack{(e^*_t)} \step \kwunpack{(e^{*'}_t)}$.
+
+            If $(e^*_t) \val$, then by \Rulename{TUPLES-3}, $\kwunpack{(e^*_t)} \step e^*_t$.
+
+        \item \Rulename{Macro}
+            Then $e^* = m = \kwmacro{\tau^*}{e^*_t}$. $m \val$ by the premise.
+
+        \item \Rulename{Call}
+            Then $e^* = e^*_x~ e_f~ \textbf{@}$. By induction, either $e_i \val~ \forall~ e_i \in e^*_x$ or $e^*_x \step e^{*'}_x$.
+
+            If $e^*_x \step e^{*'}_x$, then by \Rulename{STEP}, $e^* \step e^{*'}_x~ e_f~ \textbf{@}$.
+
+            Otherwise, $e_i \val~ \forall~ e_i \in e^*_x$ and, by induction, $e_f \val$ or $e_f \step e_f'$.
+
+            If $e_f \step e_f'$, then by \Rulename{STEP}, $e^* \step e^*_x~ e_f'~ \textbf{@}$.
+
+            Otherwise, $e_f \val$. To handwave a bit, by inversion on \Rulename{Macro}, $e_f$ is a macro with a body $e^*_b$, and by \Rulename{CALL-1}, $e^* \step e^*_b ~ \textbf{end_scope}$.
     \end{itemize}
@@
 If $\typed{\ctx}{e^*}{\tau^*}$ and $e^* \step e^{*\prime}$ then $\typed{\ctx}{e^{*\prime}}{\tau^{*\prime}}$ where $\tau^{*\prime} \tstep^* \tau^*$
-\begin{proof}
+% \subsubsection{Lemma - Convergence}
+% Part 1: $\tau^* \tstep \tau^{*\prime}$ is no longer possible when $\tau_i \prim ~ \forall~ \tau_i \in \tau^*$.
+% \begin{proof}
+% Proof by exhaustion with all rules involving $\prim$ types not stepping. To values
+% \end{proof}
+% Part 2: $\tau^* \tstep \tau^{*\prime}$ will eventually step to a $\tau^{*\prime}$ such that $\tau_i \prim ~ \forall~ \tau_i \in \tau^{*\prime}$.
+% \begin{proof}
+% This is provable by the fact that the language does not feature recursive types and the only form of recursion in the language is via the fix operator and $\kwType{}$ operator does no have to follow recursive paths in order to get it's type (Type Rules \Rulename{Macro}, \Rulename{Fix}).
+The other rules are trivial (ie - step to a $\prim$) and/or irrelevant to the assertion
+% \end{proof}
 
-\end{proof}
+% \subsubsection{Proof}
+% \begin{proof}
+% \end{proof}
 

### 2021-12-05 project/paper.tex
after presentation
expand proofs

@@
 \subsection{Type Identifiers}
-Type identifiers use the same store as used for values, however their identifiers are guaranteed to not conflict as type identifiers start with a capitol letter vs lower-case for value identifiers.
+Type identifiers use the same store as used for values, however their identifiers are guaranteed to not conflict as type identifiers start with a capital letter vs lower-case for value identifiers.
 Because we use the same store the step semantics for closures and scopes continue to work.
@@
         % fixed point combinator
-        \and \Rule{Fixed Point}{e \val \\ \tau \prim}{\kwfix{e}{\tau} \val}
+        \and \Rule{Fixed Point}
+            {\kwmacro{\tau^*~ \tau_1}{e^*} \val \\ \tau_1 \prim}
+            {\kwfix{\kwmacro{\tau^*~ \tau_1}{e^*}}{\tau_1} \val}
+        \and \Rule{Call FP}
+            {v = \kwfix{\kwmacro{\tau^*~ \tau_1}{e^*}}{\tau_1} \\ v \val}
+            {v~ \textbf{@} \step v~ \kwmacro{\tau^*~ \tau_1}{e^*}~ \textbf{@}}
 
@@
             {\tau^* \tstep \tau^{*\prime}}
-            {\sigma; \kwmacro{\tau^*}{e^*} \step \sigma; \kwmacro{\tau^{*\prime}}{e^*}}
+            {\sigma; \kwmacro{\tau^*}{e^*} \tstep \sigma; \kwmacro{\tau^{*\prime}}{e^*}}
 
@@
             {\typed{\ctx}{\kwid{x}}{\tau}}
+
         \and \Rule{Let}
@@
             Otherwise, $e_f \val$. To handwave a bit, by inversion on \Rulename{Macro}, $e_f$ is a macro with a body $e^*_b$, and by \Rulename{CALL-1}, $e^* \step e^*_b ~ \textbf{end_scope}$.
+
+        \item \Rulename{Identifier}
+            Then $e^* = \texttt{x}$.
+            By, uh, the other \Rulename{Identifier}, $e^*$ steps to a value in the store.
+
+        \item \Rulename{Let}
+            Then $e^* = e ~ \texttt{x} =$.
+            By induction, $e \val$ or $e \step e'$.
+            If $e \step e'$, then by \Rulename{STEP}, $e^* \step e' ~ \texttt{x} =$.
+            If $e \val$, then by the other \Rulename{LET}, $\sigma;e^* \step \sigma';\varepsilon$, where $\sigma'$ is, as described in the rule, $\sigma$ with $x \step e$ added to its ``topmost'' scope.
+
+        \item \Rulename{Fix}
+            Then $e^* = \kwfix{e}{\tau_1}$.
+            By induction, $e \val$ or $e \step e'$.
+            If $e \step e'$, then by \Rulename{STEP}, $e^* \step \kwfix{e'}{\tau_1}$.
+            If $e \val$, then I think I'm missing something here, but by \Rulename{Fixed Point}, $e^* \val$.
     \end{itemize}
@@
 \subsection{Theorem - Preservation}
-If $\typed{\ctx}{e^*}{\tau^*}$ and $e^* \step e^{*\prime}$ then $\typed{\ctx}{e^{*\prime}}{\tau^{*\prime}}$ where $\tau^{*\prime} \tstep^* \tau^*$
+If $\typed{\ctx}{e^*}{\tau^*}$ and $e^* \step e^{*\prime}$ then $\typed{\ctx}{e^{*\prime}}{\tau^{*\prime}}$ where $\tau^* \tstep^* \tau^{*\prime}$.
+
+\begin{proof}
+By induction on the derivation of $e^* \step e^{*\prime}$.
+\begin{itemize}
+    \item \Rulename{STEP}
+        Then $e^* = v^* e_1 ... e_n$ and $e^{*\prime} \step v^* e_1' ... e_n$ where $\etyped{e_1}{\tau}$ and $\etyped{e_1'}{\tau'}$.
+        By induction, $\tau \tstep^* \tau'$, so by \Rulename{TSTEP}, $\tau^* \tstep^* [\tau' / \tau]\tau^*$.
+
+    \item \Rulename{TUPLES-1}
+        Then $e^* = (e_{elem}^*)$ and $e^* \step (e_{elem}^{*\prime})$ and $\tau^* = (\tau_{elem}^*)$.
+        By induction, $\etyped{e_{elem}^{*\prime}}{\tau_{elem}^{*\prime}}$ where $\tau_{elem}^* \tstep^* \tau_{elem}^{*\prime}$.
+By \Rulename{Tuple Type}, $\etyped{(e_{elem}^{*\prime})}{(\tau_{elem}^{*'})}$, and by \Rulename{Tuple Types - 1}, $(\tau_{elem}^*) \tstep (\tau_{elem}^{*'})$.
+
+    \item \Rulename{LET}
+        Then $e^* = v ~ \texttt{\$x} ~ =$.
+By the other \Rulename{Let}, $\etyped{e^*}{\varepsilon}$, so $\tau^* = \varepsilon$.
+$e^* \step \varepsilon$, so $\tau^{*\prime} = \varepsilon$, so $\tau^* = \tau^{*\prime}$.
+
+    \item \Rulename{Identifier}
+        Then $e^* = \texttt{x}$, where $\texttt{x} \step v$ is in scope.
+By the other \Rulename{Identifier}, the type of an identifier expression is the same as the type of the value in the store ($\etyped{v}{\tau}$), so $\tau^* = \tau^{*\prime} = \tau$.
+
+    \item \Rulename{CALL-1}
+        Then $e^* = $.
+        % I think we need to modify (CALL-1) to require the values? Revisit.
+
+    \item \Rulename{CALL-2}
+        % do we need a type rule for end_scope?
+
+    \item \Rulename{TUPLES-3}
+        Then $e^* = \kwunpack{(v^*)}$, where $\etyped{v^*}{\tau_{elem}^*}$.
+        By \Rulename{Unpack}, $\tau^* = \tau_{elem}^*$. $e^* \step v^*$, so $\tau^{*\prime} = \tau^* = \tau_{elem}^*$.
+
+    \item \Rulename{ADD} \footnote{For these three rules, terms of the form $v_n$ are better expressed as numeric literals.}
+        Then $e^* = v_1~ v_2~ +$ and $e^* \step v_3$, where $\etyped{v_1}{\tau}$ and $\etyped{v_2}{\tau}$ for some $\tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}$.
+        By \Rulename{T-5}, $\etyped{v_1~ v_2~ +}{\tau}$, and by the appropriate choice of \Rulename{T-1}, \Rulename{T-2}, \Rulename{T-3}, or \Rulename{T-4}, $\etyped{v_3}{\tau}$., so $\tau^* = \tau^{*\prime} = \tau$.
+
+    \item \Rulename{MUL}
+        Omitted for brevity on the grounds of being mechanically identical to the above case.
+
+    \item \Rulename{EQUIV}
+        Then $e^* = v_1~ v_2~ ==$ where $\etyped{v_1}{\tau}$ and $\etyped{v_2}{\tau}$ for some $\tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}$.
+        By \Rulename{T-8}, $\tau^* = \kwInt$.
+        $e^* \step \kwnumlit{n}{\kwInt}$, and by \Rulename{T-1}, $\etyped{\kwnumlit{n}{\kwInt}}{\kwInt}$, so $\tau^{*\prime} = \tau^* = \kwInt$.
+
+    \item \Rulename{TERN-1}, \Rulename{TERN-2}
+        Then $e^* = \kwtern{\kwnumlit{n}{\kwInt}}{v_1}{v_2}$ and $\etyped{v_1}{\tau}$ and $\etyped{v_2}{\tau}$ for some $\tau$.
+        By \Rulename{T-7}, $\etyped{e^*}{\tau}$, so $\tau^* = \tau$.
+        Either $e^* \step v_1$ or $e_* \step v_2$, so in either case, $\tau^{*\prime} = \tau$.
+\end{itemize}
+
+\end{proof}
+
 % \subsubsection{Lemma - Convergence}

### 2021-12-07 project/paper.tex
Clean up first part

@@
 
-\section{Motivation}
+\section{Introduction}
+\subsection{Motivation}
 WebAssembly is a typed, stack-machine based assembly language which is widely supported across all major browsers.
@@
 
-The language this paper would be focused on is currently being used to power \href{https://ridder.xyz/game}{some insignificant demos}, but maybe in some distant future it will end up as the backend for a visual programming tool or some other application where guarantees of safety which could be produced here would be very important.
+\href{https://github.com/dvtate/postfix-haskell}{The language this paper focuses on} is currently being used to power \href{https://ridder.xyz/game}{some insignificant demos}, but maybe in some distant future it will end up as the backend for a visual programming tool or some other application where guarantees of safety which could be produced here would be very important.
 
-\section{The Language}
+\subsection{The Language}
 This section describes some of the key features of the original language. It is included only to show that the subset is a good representation of the original language and thus can be skipped.
 
-\subsection{Syntax}
+\subsubsection{Syntax}
 \begin{itemize}
@@
 
-\subsection{Preliminary Work - Type System}
+\subsubsection{Type System}
 The language currently features an incomplete type system (with some compile-time-only values not fitting into the type system).
@@
 
-\section{The Subset}
-Here we describe at a high level the subset of the language to formalize.
-
-\subsection{Branching}
-As mentioned before a ternary operator would be a reasonable subset of branching via function overloading. The tern operator is defined as an operator which
-
-\begin{verbatim}
-# Tern takes an I32 and two values of the same type
-# and gives one of those two values
-$tern ~ :type # :type - (I32 $A $A) ($A) Arrow
-
-# As it is a subset we can define it in terms of functions
-((I32 Any Any): type swap type == ) (: ( $c $a $b ) = b ) $tern fun
-((I32 Any Any): type swap type == && ) (: ( $c $a $b ) = a ) $tern fun
-\end{verbatim}
-
-\subsection{closures}
-Via the same syntax already defined
-
-\subsection{Recursion}
-Need to think about how best to do this in a formalization but there was a lot of time that had to go into making type inference work for recursive functions (implementation + 3 pages of notes).
-
-\subsection{Tuples}
-N-ary tuples represent one of the more problematic parts of the language, notably \texttt{unpack} being the only operator with a type signature dependant on the type of it's input, additionally I have no clue how to define an arrow type with a specified but non-constant number of return values.
-In order to avoid expanding the scope of this assignment we may replace the \texttt{unpack} operator with a set of two operators with similar functionality.
-
-\subsection{Unions}
-Unions should be an easy addition.
-
-\begin{verbatim}
-    (()(I32 F32 |): 12 ) $value =
-    value type I32 == 1 0 tern $value_is_int =
-\end{verbatim}
-
-\subsection{Identifiers}
-Although the system for identifiers is a bit different from most languages we've covered, it's not too strange.
-
-\subsection{Types}
-Although types, which are treated as untyped stack values, aren't pretty we could take the approach of having a second stack (for operations on types) and second store (for type variables), instead of the implementation's approach.
-Or we can have a type judgement in addition to a value judgement.
-
-\subsection{Operators, Keywords and Constants}
-So in summary our builtin operators would be: 
-\begin{itemize}
-    \item \texttt{+} - numeric add
-    \item \texttt{*} - numeric mul
-    \item \texttt{if-then-else} - ternary operator
-    \item \texttt{unpack} - destruct tuple (an attempt will be made)
-    \item \texttt{\~} - Extract value from an escaped identifier without invoking it (less important for safety but useful for defining additional operators using these as base)
-    \item \texttt{=} - the expression \texttt{e \$x = ...} is equivalent to \texttt{let x = e in ...}. But note that there are no 'bound variables' in this language.
-    \item \texttt{type} - get type of argument (equivalent to \texttt{type()} in python
-    \item \texttt{==} - current plan is to have an ambiguous == operator which accepts either 2 types or 2 equivalently typed values like in the language
-    \item \texttt{|} - used to form type unions
-    \item \texttt{Arrow} - closure types
-    \item \texttt{Unit} - keyword to distinguish from the empty tuple ()
-    \item \texttt{Void} - used due to lack of syntax for empty union
-\end{itemize}
-
-\subsection{Notably Not Included}
-
-\subsubsection{Classes}
-As much as we would like to include classes (notably recursive classes), it would likely add excess complexity, requiring subtyping (on N dimensions); untyped closures which can operate on types; and likely some other painpoints.
-And it's not clear how they could make the system unsafe as apart from being used to define recursive types are simply specialized versions of preexisting types.
-
-\subsubsection{Functions}
-Really the only reason I opted for the function overloading system instead of normal branching in the original language was in order to make it easier to extend the language.
-Function overloading seems like it would be needlessly painful to formalize and doesn't contribute to our proof of soundness any more than branching via a ternary.
-As the case for functions having potentially no possible branch results in a type error in the complete language and this is simplified by having an else clause.
-
 \section{Syntax}
@@
 \end{grammar}
-Notes
-\begin{itemize}
-    \item A program consists of a list of $op^*$ expressions
-    \item $\varepsilon$ is used to denote an empty program/expression
-    \item A syntactic fixed point combinator was chosen instead of $\kwrec{\tau^*}{\tau^*}{e^*}$. It's possible that one could be implemented in the language itself.
-    \item Tuples and Tuple types are enclosed in parens
-    \item $\kwunpack{e}$ and $\kwunpack{\tau}$ can give 0 or more outputs depending on the input tuple
-\end{itemize}
+Where $\varepsilon$ is used to denote an empty program/expression
+
+\subsection{Included}
+
+\subsection{Number Literals}
+Number literals are denoted with subscript indicating the numeric type they correspond with.
+This mirrors the actual language which uses the C-style syntax for number literal type inference.
+
+\subsubsection{Branching}
+As mentioned before a ternary operator would be a reasonable subset of branching via function overloading. The tern operator is defined as an operator which
+
+\begin{verbatim}
+# Tern takes an I32 and two values of the same type
+# and gives one of those two values
+$tern ~ :type # :type - (I32 $A $A) ($A) Arrow
+
+# As it is a subset we can define it in terms of functions
+((I32 Any Any): type swap type == ) (: ( $c $a $b ) = b ) $tern fun
+((I32 Any Any): type swap type == && ) (: ( $c $a $b ) = a ) $tern fun
+\end{verbatim}
+
+\subsubsection{Identifiers}
+Identifiers are included, however, it was found that the notation for 
+dereferencing an identifier $\kweid{x}~\texttt{~}$ was confusing and
+thus the behavior was changed and thus the default behavior was
+changed such that unescaped identifiers are dereferenced instead of 
+called and values must be explicitly called via the $\texttt{@}$
+operator.
+
+\subsubsection{Macros}
+Macros use a specific instance of the type annotations, notably where the
+input types are specified and the output types are inferred. In order to 
+enable recursion, a syntactic fixed-point combinator was provided to replace
+the $\texttt{rec}$ syntax.
+
+\subsubsection{Identical to base language}
+\begin{enumerate}
+    \item Tuples and tuple related operators
+    \item type operator
+    \item $\texttt{Arrow}$ type
+    \item some math and comparison operators
+\end{enumerate}
+
+\subsection{Not included}
+Things which were either out of scope, problematic, or uninteresting were unincluded. Here a re some of the notable ones.
+
+\subsubsection{Classes}
+As much as we would like to include classes (notably recursive classes), it 
+would likely add excess complexity, requiring subtyping (on N dimensions); 
+untyped closures which can operate on types; and likely some other painpoints.
+And it's not clear how they could make the system unsafe as apart from being
+used to define recursive types are simply specialized versions of preexisting
+types.
+
+\subsubsection{Functions}
+Really the only reason the function overloading system was preferred instead
+of normal branching in the original language was in order to make it easier 
+to extend the language. Function overloading seems like it would be needlessly
+painful to formalize and doesn't contribute to our proof of soundness any more
+than branching via a ternary. As the case for functions having potentially no
+possible branch results in a type error in the complete language and this is
+simplified by having an else clause.
+
+\subsubsection{Unions}
+Although the language currently supports some operations with union types,
+the system does not work at runtime when the true type is not known, thus,
+a flaw in the base language was discovered while working on the semantics
+for this subset and thus the original system needs to be redesigned. This
+would likely mean that the only way to make runtime union-types would be
+via the \texttt{make} operator which is related to \texttt{class} and we
+would additionally need to either make a different way to branch on unions
+or make the compiler able to determine if the condition implies a particular
+union member. Additionally Void was not included as on paper, it doesn't
+work with the type inference algorithm.
+
+\begin{verbatim}
+(()(I32 F32): ... ) $get_num =
+get_num $n =
+(: n type I32 == ) (: 
+    # How do we know that it's an i32 here?
+) $branch fun
+(: n type F32 == ) (: 
+    # How do we know that it's an f32 here?
+) $branch fun
+branch
+\end{verbatim}
+
+The above example program illustrates the problem with the language.
 
@@
             {\typed{\ctx}{\kwfix{e}{\tau_1}}{\tau_1}}
+
+        \and \Rule{end_scope type}{\strut}{\typed{\ctx}{\kw{end\_scope}}{\varepsilon}}
     \end{mathpar}
@@
     \item \Rulename{STEP}
-        Then $e^* = v^* e_1 ... e_n$ and $e^{*\prime} \step v^* e_1' ... e_n$ where $\etyped{e_1}{\tau}$ and $\etyped{e_1'}{\tau'}$.
+        Then $e^* = v^* e_1 ... e_n$ and $e^ \step v^* e_1' ... e_n$ where $\etyped{e_1}{\tau}$ and $\etyped{e_1'}{\tau'}$.
         By induction, $\tau \tstep^* \tau'$, so by \Rulename{TSTEP}, $\tau^* \tstep^* [\tau' / \tau]\tau^*$.
@@
 
-% \subsubsection{Lemma - Convergence}
-% Part 1: $\tau^* \tstep \tau^{*\prime}$ is no longer possible when $\tau_i \prim ~ \forall~ \tau_i \in \tau^*$.
-% \begin{proof}
-% Proof by exhaustion with all rules involving $\prim$ types not stepping. To values
-% \end{proof}
-% Part 2: $\tau^* \tstep \tau^{*\prime}$ will eventually step to a $\tau^{*\prime}$ such that $\tau_i \prim ~ \forall~ \tau_i \in \tau^{*\prime}$.
-% \begin{proof}
-% This is provable by the fact that the language does not feature recursive types and the only form of recursion in the language is via the fix operator and $\kwType{}$ operator does no have to follow recursive paths in order to get it's type (Type Rules \Rulename{Macro}, \Rulename{Fix}).
-The other rules are trivial (ie - step to a $\prim$) and/or irrelevant to the assertion
-% \end{proof}
-
-% \subsubsection{Proof}
-% \begin{proof}
-% \end{proof}
-
 \end{document}

### 2021-12-09 project/paper.tex
submitted

@@
 
-\title{Verifying a Postfix Language \\
-    \large CS595 Final Project}
+\title{Verifying a Postfix Language}
 \author{Dustin Van Tate Testa and Andrew Neth}
@@
 \maketitle
-\begin{abstract}
-A formalization and some proofs of soundness is given to a subset of \href{https://github.com/dvtate/postfix-haskell}{a functional, postfix language}.
-\end{abstract}
+% \begin{abstract}
+% A formalization and some proofs of soundness is given to a subset of \href{https://github.com/dvtate/postfix-haskell}{a functional, postfix language}.
+% \end{abstract}
 
@@
 \subsection{Motivation}
-WebAssembly is a typed, stack-machine based assembly language which is widely supported across all major browsers.
-A number of new languages have come about with the goal of providing better support for the platform.
-Additionally it has proven the utility of typed stack-machines as a reasonably safe way to write programs.
-
-There appears to be little research into stack-based, functional languages (likely due to these languages not being common), thus more exploration on the topic could likely produce some value.
-
-\href{https://github.com/dvtate/postfix-haskell}{The language this paper focuses on} is currently being used to power \href{https://ridder.xyz/game}{some insignificant demos}, but maybe in some distant future it will end up as the backend for a visual programming tool or some other application where guarantees of safety which could be produced here would be very important.
+\href{https://github.com/dvtate/postfix-haskell}{The functional, postfix language this paper focuses on} is currently being used to power \href{https://ridder.xyz/game}{some insignificant demos}, but maybe in some distant future it will end up as the backend for a visual programming tool or some other application where guarantees of safety like the ones here would be very important.
 
 \subsection{The Language}
-This section describes some of the key features of the original language. It is included only to show that the subset is a good representation of the original language and thus can be skipped.
+This section describes some of the key features of the original language.
+It is included only to show that the subset we chose is a good representation of the original language and thus can be skipped.
 
-\subsubsection{Syntax}
+\subsubsection{Full Syntax}
 \begin{itemize}
-    \item \textbf{Primitive Literals:} The language supports numeric and string literals although in the paper we will likely only cover numbers.
+    \item \textbf{Primitive Literals:} The language supports number and string literals
 
@@
 
-    In the language, identifiers can reference both typed and untyped values, but likely will be limited to typed values in the paper.
-
     \item \textbf{Functions:} function overloading is the only supported form of branching in the language. The syntax for the \texttt{fun} operator is as follows.
 
-    \texttt{(($..._1$)($..._2$ I32): Condition ) (: Action ) \$Identifier fun}.
-
-    In the final paper this will likely get replaced with a ternary.
+    \texttt{(($..._1$)($..._2$ I32): Condition ) (($..._1$): Action ) \$Identifier fun}.
 
     \item \textbf{Other Operators:} Although the majority of primitive operators are defined in the standard library, the compiler actually has a few builtins as well.
-Although these are very important to actually using the language they aren't worthy of listing out in their entirety here.
+Although these are very important to actually using the language, they aren't worthy of listing out in their entirety here.
 Some examples include: \texttt{require}, \texttt{use}, \texttt{import}, \texttt{export}, \texttt{asm}, \texttt{namespace}, \texttt{type}, \texttt{unpack}, \texttt{class}, \texttt{=}, \texttt{|}, \texttt{@}, \texttt{\~}, \texttt{==}, \texttt{make}
@@
 \subsection{Included}
+In addition to the following, tuples and tuple related operators,$\kwType{}$ operator, $\texttt{Arrow}$ type, and some math and comparison operators were included identically to their counterparts in base language.
 
-\subsection{Number Literals}
+\subsubsection{Number Literals}
 Number literals are denoted with subscript indicating the numeric type they correspond with.
@@
 \subsubsection{Branching}
-As mentioned before a ternary operator would be a reasonable subset of branching via function overloading. The tern operator is defined as an operator which
+As mentioned before a ternary operator would be a reasonable subset of branching via function overloading.
+The $\kwtern{}{}{}$ operator is defined as an operator which takes a condition and two values of the same type, and gives one depending on the condition.
+Which could be defined like so:
 
 \begin{verbatim}
-# Tern takes an I32 and two values of the same type
-# and gives one of those two values
-$tern ~ :type # :type - (I32 $A $A) ($A) Arrow
-
-# As it is a subset we can define it in terms of functions
-((I32 Any Any): type swap type == ) (: ( $c $a $b ) = b ) $tern fun
-((I32 Any Any): type swap type == && ) (: ( $c $a $b ) = a ) $tern fun
+((I32 Any Any): type swap type == ) (: ( $c $a $b ) = b ) $select fun
+((I32 Any Any): type swap type == && ) (: ( $c $a $b ) = a ) $select fun
 \end{verbatim}
@@
 dereferencing an identifier $\kweid{x}~\texttt{~}$ was confusing and
-thus the behavior was changed and thus the default behavior was
-changed such that unescaped identifiers are dereferenced instead of 
+thus the subset was changed such that unescaped identifiers are dereferenced instead of 
 called and values must be explicitly called via the $\texttt{@}$
@@
 \subsubsection{Macros}
-Macros use a specific instance of the type annotations, notably where the
+Macros use a specific variation of the type annotations, notably where the
 input types are specified and the output types are inferred. In order to 
@@
 the $\texttt{rec}$ syntax.
-
-\subsubsection{Identical to base language}
-\begin{enumerate}
-    \item Tuples and tuple related operators
-    \item type operator
-    \item $\texttt{Arrow}$ type
-    \item some math and comparison operators
-\end{enumerate}
 
@@
 \begin{verbatim}
-(()(I32 F32): ... ) $get_num =
-get_num $n =
-(: n type I32 == ) (: 
+(: type I32 == ) (: 
     # How do we know that it's an i32 here?
 ) $branch fun
-(: n type F32 == ) (: 
-    # How do we know that it's an f32 here?
-) $branch fun
-branch
+(()(I32 F32): ... ) @ branch
 \end{verbatim}
-
-The above example program illustrates the problem with the language.
 
@@
 To annotate stack values we used the notation of two values being written next to each other, as they would be written in the language itself.
-The values at the bottom of the stack (furthest to the left) are often irrelevant, and thus can be omitted with this notation.
+The values at the bottom of the stack (furthest to the left) are often irrelevant, and thus for most rules can be omitted with this notation.
 The superscript * notation (ie - $e^*$) is used to annotate 0 or more of the preceding.
-This is explicitly defined in the syntax but implicitly expanded in the semantics.
 
@@
         % Derivation of e* -> e*'
-        \Rule{STEP}
+        \Rule{Step}
             {e_1 \step e_1'
@@
         % number literals
-        \Rule{V-1}
+        \Rule{V-Num}
              {\forall ~ \tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
@@
         % Tuple literals evaluation
-        \Rule{TUPLES-1}
+        \Rule{S-Tuple}
             {e^* \step e^{*\prime}}
             {( e^* ) \step ( e^{*\prime} )}
-        \Rule{TUPLES-2}
+        \Rule{V-Tuple}
             {v_i \val ~ \forall v_i \in v^*}
@@
 \subsection{Macros, Identifiers and Scoping}
-The store, $\sigma$, consists of a list of \textit{scopes}, where each scope is a mapping from identifiers to values.
+The store, $\sigma$, consists of a list of scopes, where each scope is a mapping from identifiers to values.
 Notice the addition of an auxiliary $\kw{end\_scope}$ operator which removes the current scope.
-A macro literal steps to a macro value by first stepping its input types until they're all values before substituting all of the values in the store within its body (forming a closure).
-Calling a macro value pushes a new empty mappings list onto the store, then places the macro's expressions and an $\kw{end\_scope}$ operator onto the stack, enabling local identifiers.
+A macro literal steps to a macro value by first stepping its input types until they're all values before substituting all of the values in the store within its body (forming a closure) (this rule comes later).
+Calling a macro value pushes a new empty mappings list onto the store, then places the macro's body and an $\kw{end\_scope}$ operator onto the stack, enabling locals.
 {
@@
         % Closures capture lexically scoped variables via substitution 
-        \Rule{LET}
+        \Rule{S-Let}
             {v \val
@@
 
-        \and \Rule{Identifier}
+        \and \Rule{S-Ident}
             { \kwid{x} \step v \in \sigma(-1)}
@@
 
-        \and \Rule{CALL-1}
+        \and \Rule{S-Call}
             {\sigma' = \sigma \cup [~]
@@
 
-        \and \Rule{CALL-2}
+        \and \Rule{S-Fix}
+            %{v = \kwfix{\kwmacro{\tau^*~ \tau_1}{e^*}}{\tau_1} \\ v \val}
+            {f = \kwfix{v}{\tau} \\ f \val}
+            %{v~ \textbf{@} \step v~ \kwmacro{\tau^*~ \tau_1}{e^*}~ \textbf{@}}
+            {f~ \textbf{@} \step f~ v~ \textbf{@}}
+
+        \and \Rule{S-End}
             {\sigma = [s_1,...s_n,s_{n+1}] \\ \sigma' = [s_1,...s_n]}
@@
 \subsection{Others}
-Notice the use of the \href{https://en.wikipedia.org/wiki/Kronecker_delta}{Kronecker Delta} to define the equals operator.
+Notice the use of the \href{https://en.wikipedia.org/wiki/Kronecker_delta}{Kronecker Delta} to define the equals operator. Here $\tau$ indicates the rule is defined for all numeric types $\tau$
 {
@@
         % unpack operator
-        \Rule{TUPLES-3}
+        \Rule{S-Unpack}
             {v_i \val ~\forall~ v_i \in v^*}
@@
         % math
-        \Rule{ADD}
-            {v_1 \val \\ v_2 \val \\ v_3 = v_1 + v_2}
-            {v_1~ v_2~ + \step v_3}
+        \Rule{S-Add}
+            {\tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
+            {\kwnumlit{n_1}{\tau}~ \kwnumlit{n_2}{\tau}~ +~ \step \kwnumlit{(n_1 + n_2)}{\tau}}
         \and
-        \Rule{MUL}
-            {v_1 \val \\ v_2 \val \\ v_3 = v_1 * v_2}
-            {v_1~ v_2~ \times \step v_3}
+        \Rule{S-Mul}
+            {\tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
+            {\kwnumlit{n_1}{\tau}~ \kwnumlit{n_2}{\tau}~ \times~ \step \kwnumlit{(n_1 \times n_2)}{\tau}}
         \and
@@
         % Comparisons
-        \Rule{EQUIV}
-            {v_1 \val \\ v_2 \val}
-            {v_1~ v_2~ ==~ \step \kwnumlit{(\delta_{v_1 v_2})}{\kwInt}}
+        \Rule{S-Eq}
+            {\tau \in \{ \kwInt, \kwLong, \kwFloat, \kwDouble\}}
+            {\kwnumlit{n_1}{\tau}~ \kwnumlit{n_2}{\tau}~ ==~ \step \kwnumlit{(\delta_{n_1 n_2})}{\kwInt}}
 
         % Ternary
-        \Rule{TERN-1}
+        \Rule{S-Tern-0}
             {v_1 \val \\ v_2 \val}
             {\kwtern{\kwnumlit{0}{\kwInt}}{v_1}{v_2} \step v_2}
-        \Rule{TERN-2}
-            {v_1 \val \\ v_2 \val \\ n \neq 0}
-            {\kwtern{\kwnumlit{n}{\kwInt}}{v_1}{v_2} \step v_1}
+        \Rule{S-Tern-1}
+            {v_1 \val \\ v_2 \val \\ n \neq \kwnumlit{0}{\kwInt}}            {\kwtern{\kwnumlit{n}{\kwInt}}{v_1}{v_2} \step v_1}
         \and
@@
     \begin{mathpar}
-        \Rule{P-I32}{\strut}{\kwInt \prim} \and
-        \Rule{P-I64}{\strut}{\kwLong \prim} \and
-        \Rule{P-F32}{\strut}{\kwFloat \prim} \and
-        \Rule{P-F64}{\strut}{\kwDouble \prim} \and
+        \Rule{P-Num}{\tau \in {\kwInt, \kwLong, \kwFloat, \kwDouble}}{\tau \prim} \and
         %\Rule{P-sum}{\tau_1 \prim \\ \tau_2 \prim}{\tau_1 \tau_2 \texttt{|} \prim} \and
@@
         % Derivation of T* -> T*'
-        \Rule{TSTEP}
+        \Rule{Ty-Step}
             {\tau \tstep \tau'
@@
 
-        \and \Rule{TSTEP - Mixed}
+        \and \Rule{Ty-Step-Mixed}
             {\tau^* \tstep \tau^{*\prime}
@@
             {v^*~ \tau^*~ e^*~ \tstep v^*~ \tau^{*\prime}~ e^*}
-        \and \Rule{STEP - Mixed}
+        \and \Rule{Val-Step-Mixed}
             {e^* \step e^{*\prime}
@@
     \begin{mathpar}
-        \Rule{Type-Let}
+        \Rule{Ty-Let}
             {\tau \prim
@@
 
-        \and \Rule{MACRO-2}
+        \and \Rule{Macro-Subst}
             {\sigma = [ ..., [x_1 \step v_1 ~ ... ~ x_n \step v_n, 
@@
 
-        \and \Rule{Type-Id}
+        \and \Rule{Ty-Ident}
             { \kwid{T} \step \tau \in \sigma(-1)}
@@
         % Tuple type literals evaluation
-        \Rule{Tuple Types - 1}
+        \Rule{Ty-Tuple}
             {\tau^* \tstep \tau^{*\prime}}
             {( \tau^* ) \tstep ( \tau^{*\prime} )}
-        \Rule{Tuple Types - 2}
+        \Rule{P-Tuple}
             {\tau_i \prim ~ \forall \tau_i \in \tau^*}
@@
         % unpack operator
-        \Rule{Tuple Types - 3}
+        \Rule{Ty-Unpack}
             {\tau_i \prim ~ \forall~ \tau_i \in \tau^*}
@@
     \begin{mathpar}
-        \Rule{Type Equiv}
+        \Rule{S-TyEq}
             {\tau_1 \prim \\ \tau_2 \prim}
-            {\tau_1~ \tau_2~ ==~ \tstep~ \kwnumlit{(\delta_{\tau_1 \tau_2})}{\kwInt}}
+            {\tau_1~ \tau_2~ ==~ \step~ \kwnumlit{(\delta_{\tau_1 \tau_2})}{\kwInt}}
 
         % fixed point combinator
-        \and \Rule{Fixed Point}
-            {\kwmacro{\tau^*~ \tau_1}{e^*} \val \\ \tau_1 \prim}
-            {\kwfix{\kwmacro{\tau^*~ \tau_1}{e^*}}{\tau_1} \val}
-        \and \Rule{Call FP}
-            {v = \kwfix{\kwmacro{\tau^*~ \tau_1}{e^*}}{\tau_1} \\ v \val}
-            {v~ \textbf{@} \step v~ \kwmacro{\tau^*~ \tau_1}{e^*}~ \textbf{@}}
+        \and \Rule{V-Fix}
+            {e \val \\ \tau_1 \prim}
+            {\kwfix{e}{\tau_1} \val}
 
         % input types for macros can step
-        \and \Rule{Macro Type Step}
+        \and \Rule{Ty-Macro}
             {\tau^* \tstep \tau^{*\prime}}
@@
         % typeof operator
-        \Rule{TYPE-1}{v \val \\ \etyped{v}{\tau}}{\kwType{v} \tstep \tau} 
+        \Rule{Ty-Typeof}{v \val \\ \etyped{v}{\tau}}{\kwType{v} \tstep \tau} 
     \end{mathpar}
@@
     \begin{mathpar}
-    \Rule{T-0}
+    \Rule{T-Step}
         {\typed{\ctx}{e}{\tau} \\ \tau \tstep \tau'}
@@
     \begin{mathpar}
-        \Rule{T-1}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwInt}}{\kwInt}}
-        \and \Rule{T-2}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwLong}}{\kwLong}}
-        \and \Rule{T-3}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwFloat}}{\kwFloat}}
-        \and \Rule{T-4}{\strut}{\typed{\ctx}{\kwnumlit{n}{\kwDouble}}{\kwDouble}}
-        \and \Rule{T-5}
+        \Rule{T-Num}{\tau \in {\kwInt, \kwLong, \kwFloat, \kwDouble}}{\typed{\ctx}{\kwnumlit{n}{\tau}}{\tau}}
+        \and \Rule{T-Add}
             {\typed{\ctx}{e_1}{\tau} 
@@
             {\typed{\ctx}{e_1~ e_2~ +}{\tau}}
-        \and \Rule{T-6}
+        \and \Rule{T-Mul}
             {\typed{\ctx}{e_1}{\tau} 
@@
             {\typed{\ctx}{e_1~ e_2~ \times}{\tau}}
-        \and \Rule{T-7}
+        \and \Rule{T-Tern}
             {\typed{\ctx}{e_1}{\kwInt}
@@
             {\typed{\ctx}{\kwtern{e_1}{e_2}{e_3}}{\tau}}
-        \and \Rule{T-8}
+        \and \Rule{T-Eq}
             {\typed{\ctx}{e_1}{\tau}
@@
             {\typed{\ctx}{e_1~ e_2~ ==~}{\kwInt}}
-        \and \Rule{T-9}
+        \and \Rule{T-TyEq}
             {\tau_1 \prim
@@
     \begin{mathpar}
-        \Rule{Empty Tuple}{\strut}{\typed{\ctx}{()}{\kwUnit}}
-        \and \Rule{Tuple Type}
+        \Rule{T-Unit}{\strut}{\typed{\ctx}{()}{\kwUnit}}
+        \and \Rule{T-Tuple}
             {\tau^* = \{ \tau_i ~\text{such that}~ \typed{\ctx}{e_i}{\tau_i}~ \forall ~ e_i \in e^* \} }
             {\typed{\ctx}{(e^*)}{(\tau^*)}}
-        \and \Rule{Unpack}
+        \and \Rule{T-Unpack}
             {\typed{\ctx}{(e^*)}{(\tau^*)}}
@@
         % Move this to type rules   
-        \Rule{Macro}
+        \Rule{T-Macro}
             {\tau^* = \tau_1 ~...~ \tau_n \\
@@
 
-        \and \Rule{Call}
-            {\typed{\ctx}{e_1}{\kwArrow{\tau^*}{\tau^{*\prime}}}
+        \and \Rule{T-Call}
+            {\typed{\ctx}{e_1}{\kwArrow{(\tau^*)}{(\tau^{*\prime})}}
                 \\ \typed{\ctx}{e^*}{\tau^*}}
@@
 
-        \and \Rule{Identifier}
+        \and \Rule{T-Ident}
             {\ctx(\kwid{x}) = \tau}
@@
 
-        \and \Rule{Let}
+        \and \Rule{T-Let}
             {\typed{\ctx}{e}{\tau}
@@
 
-        \and \Rule{Fix}
+        \and \Rule{T-Fix}
             {\tau_1 \prim
@@
 
-        \and \Rule{end_scope type}{\strut}{\typed{\ctx}{\kw{end\_scope}}{\varepsilon}}
+        \and \Rule{T-End}{\strut}{\typed{\ctx}{\kw{end\_scope}}{\varepsilon}}
     \end{mathpar}
@@
     \begin{itemize}
-        \item (T-\{1,2,3,4\})
+        \item \Rname{T-Num}
             Then $e^* = \kwnumlit{n}{\tau}$, where $\tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}$.
-            By (V-1), $\kwnumlit{n}{\tau} \val$, so $e^* \val$.
+            By \Rname{V-Num}, $\kwnumlit{n}{\tau} \val$, so $e^* \val$.
 
-        \item (T-5)
+        \item \Rname{T-Add}
             Then $e^* = e_1~ e_2~ +$ and $\etyped{e_1}{\tau}$ and $\etyped{e_2}{\tau}$ for some $\tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}$.
@@
             By induction, $e_1 \step e_1'$ or $e_1 \val$.
-            If $e_1 \step e_1'$, then by \Rulename{STEP}, $e^* \step e_1'~ e_2~ +$.
+            If $e_1 \step e_1'$, then by \Rname{Step}, $e^* \step e_1'~ e_2~ +$.
 
             If $e_1 \val$, then by induction, $e_2 \step e_2'$ or $e_2 \val$.
-            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP}, $e^* \step e_1~ e_2'~ +$.
+            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rname{Step}, $e^* \step e_1~ e_2'~ +$.
 
-            If $e_1 \val$ and $e_2 \val$, then by CF, $e_1 = \kwnumlit{n_1}{\tau}, e_2 = \kwnumlit{n_2}{\tau}$, and by \Rulename{ADD}, $e^* \step \kwnumlit{n_1 + n_2}{\tau}$.
+            If $e_1 \val$ and $e_2 \val$, then by CF, $e_1 = \kwnumlit{n_1}{\tau}, e_2 = \kwnumlit{n_2}{\tau}$, and by \Rname{S-Add}, $e^* \step \kwnumlit{n_1 + n_2}{\tau}$.
 
-        \item (T-6)
-            $[\times / +][\Rulename{MUL} / \Rulename{ADD}]$ (T-5)
+        \item \Rname{T-Mul}
+            $[\times / +][\Rname{S-Mul} / \Rname{S-Add}]$ \Rname{T-Add}
 
-        \item (T-7)
+        \item \Rname{T-Tern}
             Then $e^* = \kwtern{e_1}{e_2}{e_3}$ and $\etyped{e_1}{\kwInt}$. %and $\etyped{e_2}{\tau}$ and $\etyped{e_3}{\tau}$ for some $\tau$.
@@
             By induction, $e_1 \step e_1'$ or $e_1 \val$.
-            If $e_1 \step e_1'$, then by \Rulename{STEP}, $e^* \step \kwtern{e_1'}{e_2}{e_3}$.
+            If $e_1 \step e_1'$, then by \Rname{Step}, $e^* \step \kwtern{e_1'}{e_2}{e_3}$.
 
             If $e_1 \val$, then by induction, $e_2 \step e_2'$ or $e_2 \val$.
-            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rulename{STEP}, $e^* \step \kwtern{e_1}{e_2'}{e_3}$.
+            If $e_1 \val$ and $e_2 \step e_2'$, then by \Rname{Step}, $e^* \step \kwtern{e_1}{e_2'}{e_3}$.
 
             If $e_1 \val$ and $e_2 \val$, then by induction, $e_3 \step e_3'$ or $e_3 \val$.
-            If $e_1 \val$ and $e_2 \val$ and $e_3 \step e_3'$, then by \Rulename{STEP}, $e \step \kwtern{e_1}{e_2}{e_3'}$.
+            If $e_1 \val$ and $e_2 \val$ and $e_3 \step e_3'$, then by \Rname{Step}, $e \step \kwtern{e_1}{e_2}{e_3'}$.
 
@@
 
-            If $n = 0$, then by \Rulename{TERN-1}, $e^* \step e_3$.
-            Otherwise, $n \neq 0$, so by \Rulename{TERN-2}, $e^* \step e_2$.
+            If $n = 0$, then by \Rname{S-Tern-0}, $e^* \step e_3$.
+            Otherwise, $n \neq 0$, so by \Rname{S-Tern-1}, $e^* \step e_2$.
 
-        \item \Rulename{Empty Tuple}
-            Then $e^* = ()$. By \Rulename{TUPLES-2}, $() \val$.
+        \item \Rname{T-Unit}
+            Then $e^* = ()$. All zero of the tuple's elements are values, so by \Rname{V-Tuples}, $() \val$.
 
-        \item \Rulename{Tuple Type}
+        \item \Rname{T-Tuple}
             Then $e^* = (e^*_t) = (e_1~ e_2~ ...~ e_n)$.
@@
 
-            If all elements are values, then $(e^*_t) \val$ by \Rulename{TUPLES-2}.
+            If all elements are values, then $(e^*_t) \val$ by \Rname{V-Tuple}.
 
-            Otherwise, one or more elements can step, including a leftmost steppable element $e_j \step e_j'$.
-            Then, by \Rulename{STEP}, $e^*_t \step [e_j' / e_j] e^*_t$, so by \Rulename{Tuples-1}, $e^* \step ([e_j' / e_j] e^*_t)$.
+            Otherwise, one or more elements can step, including a ``leftmost steppable element'' $e_j \step e_j'$.
+            Then, by \Rname{Step}, $e^*_t \step [e_j' / e_j] e^*_t$, so by \Rname{S-Tuple}, $e^* \step ([e_j' / e_j] e^*_t)$.
 
-        \item \Rulename{Unpack}
+        \item \Rname{T-Unpack}
             Then $e^* = \kwunpack{(e^*_t)}$.
@@
 
-            If $(e^{*'}_t) \step (e^{*'}_t)$, then by \Rulename{STEP}, $\kwunpack{(e^*_t)} \step \kwunpack{(e^{*'}_t)}$.
+            If $(e^{*'}_t) \step (e^{*'}_t)$, then by \Rname{Step}, $\kwunpack{(e^*_t)} \step \kwunpack{(e^{*'}_t)}$.
 
-            If $(e^*_t) \val$, then by \Rulename{TUPLES-3}, $\kwunpack{(e^*_t)} \step e^*_t$.
+            If $(e^*_t) \val$, then by \Rname{S-Unpack}, $\kwunpack{(e^*_t)} \step e^*_t$.
 
-        \item \Rulename{Macro}
-            Then $e^* = m = \kwmacro{\tau^*}{e^*_t}$. $m \val$ by the premise.
+        \item \Rname{T-Macro}
+            Then $e^* = m = \kwmacro{\tau^*}{e^*_t}$.
+$m \val$ by the premise.
+A macro isn't well-typed if it's not a value, and it's not a value if it hasn't had its free variables bound using \Rname{Macro-Subst}.
 
-        \item \Rulename{Call}
+        \item \Rname{T-Call}
             Then $e^* = e^*_x~ e_f~ \textbf{@}$. By induction, either $e_i \val~ \forall~ e_i \in e^*_x$ or $e^*_x \step e^{*'}_x$.
 
-            If $e^*_x \step e^{*'}_x$, then by \Rulename{STEP}, $e^* \step e^{*'}_x~ e_f~ \textbf{@}$.
+            If $e^*_x \step e^{*'}_x$, then by \Rname{Step}, $e^* \step e^{*'}_x~ e_f~ \textbf{@}$.
 
@@
 
-            If $e_f \step e_f'$, then by \Rulename{STEP}, $e^* \step e^*_x~ e_f'~ \textbf{@}$.
+            If $e_f \step e_f'$, then by \Rname{Step}, $e^* \step e^*_x~ e_f'~ \textbf{@}$.
 
-            Otherwise, $e_f \val$. To handwave a bit, by inversion on \Rulename{Macro}, $e_f$ is a macro with a body $e^*_b$, and by \Rulename{CALL-1}, $e^* \step e^*_b ~ \textbf{end_scope}$.
+            Otherwise, $e_f \val$. By CF, $e_f$ is either a macro or an application of $\kwfix{}{}$.
+            If it's a macro with a body $e^*_b$, then by \Rname{S-Call}, $e^* \step e^*_b ~ \textbf{end_scope}$.
+            If it's an application of $\kwfix{}{}$ to some $f$, then by \Rname{S-Fix}, $e^* \step e^*_x~ e_f~ f~ \textbf{@}$.
 
-        \item \Rulename{Identifier}
-            Then $e^* = \texttt{x}$.
-            By, uh, the other \Rulename{Identifier}, $e^*$ steps to a value in the store.
+        \item \Rname{T-Ident}
+            Then $e^* = \texttt{x}$ and some binding $\texttt{x} \step v$ is in scope.
+            By \Rname{S-Ident}, $e^* \step v$.
 
-        \item \Rulename{Let}
-            Then $e^* = e ~ \texttt{x} =$.
+        \item \Rname{T-Let}
+            Then $e^* = e ~ \kweid{x} =$.
             By induction, $e \val$ or $e \step e'$.
-            If $e \step e'$, then by \Rulename{STEP}, $e^* \step e' ~ \texttt{x} =$.
-            If $e \val$, then by the other \Rulename{LET}, $\sigma;e^* \step \sigma';\varepsilon$, where $\sigma'$ is, as described in the rule, $\sigma$ with $x \step e$ added to its ``topmost'' scope.
+            If $e \step e'$, then by \Rname{Step}, $e^* \step e' ~ \kweid{x} =$.
+            If $e \val$, then by \Rname{S-Let}, $\sigma;e^* \step \sigma';\varepsilon$, where $\sigma'$ is, as described in the rule, $\sigma$ with $\texttt{x} \step e$ added to its ``topmost'' scope.
 
-        \item \Rulename{Fix}
-            Then $e^* = \kwfix{e}{\tau_1}$.
+        \item \Rname{T-Fix}
+            Then $e^* = \kwfix{e}{\tau_1}$ where $\etyped{e}{\kwArrow{(\tau_{in}^* ~ \tau_1)}{\tau_{out}}}$ and $\tau_1 \prim$.
             By induction, $e \val$ or $e \step e'$.
-            If $e \step e'$, then by \Rulename{STEP}, $e^* \step \kwfix{e'}{\tau_1}$.
-            If $e \val$, then I think I'm missing something here, but by \Rulename{Fixed Point}, $e^* \val$.
+            If $e \step e'$, then by \Rname{Step}, $e^* \step \kwfix{e'}{\tau_1}$.
+            If $e \val$, then by \Rname{V-Fix}, $e^* \val$.
+
+        \item \Rname{T-End}
+            Lemma: The presence of an \texttt{end_scope} operator on the stack to execute implies the presence of a scope in the store to pop. \\
+            Proof: \texttt{end_scope} is absent from the formal syntax, so it can only exist if it is introduced during execution.
+It is introduced exclusively via \Rname{S-Call}, which pushes exactly one scope, and eliminated exclusively via \Rname{S-End}, which pops exactly one scope.
+
+            Then $e^* = \texttt{end_scope}$, and by the lemma, there is a scope to pop. Then by \Rname{S-End}, $e^* \step \varepsilon$.
     \end{itemize}
+\end{proof}
+
+\subsection{Lemma - Primitive Values}
+If $v \val$ and $\typed{v}{\tau}$, then $\tau \prim$.
+\begin{proof}
+By induction on the derivation of $v \val$.
+\begin{itemize}
+    \item \Rname{V-Num}
+        Then $v = \kwnumlit{n}{\tau}$ and $\tau \in {\kwInt, \kwLong, \kwFloat, \kwDouble}$. By \Rname{P-Num}, $\tau \prim$.
+
+    \item \Rname{V-Tuple}
+        Then $v = (v^*)$ and $\tau = (\tau_i^*)$ where $\etyped{v_i}{\tau_i}$ and $v_i \val~ \forall v_i \in v^*$.
+        By induction, $\tau_i \prim~ \forall \tau_i \in \tau^*$.
+        So by \Rname{P-Tuple}, $\tau \prim$.
+
+    \item \Rname{Macro-Subst}
+        Then $v$ is a macro with its captures substituted in.
+        By \Rname{T-Macro}, $\tau$ is of the form $\kwArrow{\tau_1}{\tau_2}$.
+        By \Rname{P-Arrow}, $\tau \prim$.
+
+    \item \Rname{V-Fix}
+        Then $v = \kwfix{m}{\tau_1}$ where $t_1 = \kwArrow{\tau_3^*}{\tau_4^*}$ and $\etyped{m}{\kwmacro{t_3^* ~ t_1}{t_4^*}}$.
+        Then $\etyped{v}{\tau_1}$ by \Rname{T-Fix}, meaning $\tau = \tau_1$, so by \Rname{P-Arrow}, $\tau \prim$.
+\end{itemize}
+\end{proof}
+
+\subsubsection{Lemma - Convergence}\footnote{a.k.a. Progress for Types, kinda}
+Part 1: $\tau^* \tstep \tau^{*\prime}$ is no longer possible when $\tau_i \prim ~ \forall~ \tau_i \in \tau^*$.
+\begin{proof}
+By induction on the derivation of $\tau \prim$.
+
+\begin{itemize}
+    \item \Rname{P-Num}
+        Then $\tau \in {\kwInt, \kwLong, \kwFloat, \kwDouble}$.
+        By exhaustion, there is no rule of the form $\tau \tstep \tau'$ that matches.
+
+    \item \Rname{P-Arrow}
+        Then $\tau = \kwArrow{\tau_1}{\tau_2}$ and $\tau_1 \prim$ and $\tau_2 \prim$.
+        By \Rname{Ty-Step}, $\tau$ can step if $\tau_1$ can step or if $\tau_1 \prim$ and $\tau_2$ can step, but by induction, neither $\tau_1$ nor $\tau_2$ can step.
+        By exhaustion, there is no other rule that would allow $\tau$ to step.
+
+    \item \Rname{P-Tuple}
+        Then $\tau = (\tau^*)$ where $\tau_i \prim ~ \forall \tau_i \in \tau^*$.
+        By \Rname{Ty-Tuple}, a tuple type can step if its elements collectively can.
+        By induction, they cannot.
+        By exhaustion, there is no other rule that would allow $\tau$ to step.
+\end{itemize}
+\end{proof}
+Part 2: $\tau^* \tstep \tau^{*\prime}$ will eventually step to a $\tau^{*\prime}$ such that $\tau_i \prim ~ \forall~ \tau_i \in \tau^{*\prime}$.
+\begin{proof}
+This is provable by the fact that the language does not feature recursive types and the only form of recursion in the language is via the fix operator and $\kwType{}$ operator does not have to follow recursive paths in order to get its type (Type Rules \Rname{T-Macro}, \Rname{T-Fix}).
+The other rules are trivial (i.e., step to a $\prim$) and/or irrelevant to the assertion.
 \end{proof}
@@
 \begin{itemize}
-    \item \Rulename{STEP}
-        Then $e^* = v^* e_1 ... e_n$ and $e^ \step v^* e_1' ... e_n$ where $\etyped{e_1}{\tau}$ and $\etyped{e_1'}{\tau'}$.
-        By induction, $\tau \tstep^* \tau'$, so by \Rulename{TSTEP}, $\tau^* \tstep^* [\tau' / \tau]\tau^*$.
+    \item \Rname{Step}
+        Then $e^* = v^* e_1 ... e_n$ and $e^* \step v^* e_1' ... e_n$ where $\etyped{e_1}{\tau_1}$ and $\etyped{e_1'}{\tau_1'}$ and $v_i \val~ \forall v_i \in v^*$.
+        By induction, $\tau_1 \tstep^* \tau_1'$.
+        By the Primitive Values lemma, the types of those values on the left are all primitive, so by \Rname{Ty-Step}, $\tau^* \tstep^* [\tau_1' / \tau_1]\tau^* = \tau^{*\prime}$.
 
-    \item \Rulename{TUPLES-1}
+    \item \Rname{S-Tuple}
         Then $e^* = (e_{elem}^*)$ and $e^* \step (e_{elem}^{*\prime})$ and $\tau^* = (\tau_{elem}^*)$.
         By induction, $\etyped{e_{elem}^{*\prime}}{\tau_{elem}^{*\prime}}$ where $\tau_{elem}^* \tstep^* \tau_{elem}^{*\prime}$.
-By \Rulename{Tuple Type}, $\etyped{(e_{elem}^{*\prime})}{(\tau_{elem}^{*'})}$, and by \Rulename{Tuple Types - 1}, $(\tau_{elem}^*) \tstep (\tau_{elem}^{*'})$.
+By \Rname{T-Tuple}, $\etyped{(e_{elem}^{*\prime})}{(\tau_{elem}^{*'})}$, and by \Rname{Ty-Tuple}, $(\tau_{elem}^*) \tstep (\tau_{elem}^{*'})$.
 
-    \item \Rulename{LET}
-        Then $e^* = v ~ \texttt{\$x} ~ =$.
-By the other \Rulename{Let}, $\etyped{e^*}{\varepsilon}$, so $\tau^* = \varepsilon$.
-$e^* \step \varepsilon$, so $\tau^{*\prime} = \varepsilon$, so $\tau^* = \tau^{*\prime}$.
+    \item \Rname{S-Let}
+        Then $e^* = v ~ \texttt{\$x} ~ =$. By \Rname{T-Let}, $\etyped{e^*}{\varepsilon}$, so $\tau^* = \varepsilon$. $e^* \step \varepsilon$, so $\tau^* = \tau^{*\prime} = \varepsilon$.
 
-    \item \Rulename{Identifier}
-        Then $e^* = \texttt{x}$, where $\texttt{x} \step v$ is in scope.
-By the other \Rulename{Identifier}, the type of an identifier expression is the same as the type of the value in the store ($\etyped{v}{\tau}$), so $\tau^* = \tau^{*\prime} = \tau$.
+    \item \Rname{S-Ident}
+        Then $e^* = \texttt{x}$ and some binding $\texttt{x} \step v$ is in scope where $\etyped{v}{\tau_v}$.
+$e^* \step v$ and, by \Rname{T-Ident}, $\etyped{e^*}{\tau_v}$, so $\tau^* = \tau^{*\prime} = \tau_v$.
 
-    \item \Rulename{CALL-1}
-        Then $e^* = $.
-        % I think we need to modify (CALL-1) to require the values? Revisit.
+    \item \Rname{S-Call}
+        Then $e^* = m ~\textbf{@}$, where $m = \kwmacro{\tau_{in}^*}{e^*}$ and $m \val$.
 
-    \item \Rulename{CALL-2}
-        % do we need a type rule for end_scope?
+        This is going to get a little weird.
 
-    \item \Rulename{TUPLES-3}
+        First, \Rname{Macro-Subst} means that $m \val$ iff all its free variables have been replaced with captures.
+
+        Next, \Rname{T-Macro} states that once that's done, a macro is well-typed only if it's ``ready to be called'' (appropriate arguments are on the stack), because its return type is ``whatever you get when you call it'', i.e., we're doing some kind of type inference because we tried too hard.
+        $e^*$ being well-typed is part of the theorem's premise, so the arguments are indeed in place, and we can call them $v^*$ (with types $\tau_{in}^*$, of course).
+
+        To get the return type, we start with the $\kwType{(v^* ~ m ~ \textbf{@})}$ described in \Rname{T-Macro}.
+        By the Convergence lemma, we can step this type-expression to a primitive, and by \Rname{Ty-Typeof} and \Rname{T-Tuple}, the result will be a tuple type, which we can name $(\tau_{out}^*)$.
+
+        That then gives us $\etyped{m}{\kwArrow{(\tau_{in}^*)}{(\tau)}}$.
+        As previously established, the necessary arguments $\etyped{v^*}{\tau_{in}^*}$ are on the stack,
+        so by \Rname{T-Call}, $\etyped{e^*}{\tau_{out}^*}$.
+
+        $e^*$, a macro call, steps to the macro's expansion (i.e., its body plus an \texttt{end_scope}).
+The ``type signature'' of the macro's expansion steps to $\tau_{out}^*$.
+The type of $e^*$ also steps to $\tau_{out}^*$.
+That's preservation.
+
+    \item \Rname{S-Fix}
+        Then $e^* = f~ \textbf{@}$ where $f = \kwfix{v}{\tau}$ and $e^* \step e^{*\prime} f~ v~ \textbf{@}$.
+
+        By \Rname{T-Call}, the type of $e^{*\prime}$ is inferred\footnote{I'm tired.} based on the expansion of $v$.
+        By \Rname{T-Fix}, this is also the case for $e^*$.
+
+    \item \Rname{S-End}
+        Then $e^* = \texttt{end_scope}$. By \Rname{T-End}, $\etyped{e^*}{\varepsilon}$, so $\tau^* = \varepsilon$. $e^* \step \varepsilon$, so $\tau^* = \tau^{*\prime} = \varepsilon$.
+
+    \item \Rname{S-Unpack}
         Then $e^* = \kwunpack{(v^*)}$, where $\etyped{v^*}{\tau_{elem}^*}$.
-        By \Rulename{Unpack}, $\tau^* = \tau_{elem}^*$. $e^* \step v^*$, so $\tau^{*\prime} = \tau^* = \tau_{elem}^*$.
+        By \Rname{T-Unpack}, $\tau^* = \tau_{elem}^*$. $e^* \step v^*$, so $\tau^{*\prime} = \tau^* = \tau_{elem}^*$.
 
-    \item \Rulename{ADD} \footnote{For these three rules, terms of the form $v_n$ are better expressed as numeric literals.}
-        Then $e^* = v_1~ v_2~ +$ and $e^* \step v_3$, where $\etyped{v_1}{\tau}$ and $\etyped{v_2}{\tau}$ for some $\tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}$.
-        By \Rulename{T-5}, $\etyped{v_1~ v_2~ +}{\tau}$, and by the appropriate choice of \Rulename{T-1}, \Rulename{T-2}, \Rulename{T-3}, or \Rulename{T-4}, $\etyped{v_3}{\tau}$., so $\tau^* = \tau^{*\prime} = \tau$.
+    \item \Rname{S-Add}
+        Then $e^* = \kwnumlit{n_1}{\tau}~ \kwnumlit{n_2}{\tau}~ +$ and $e^* \step \kwnumlit{n_1 + n_2}{\tau}$ for some $\tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}$.
+        By \Rname{T-Add}, $\etyped{\kwnumlit{n_1}{\tau}~ \kwnumlit{n_2}{\tau}~ +}{\tau}$, and by the appropriate choice of \Rname{T-Int}, \Rname{T-Long}, \Rname{T-Float}, or \Rname{T-Double}, $\etyped{v_3}{\tau}$, so $\tau^* = \tau^{*\prime} = \tau$.
 
-    \item \Rulename{MUL}
+    \item \Rname{S-Mul}
         Omitted for brevity on the grounds of being mechanically identical to the above case.
 
-    \item \Rulename{EQUIV}
-        Then $e^* = v_1~ v_2~ ==$ where $\etyped{v_1}{\tau}$ and $\etyped{v_2}{\tau}$ for some $\tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}$.
-        By \Rulename{T-8}, $\tau^* = \kwInt$.
-        $e^* \step \kwnumlit{n}{\kwInt}$, and by \Rulename{T-1}, $\etyped{\kwnumlit{n}{\kwInt}}{\kwInt}$, so $\tau^{*\prime} = \tau^* = \kwInt$.
+    \item \Rname{S-Eq}
+        Then $e^* = \kwnumlit{n_1}{\tau}~ \kwnumlit{n_2}{\tau}~ ==$ and $e^* \step \kwnumlit{(\delta_{n_1 n_2})}{\kwInt}$ for some $\tau \in \{\kwInt, \kwLong, \kwFloat, \kwDouble\}$.
+        By \Rname{T-Eq}, $\tau^* = \kwInt$.
+        By \Rname{T-Int}, $\etyped{\kwnumlit{(\delta_{n_1 n_2})}{\kwInt}}{\kwInt}$, so $\tau^{*\prime} = \tau^* = \kwInt$.
 
-    \item \Rulename{TERN-1}, \Rulename{TERN-2}
+    \item \Rname{S-Tern-0}, \Rname{S-Tern-1}
         Then $e^* = \kwtern{\kwnumlit{n}{\kwInt}}{v_1}{v_2}$ and $\etyped{v_1}{\tau}$ and $\etyped{v_2}{\tau}$ for some $\tau$.
-        By \Rulename{T-7}, $\etyped{e^*}{\tau}$, so $\tau^* = \tau$.
-        Either $e^* \step v_1$ or $e_* \step v_2$, so in either case, $\tau^{*\prime} = \tau$.
+        By \Rname{T-Tern}, $\etyped{e^*}{\tau}$, so $\tau^* = \tau$.
+        Either $e^* \step v_1$ or $e_* \step v_2$, so in either case, $\tau^* = \tau^{*\prime} = \tau$.
+
+    \item \Rname{Macro-Subst}
+        Then $e^*$ is a macro with free variables that steps to a macro with captures.
+        The fact that macros perform type inference based on the stack with which they're called (which isn't really known in the context of this rule) makes things pretty complicated, but it should still be the case that by the substitution lemma, this capturing changes nothing.
 \end{itemize}
@@
 
+\section{Conclusion}
+As mentioned before the big takeaway was that we learned some flaws with the base language, notably that union types need to be redesigned.
+The semantics for the language are overly complicated, and while working on proofs we came up with some ideas for simplifying them further.
+We made light use of the lemmas for canonical forms and substitution, but for the former the cases were fairly trivial anyway, and for the latter, the proof case (and the rule it concerns) is somewhat weird and handwavy even with the lemma.
+
 \end{document}

### 2021-12-09 project/template.tex
this one uses the ACM template

--- 2021-12-09 project/paper.tex
@@
-\documentclass{article}
+\documentclass[acmsmall, review]{acmart}
+\settopmatter{printfolios=false,printccs=false,printacmref=false}
 
-% Language setting
-% Replace `english' with e.g. `spanish' to change the document language
-\usepackage[english]{babel}
+%% Rights management information.  This information is sent to you
+%% when you complete the rights form.  These commands have SAMPLE
+%% values in them; it is your responsibility as an author to replace
+%% the commands and values with those provided to you when you
+%% complete the rights form.
+\setcopyright{none}
+%\acmJournal{Topics and Applications in PL}
+\acmYear{2021} \acmVolume{595} \acmNumber{09} \acmArticle{} \acmMonth{12} \acmPrice{}\acmDOI{}
 
-% Set page size and margins
-% Replace `letterpaper' with`a4paper' for UK/EU standard size
-\usepackage[letterpaper,top=2cm,bottom=2cm,left=3cm,right=3cm,marginparwidth=1.75cm]{geometry}
+%%
+%% Submission ID.
+%% Use this when submitting an article to a sponsored event. You'll
+%% receive a unique submission ID from the organizers
+%% of the event, and this ID should be used as the parameter to this command.
+%%\acmSubmissionID{123-A56-BU3}
+
+%%
+%% The majority of ACM publications use numbered citations and
+%% references.  The command \citestyle{authoryear} switches to the
+%% "author year" style.
+%%
+%% If you are preparing content for an event
+%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
+%% citations and references.
+%% Uncommenting
+%% the next command will enable that style.
+\citestyle{acmauthoryear}
+
+%% Some recommended packages.
+\usepackage{booktabs}   %% For formal tables:
+%% http://ctan.org/pkg/booktabs
+\usepackage{multirow}
+\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
+                        %% http://ctan.org/pkg/subcaption
+\usepackage{wrapfig}
+\usepackage{listings}
+\usepackage{amsmath, amsthm, mathtools, mathrsfs, verbatim, enumerate, xcolor, mathpartir}
+\usepackage{mathpartir}
+
+\usepackage{footnote}
 
 % Useful packages
-\usepackage{amsmath, amsthm, verbatim, enumerate, xcolor, mathpartir}
 \usepackage{graphicx}
-\usepackage[colorlinks=true, allcolors=blue]{hyperref}
+% \usepackage[colorlinks=true, allcolors=blue]{hyperref}
 \usepackage{syntax}
@@
 
+\makesavenoteenv{tabular}
+
+\newtheorem{lemma}{Lemma}
+\newtheorem{corollary}{Corollary}
+\newtheorem{theorem}{Theorem}
+\newtheorem{proposition}{Proposition}
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}
+\newtheorem{remark}{Remark}
+
+%%
+%% end of the preamble, start of the body of the document source.
+\begin{document}
+
+%%
+%% The "title" command has an optional parameter,
+%% allowing the author to define a "short title" to be used in page headers.
 \title{Verifying a Postfix Language}
-\author{Dustin Van Tate Testa and Andrew Neth}
 
-\begin{document}
+%% Copy and paste the block of code between here...
+
+\author{Dustin Van Tate Testa}
+\email{dtesta1@hawk.iit.edu}
+\affiliation{
+  \institution{Illinois Institute of Technology}
+  \country{USA}
+}
+
+\author{Andrew Neth}
+\email{aneth@hawk.iit.edu}
+\affiliation{
+  \institution{Illinois Institute of Technology}
+  \country{USA}
+}
+
+%% ... and here for each member of your group.
+
+%%
+%% This command processes the author and affiliation and title
+%% information and builds the first part of the formatted document.
 \maketitle
-% \begin{abstract}
-% A formalization and some proofs of soundness is given to a subset of \href{https://github.com/dvtate/postfix-haskell}{a functional, postfix language}.
-% \end{abstract}
 
@@
 
-\subsubsection{Lemma - Convergence}\footnote{a.k.a. Progress for Types, kinda}
+\subsubsection{Lemma - Convergence}%\footnote{a.k.a. Progress for Types, kinda}
 Part 1: $\tau^* \tstep \tau^{*\prime}$ is no longer possible when $\tau_i \prim ~ \forall~ \tau_i \in \tau^*$.
@@
 
-        By \Rname{T-Call}, the type of $e^{*\prime}$ is inferred\footnote{I'm tired.} based on the expansion of $v$.
+        By \Rname{T-Call}, the type of $e^{*\prime}$ is inferred based on the expansion of $v$.
         By \Rname{T-Fix}, this is also the case for $e^*$.
@@
 
+%% There's no need to have a bibliography. If you do cite papers, uncomment
+%% the following two lines.
+%\bibliographystyle{ACM-Reference-Format}
+%\bibliography{bibfile} % replace "bibfile" with the name of your .bib file
+
 \end{document}
